Dissecting muscle synergies in the task space

Version 1

Abstract

Abstract

The muscle synergy is a guiding concept in motor control research that relies on the general notion of muscles ‘working together’ towards task performance. However, although the synergy concept has provided valuable insights into motor coordination, muscle interactions have not been fully characterised with respect to task performance. Here, we address this research gap by proposing a novel perspective to the muscle synergy that assigns a specific functional role to each muscle coupling by characterising its task-relevance. Crucially, our novel perspective provides nuance to the muscle synergy concept, demonstrating how muscular interactions can ‘work together’ in different ways: a) irrespective of the task at hand but also b) redundantly or c) complementarily towards common task-goals. To establish this perspective, we leverage information- and network-theory and dimensionality reduction methods to include discrete and continuous task parameters directly during muscle synergy extraction. Specifically, we introduce co-information as a measure of the task relevance of muscle interactions and use it to categorise such interactions as task-irrelevant (present across tasks), redundant (sharing the same task information) or synergistic (representing different aspects of the task). To demonstrate these types of interactions in real data, we apply the framework to large-scale datasets of human movements and extract generalizable and scale-invariant representations consisting of subnetworks of synchronised muscle couplings and distinct temporal patterns. The representations effectively capture the functional interplay between task end-goals and biomechanical affordances and the concurrent processing of functionally similar (redundant) and complementary (synergistic) task information. The proposed framework unifies the capabilities of current approaches in capturing distinct motor features while providing novel insights and research opportunities through a nuanced perspective to the muscle synergy.

Introduction

Human movement is a highly complex behaviour, with a broad spectrum of multiplexed spatiotemporal dynamics typically exhibited for basic activities-of-daily-living [1,2]. How the central nervous system controls movement in the face of this inherent complexity to ensure efficient and reliable navigation of the environment and task performance is a nontrivial question currently under investigation in the motor control field [3,4]. The muscle synergy hypothesis is a long-withstanding proposition on the underlying neural constraints producing coordinated movement, stating that this complexity is offset by the allocation of computational resources to the spinal-level in the form of motor primitives [3–6]. These motor primitives modularly activate functional groups of muscles that are flexibly combined for the efficient construction of a movement. The conceptual underpinning of the ‘muscle synergy’ entails the idea of combinations of muscles ‘working together’ for the purpose of effective goal-orientated behaviour [7]. This emergent cohesion involves the following qualifying attributes: a repeatable muscle activation pattern common across trials and participants, a reciprocal relationship among functional muscle groups such that changes occur in one group to compensate for changes in another, and task dependence (i.e. the pattern of interdependencies among muscles must map onto task performance) [7,8]. A common approach to analyse the neural constraints underlying these motor patterns is to apply unsupervised machine-learning algorithms to electromyographic (EMG) data [9,10], with the aim of extracting a latent, low-dimensional representation.

In [11], we considered, key limitations among current approaches to muscle synergy analysis in extracting functionally relevant and interpretable patterns of muscle activity [12]. We proposed a combinatorial approach based on information- and network-theory and dimensionality reduction (the network-information framework (NIF)) that significantly improved the generalisability of the extraction process by, among others, removing restrictive model assumptions (e.g. linearity, same mixing coefficients) and the reliance on variance-accounted-for (VAF) metrics [12]. To elaborate on the significance of this development, the extraction of motor patterns in isolation of the task space comes at the expense of both functional and physiological relevance [12,13]. Thus, here we propose a computational framework that allows for the direct inclusion of task space parameters which is of particular importance for the progression of the research field and successful engineering applications [12,14,15]. Furthermore, effective methods for mapping large-scale physiological dynamics to behaviour is a current gap across the neurosciences [16]. Here, we address this research gap by directly including task parameters in the extraction of muscle synergies using the NIF [11]. This enables us to dissect the concept of muscle synergy and therefore quantify different types of interactions between muscle activations with distinct but complementary functional roles in task space.

In its currently defined state, the muscle synergy concept describes the role of common neural drives to functional muscle groupings working redundantly towards a common task-goal [3–5,7,14]. However, recent influential works have highlighted several other important mechanisms involved in this low-dimensional control strategy that are not well recognised by the muscle synergy concept [17–25]. Such insights include the partitioning of motor variability by the nervous system across task-relevant and -irrelevant spaces and the cooperation between functionally distinct muscle groupings in the form of cross-module functional connectivities. These observations highlight the need for a refinement of the muscle synergy concept to comprehensively describe muscle interactions during movement, including their partitioning into task-relevant and –irrelevant spaces and the characterisation of their functional roles.

We thus motivate the development of a more nuanced perspective to the muscle synergy concept and the general notion of ‘working together’ that comprehensively describes the muscle interactions underlying motor behaviour. To do so, we propose an information-theoretic approach (based on the NIF pipeline) that characterises the contributions of muscle couplings to motor task performance. In other words, we frame the notion of ‘working together’ in more specific terms of shared information between pairs of spatiotemporal muscle activations ([mx, my] (Fig.1.3(a))) (red and green sets respectively) and a corresponding task parameter (τ) (blue set) (see Venn diagrams in Fig.1.3(b) below). Among current approaches to muscle synergy analysis, the shared information between muscles (yellow and white area in Fig.1.1(a)) is quantified, using dimensionality reduction, as common patterns of variability. These common patterns are essentially task-agnostic and may contain patterns of variability a) present in specific tasks (i.e. task-relevant (white shaded area in Fig.1.1(a)) as well as b) shared across tasks (i.e. task–irrelevant (yellow shaded area in Fig.1.1(a)). Our proposed approach dissects patterns of muscle variability in space, time and across trials in terms of their task-relevance and functional similarity in a generalizable manner using mutual information (MI) (Fig.1.3(a-c)). This enables us to decompose muscle activations into muscle pair – task parameter couplings and characterise their combined functional roles. We can then extract low-dimensional representations of these muscle couplings, i.e. muscle networks with specific spatial and temporal signatures, across participants and tasks (Fig.1.3(c)).

A general outline of the proposed approach.

(1.1(a-b)) We propose a novel approach to mapping muscle couplings to the task space. Among current muscle synergy analysis approaches, muscle couplings are quantified in isolation of the task solely using dimensionality reduction. Using our approach, the functional characteristics of muscle interactions can be quantified in terms of the similarity of their encoded task information. We do so by determining the coupling between [mx, my] and a corresponding task parameter (τ) using mutual information (MI). From this perspective, task-redundant muscle couplings (pink shaded area in pink-orange intersection) represent muscles cooperating towards similar task goals, while task-synergistic muscle couplings (orange shaded area in pink-orange intersection) encapsulate the task information provided by a muscle pairing acting towards complementary task goals. Muscle couplings present across tasks (i.e., task-irrelevant) are quantified by conditioning the MI between [mx, my] pairs with respect to τ (yellow intersection). (1.2) A description of redundant and synergistic interactions. (a) Net redundant interactions are defined by a greater amount of information generated by the sum of individual observation of mx and my ([mx+ my]) than their simultaneous observation ([mx, my]). (b) In a net synergistic interaction, [mx, my] provides more information than [mx+ my]. (1.3(a-c)) An overview of the approach. Spatiotemporal muscle activation samples are extracted across trials from large-scale EMG datasets and concatenated into vectors, forming [mx, my] pairs.The derived muscle couplings are then run through the NIF pipeline [11], producing low-dimensional, multiplexed space-time muscle networks.

Crucially, using this novel framework, we can separately quantify the task-irrelevant (i.e. muscle interactions present across tasks) information conveyed by a muscle coupling (yellow intersection in Fig.1.1(b)) from the task-relevant information (pink-and-orange shaded area in Fig.1.1(b)). These task-relevant interactions can be either sub-additive/redundant (i.e. the muscle coupling provides less information about the task compared to the sum of the individual muscle patterns) or super-additive/synergistic (i.e. the muscle coupling conveys more task information than the sum of individual muscle-task encodings). Conceptually, the information a muscle interaction provides is considered redundant when all the information can essentially be found in one of the muscles (pink shaded area (Fig.1.2(a)). This redundant task information thus reveals a functional similarity between muscle activations. Alternatively, we can also identify muscles that act synergistically towards complementary task goals, meaning their variations provide different information about a motor behaviour. A key, quantifiable attribute of this complementary interaction is the emergent task information (‘synergy’) they provide when considered together (orange shaded area (fig.1.2(b)). From this novel perspective, muscle activations can ‘work together’ not just similarly towards a common task-goal but also complementarily towards different aspects of motor behaviour and concurrently towards objectives functionally irrelevant to overt task performance, thus providing a comprehensive view of the muscle interactions governing coordinated movement.

To illustrate this novel conceptual and analytical framework, we applied it to three benchmark datasets of human participants performing naturalistic movements, extracting generalizable and functionally interpretable space-time muscle networks with respect to both discrete and continuous task spaces. We have also made available open-source Matlab routines for readers to apply this approach to their own data (https://github.com/DelisLab/EMG2Task).

Results

Our primary aim here is to provide a new definition for the muscle synergy in task space by quantifying the contributions of muscle couplings to task performance. To achieve this, we essentially reverse the analytical approach typically used in muscle synergy studies (i.e. muscle groupings are identified and inferences then made about their functional roles) [9]. More specifically, we first identify functional couplings between paired muscle activations by evaluating their task-relevance and then extracting representative patterns of such couplings using dimensionality reduction methods.

To begin, we derived pairs of muscle activation vectors [mx, my] from three benchmark datasets of human participants executing naturalistic movements, namely arm reaching (dataset 1), whole-body point- to-point reaching (dataset 2) and various locomotion modes (dataset 3) [26–28] (see Fig.3 for the experimental design of each dataset and ‘Materials and methods’ for an outline of the experimental setups and EMG data pre-processing)(fig.2(A)). For datasets 1 and 2, we determine the MI between [mx, my] vectors with respect to several discrete task parameters representing specific task attributes (e.g. reaching direction, speed etc.), while for dataset 3 we determine the MI with respect to corresponding kinematic, dynamic and inertial motion unit (IMU) features sampled across trials.

A summary of the NIF pipeline. (A) Large-scale datasets of EMG signals are captured while participants perform various motor tasks [26–28]. (B) The MI between all unique muscle-timepoint vector ([mx, my]) combinations with respect to a corresponding task parameter (τ) is determined [41], forming a network of functional connectivities. (C) These adjacency matrices are then analysed in terms of statistical significance and modular structure using percolation theory [11]. (D) The optimal spatial and temporal model-ranks are determined using generalised, consensus-based network community detection methods [32–34,36]. (E) The optimal model-ranks are used as input parameters for dimensionality reduction, where space-time muscle networks along with their underlying activation coefficients are concurrently extracted [26].

Graphical illustrations of each of the datasets analysed in the current study. (A) Dataset 1 consisted of participants executing table-top point-to-point reaching movements across four targets in forward (P1-P4) and backwards (P5-P8) directions at both fast and slow speeds [26]. The muscles recorded included the finger extensors (FE), brachioradialis (BR), biceps brachii (BI), medial-triceps (TM), lateral-triceps (TL), anterior deltoid (AD), posterior deltoid (PD), pectoralis major (PE), latissimus dorsi (LD) of the right, reaching arm. (B) For dataset 2, the activity of 30 muscles was recorded while participants performed whole-body point-to-point reaching movements across three different heights and bars and in various directions, accumulating to 72 unique reaching tasks [27]. (C) The circuit navigated by participants in dataset 3 as they executed various locomotion modes is illustrated, of which level-ground walking, stair- and ramp-ascent/descent were analysed in the current study [28]. Several sub-conditions were undertaken by participants for each locomotion mode including different walking-speeds, clockwise vs. counter-clockwise direction, different stair heights and ramp inclines etc. Participants executed these tasks while the EMG of 11 muscles on the right leg ((Gluteus medius (GlutM), right external oblique (Obl), semitendinosus (ST), gracilis (GR), biceps femoris (BF), rectus femoris (RF), vastus lateralis (VL), vastus medialis (VM), soleus (SO), tibialis anterior (TA), gastrocnemius medialis (GM)) along with kinematic, dynamic and IMU signals were captured.

Having extracted the muscle pair-task interdependencies representing a specific intersection in fig.1.1(b), we next sought to find a parsimonious representation of motor behaviour that is consistent across tasks and participants (fig.2(E)) [26]. To produce this sparse, low-dimensional representation, we undertook the following intermediary steps:

We modelled the MI values as adjacency matrices in the spatial or temporal domain and identified dependencies that were statistically significant using percolation theory (fig.2(C)) [29,30]. By assuming the muscle networks operate near a state of self-organised criticality [31], we effectively isolated dependencies that were above chance-level occurrence, thus empirically sparsifying the networks.

To empirically determine the number of components to extract, we then concatenated these adjacency matrices into a multiplex network and employed network community-detection protocols to identify modules across spatial and temporal scales (fig.2(D)) [32–36]. Having detected the spatiotemporal modular structure, we then returned the sparsified networks to their original format and used the number of modules identified as input parameters into dimensionality reduction (fig.2(E)) [26].

By optimising a modularity-maximising cost-function [37,38], the community detection protocols we employed consistently identified three spatial (S1-S3) and two temporal (T1-T2) modules as representative of the underlying task-redundant, -synergistic and -irrelevant informational dynamics. Following their extraction, we further analysed the spatial networks from each dataset in terms of their submodular structure by applying network-theoretic tools [33,34,39]. In doing so, we identified subnetworks within each spatial network and interesting patterns of network centrality, i.e. the relative importance of a node in a network. The spatial and temporal networks of each dataset output are illustrated in panels A and B (fig.4-9) of the following sections. They are accompanied by human body models where node colour and size indicate subnetwork community affiliation and network centrality respectively [40]. The networks we extracted operate in parallel within spatial and temporal domains while having an all-to-all correspondence across domains, i.e. any spatial component can be combined with any temporal component via a task-specific coefficient (illustrated in panel C for dataset 1 and 2 outputs and in the supplementary materials (fig.4-6) for dataset 3) [26]. Unlike similar muscle synergy extraction approaches, dimensionality reduction in the NIF pipeline doesn’t seek to approximate the recorded EMG data but to identify sets of muscles that share the same type of interaction. Thus, the multiplexing coefficients extracted in this framework are instead interpreted as the participant- and task-specific scaling of information overlap.

A simplified example output from the proposed framework applied to a single trial of turning gait from Dataset 3. (A) Task-irrelevant, (B) Task-redundant and (C) Task-synergistic synchronous muscle couplings were quantified with respect to the heel kinematic marker (anterior-posterior direction). Human body models accompanying each spatial network illustrate their respective submodular structure with node colour and size and edge width indicating community affiliation [33], network centrality and connection strength respectively [39,40].

Three spatial (S1-S3) and two temporal task-irrelevant muscle networks (T1-T2) were empirically identified and extracted across participants and task parameters from dataset 2 using the NIF pipeline (Panel A-B) [11,27]. (Panel C) Activation coefficients are presented to the right of the networks, indicating their task parameter-specific scaling averaged across participants. Human body models accompanying each spatial network illustrate their respective submodular structure with node colour and size and edge width indicating community affiliation [33], network centrality and connection strength respectively [39,40].

Three spatial (S1-S3) and two temporal task-irrelevant muscle networks (T1-T2) were empirically identified and extracted across participants and task parameters from dataset 3 using the NIF pipeline (Panel A-B) [11,28]. Activation coefficients are presented in supplementary materials (fig.4), indicating their task parameter-specific scaling averaged across participants in the dynamic, IMU and kinematic spaces. Human body models accompanying each spatial network illustrate their respective submodular structure with node colour and size and edge width indicating community affiliation [33], network centrality and connection strength respectively [39,40].

Three spatial (S1-S3) and two temporal task-redundant muscle networks (T1-T2) were empirically identified and extracted across participants and task parameters from dataset 2 using the NIF pipeline (Panel A-B) [11,27]. (Panel C) Activation coefficients are presented to the right of the networks, indicating their task parameter-specific scaling averaged across participants. Human body models accompanying each spatial network illustrate their respective submodular structure with node colour and size and edge width indicating community affiliation [33], network centrality and connection strength respectively [39,40].

Three spatial (S1-S3) and two temporal task-redundant muscle networks (T1-T2) were empirically identified and extracted across participants and task parameters from dataset 3 using the NIF pipeline (A-B) [11,28]. Activation coefficients are presented in supplementary materials document 2 (fig.2), indicating their task parameter-specific scaling averaged across participants in the dynamic, IMU and kinematic spaces. Human body models accompanying each spatial network illustrate their respective submodular structure with node colour and size and edge width indicating community affiliation [33], network centrality and connection strength respectively [39,40].

Three spatial (S1-S3) and two temporal task-synergistic muscle networks (T1-T2) were empirically identified and extracted across participants and task parameters from dataset 2 using the NIF pipeline (Panel A-B) [11,27]. (Panel C) Activation coefficients are presented to the right of the networks, indicating their task parameter-specific scaling averaged across participants. Human body models accompanying each spatial network illustrate their respective submodular structure with node colour and size and edge width indicating community affiliation [33], network centrality and connection strength respectively [39,40].

Types of muscle interactions

To provide intuition on the types of muscle couplings that can be identified by the proposed frameworks, we first applied the methodology to EMG recordings from a single trial of a participant walking on level-ground in the counter-clockwise direction around the circuit depicted in fig.3(C). This simple dataset can exemplify the different functional roles of muscle couplings with respect to the task at hand. Here we identified task-irrelevant (A), -redundant (B) and –synergistic (C) couplings with respect to a single, continuous task parameter, the heel kinematic marker in the anterior-posterior direction.

For example, the hamstring muscles (ST and BF) controlling knee flexion work together with a) muscles involved in hip abduction (GR, GlutM and Obl) to move the heel forward with a redundant interaction (Fig. 4B) and b) calf muscles involved in ankle flexion (GM, TA) cooperatively determining heel position with a synergistic interaction (Fig. 4C). Also, these two hamstring muscles together with their antagonist RF form a task-irrelevant network, i.e. their couplings are not predictive of heel position (Fig. 4A). Overall, task-irrelevant muscle couplings capture primarily interactions between co-agonist and agonist-antagonist muscles across the right-leg that are indiscriminative of anterior-posterior heel marker variations. Also, task-redundant and - synergistic couplings are identified for functionally similar and dissimilar muscle combinations respectively that provide sub-additive (i.e. partly shared or redundant) and super-additive (i.e. complementary or synergistic) information about heel marker position.

In the following, we sequentially present in more detail the three types of muscular interactions in the spatiotemporal domain across three EMG datasets and show the robustness of the approach and its outputs.

Representations of motor behaviour in muscle couplings

Task-irrelevant muscle couplings

To quantify the task-irrelevant contributions of muscular interactions to motor behaviour, we conditioned the MI between [mx, my] with respect to τ (see ‘Materials and methods’ section) [41]. This conditioning effectively removes the task-relevant information, leaving information produced by pairwise muscle variations that are task-indiscriminative. Following a run through the NIF pipeline (fig.2) [11], the output from dataset 2 and 3 are presented in fig.5-6 respectively while the output from dataset 1 is presented in the supplementary materials (fig.1). The task-irrelevant space-time muscle networks we extracted from dataset 1 and 2 shared several structural features with their task-agnostic counterparts extracted in the preliminary study [11], supporting recent work showing that functional muscle network structure are heavily influenced by task-irrelevant factors such as anatomical constraints [42]. The functional connectivities identified here captured known contributions of spatiotemporal muscular interactions to aspects of motor behaviour common across tasks and participants which we outline briefly here.

The temporal networks from dataset 1 and 2 captured mostly co-activations from movement onset – mid-movement and movement cessation, indicating that some co-contraction mechanisms were consistently task-irrelevant across trials. The temporal networks for dataset 2 were more diffuse compared to dataset 1 (fig.4B) probably reflecting the more variable role of passive forces in generating movements to different heights captured in this dataset’s experimental design [27,43]. Furthermore, this co-contraction mechanism was more parsimoniously represented as a single network in dataset 3 (T1 in fig.6B), where passive forces in contrast likely played a consistently resistive role during locomotion. Interestingly, T1 for dataset 3 corresponded equivalently high for all three task spaces when corresponding with S2 which consisted of upper-leg extensors (see Supplementary materials (fig.4)). Muscle couplings indicative of agonist-antagonist pairings were also identified as separate subnetworks in S3 of dataset 3 (fig.6A). More specifically, their functional segregation appeared to be based on their distinct functional roles in forward propulsion (red nodes) and deceleration (blue nodes) during the mid-stance phase of gait, as indicated by the prominent correspondence with T2 across task spaces (see Supplementary materials (fig.4)). This finding reflects the consistent agonistic and antagonistic contributions of muscular interactions across locomotive tasks.

The gross motor function of muscle couplings was another characteristic of task-irrelevant muscle couplings that pervaded across the datasets analysed here. For instance, AD had a central role in S2 of dataset 1 while also displaying a unique pattern of connectivity with tibial musculature in S3 of dataset 2. Similarly, GlutM had a central role in S1 of dataset 3 (fig.6A). We further found a common pattern of task-irrelevant connectivity in S2 across datasets, namely the musculature about a hinge joint (elbow in dataset 1 and 2, knee in dataset 3) coupled with proximal shoulder or hip musculature, indicative of their biomechanical affordances. Finally, the passive, left-arm was connected with the tibial musculature of S3 in dataset 2 (green nodes). To probe the underlying function of this connectivity in the left-arm, we inspected the original EMG signals. We observed periodic, tonic activations across tasks, reflective of reciprocal inhibition of contralateral limb musculature that enables unilateral movement [44].

Task-redundant muscle couplings

To characterise the functional role of task-relevant muscle couplings, we employed a higher-order information-theoretic concept known as co-information (co-I) [41,45,46]. This metric quantifies the MI between three random variables and may take on positive values (net synergistic) and negative values (net redundant) (see ‘Materials and methods’ section). Co-I quantifies the task-relevant information shared between [mx, my] independently of the information generated by task-irrelevant muscular interactions. In doing so, it also defines the functional relationship between [mx, my] overall as redundant or complementary. Following the quantification of co-I for all unique [mx, my] and corresponding τ (pink area in orange-and-pink intersection (fig.1.3(B)), we parsed the negative values indicating redundancy into a separate matrix and rectified them. In fig.7-8, we illustrate the output following the extraction of task-redundant space-time muscle networks from datasets 2-3 across tasks and participants respectively while the output for dataset 1 is presented in supplementary materials (fig.2). In the co-I formulation, task-redundant muscle couplings can be interpreted as muscle couplings that overall shared a common task-relevant functional role. For example, with reference to Figure 7, muscles in the networks presented in S1 (fig.7A) carry redundant information about the movement endpoint (fig.7C) with the temporal profile T1 (fig.7B) whereas S3 (fig.7A) contains muscle networks carrying redundant information about the starting point (fig.7C) with the same temporal profile T1 (fig.7B).

Both dataset 1 (supplementary materials (fig.2)) and dataset 3 (fig.8) outputs display similar patterns of muscle couplings at the same spatial scale of an individual, task-relevant limb musculature, with an emphasis on the coupling of specific muscles with all other muscles. For dataset 1, FE, BI and BR displayed this integrative pattern across S1-S3 respectively while BF, TA and ST demonstrated this pattern in dataset 3 also. The muscle networks encapsulated several functionally interpretable couplings such as the agonist-antagonist pairings of the BI and TM of S2 (dataset 1 (supplementary materials fig.2(A)) and the task redundancy of ankle dorsi-flexors, and knee/hip flexors during sloped walking for example in S2 of dataset 3 (fig.8(A)) [47]. The functional interpretation of these muscle connectivity patterns was in line with the extracted task-specific activations. For instance, S2 of dataset 1 was modulated most prominently by reaching direction when corresponding with T2, commensurate with the biomechanical affordances of this upper-arm muscle network. Furthermore, S2 of dataset 3 was specifically modulated by the right-thigh kinematic marker along the y-axis (up-down direction) for both T1 and T2 (see Supplementary materials (fig.5)). The centrality of task-redundant muscle couplings in dataset 1 and 3 suggests particular muscle activations drive the task-specific variations in the reaching arm and stepping leg muscle activities towards a common behavioural goal. It is also worth noting that the magnitude of these functional connectivity patterns appeared to be proportional to anatomical distance, as evidenced by the magnitude of connection strengths, a finding supportive of previous related research [42].

Meanwhile at the greater spatial scale of dataset 2 (fig.7(A))), task-redundant muscle couplings were anatomically compartmentalised to the upper- and lower-body. This functional segregation was emphasised at the subnetwork level also, where the upper- and lower-body musculature of S3 for instance formed distinct submodules (blue and red nodes). Amongst the task-specific activations in dataset 2, S1 carried redundant task information about end-point target, -height and up-down direction when corresponding with T1. T2 for dataset 2 on the other hand contained mostly temporally proximal dependencies along the diagonal, suggestive of co-contraction mechanisms, which became more diffuse near movement cessation. These endpoint trajectory and co-contraction related temporal patterns were qualitatively similar to T1 and T2 of both dataset 1 and 3 respectively (see fig.8 below and Supplementary materials (fig.2) respectively).

Task-synergistic muscle couplings

Similarly to the task-redundant muscle couplings, we isolated task-synergistic muscle couplings by parsing the positive co-I values from the computations conducted across all unique [mx, my] and corresponding τ into a sparse matrix where all redundant couplings were set to zero (see ‘Materials and methods’ section). Task-synergistic muscle couplings here can be interpreted as a [mx, my] pair that provide complementary (i.e. functionally dissimilar) task information, thus more information is gained by observing [mx, my] together rather than separately (orange area in orange-and-pink intersection (fig.1.3(B)). In Fig.8-9, we illustrate the task-synergistic space-time muscle networks from datasets 2-3 respectively (dataset 1 output is presented in the supplementary materials, Fig.3).

Across datasets, the muscle networks could be characterised by the transmission of complementary task information between functionally specialised muscle groups. The overall structure of the synergistic muscle networks was similar to their task-redundant counterparts. For instance, the emphasis of an individual muscle’s connectivity with all other muscles was evident among synergistic couplings in dataset 3 (see VM, VL and RF of S1, GR, TA and SO of S2 and GlutM of S3 in fig.9A). This correspondence with task-redundant networks demonstrates that parallel and synchronous exchanges of redundant and synergistic task information underlie task-specific variations across trials (see e.g. S3 of dataset 2 in fig.6A and fig.8A). Interestingly, despite the similarity between the redundant and synergistic muscle networks, the way they are combined to encode task information differs depending on the type of interaction (synergistic vs. redundant, see e.g. panel C of fig.8 in comparison to panel C of fig.6).

Concerning the temporal activations of these networks, the task-synergistic structure of dataset 2 (fig.8B) was also relatively unchanged compared to the task-redundant structure (fig.6B). This suggests that the task end-goal and co-contraction related mechanisms provided both redundant and synergistic task information concurrently during whole-body reaching movements. In contrast, a different view of synergistic information exchange is provided in datasets 1 and 3 (fig.9 and supp. fig.3), where T1 and T2 consist of more idiosyncratic activations that together appear to reflect the task end-goal related patterns found elsewhere. More specifically, in both datasets we found two distinct patterns of task-end goal related activity where early and late timepoints during movement initiation operated in parallel to provide complementary task information (see fig.9B and supp. fig.3B).

Generalisability of the extracted space-time muscle networks

To ascertain the generalisability of the extracted representations presented here beyond any subset of the input data, we conducted a similarity analysis through a leave-n-out cross validation procedure. In more detail, we compared the space-time networks extracted from the full dataset (illustrated in fig.4-9 and supplementary materials here) to the networks extracted from a subset of the data (see ‘Materials and methods’ section). Across datasets, a high level of concordance was found on average (∼0.9 correlation, see table1). This trend was evident across all datasets and for task-redundant, -synergistic and -irrelevant spatial and temporal networks. Dataset 1 and 2 findings demonstrate that the extracted networks are generalisable beyond any individual participant or task. Dataset 3 results go further by demonstrating that the extracted patterns are generalizable beyond any randomly selected and randomly sized subset of the input data. The highest correlations on average were consistently among temporal networks, replicating previous findings [11,48,49]. Although the spatial networks demonstrated a lower average correlation, this was substantially higher compared to previous applications of datasets 1 and 2 [11,26,48], suggesting that the inclusion of task parameters here captures the inter-participant differences more effectively. When comparing representations extracted from each continuous task space in dataset 3, kinematic features consistently had the highest average correlation and lowest variability compared to dynamic and IMU feature spaces. These findings support our change in the interpretation of the extracted activation coefficients away from conventional approaches where representational biases towards particular participants and/or tasks are inferred.

To further probe how the underlying assumption of an all-to-all correspondence between spatial and temporal representations made by sNM3F influenced the generalisability of the extracted networks, we compared its performance to non-negative Canonical-Polyadic (CP) tensor decomposition, which assumes an opposing one-to-one correspondence between components. We found that although CP has demonstrated a considerable capacity to de-mix neural data into simplified and interpretable low-dimensional components [50], its application here resulted in poor generalisability of the extracted patterns (∼0.5 correlation on average). This finding suggests that the all-to-all correspondence implemented by sNM3F identifies a more generalizable representation and should be favourably considered in future applications of this framework.

Discussion

The aim of the current study was to dissect the muscle synergy concept and offer a novel, more nuanced perspective on how muscles ‘work together’ to achieve a common behavioural goal. To do so, we introduced a computational approach based on the NIF pipeline (fig.2), enabling the effective decomposition of muscle interactions and the comprehensive description of their functional roles. Through the direct inclusion of task parameters in the extraction of muscle synergies, a novel perspective was produced where muscles ‘work together’ not just towards a common task-goal, but also concomitantly towards complementary task objectives and lower-level functions irrelevant for overt task performance. The functional architectures we uncovered were comprised of distributed subnetworks of synchronous muscle couplings and driven by distinct temporal patterns operating in parallel. These representations were also scale-invariant to dataset complexity and motor behaviours whilst being generalizable beyond any subset of the data. We thus present the proposed framework as a useful analytical approach for mechanistic investigations on large-scale neural data through this novel perspective to the muscle synergy.

The ‘muscle synergy’ is a major guiding concept in the motor control field [3–7,14], providing a conceptual framework to the problem of motor redundancy that centres around the general notion of ‘working together’. In its current conceptualisation, ‘working together’ describes how the nervous system functionally groups muscles in a task-dependent manner through common neural drives to simplify movement control. This idea has undergone continued refinement since its early conception [3,6,7], with a notable progression being the introduction of the qualifying attributes: a sharing pattern, reciprocity and task dependence [7]. Nonetheless, recent influential works revealing other important mechanisms for the simplification of motor control highlight the generality of the current perspective offered by the muscle synergy concept [17–25]. We thus sought to provide greater nuance to the notion of ‘working together’ by defining motor redundancy and synergy in information-theoretic terms [6,51]. In our framework, redundancy and synergy are terms describing functionally similar and complementary motor signals respectively, introducing a new perspective that is conceptually distinct from the traditional definition of muscle synergy as a solution to the motor redundancy problem [3,6,7]. In this new definition of muscle interactions in the task space, a group of muscles can ‘work together’ either synergistically or redundantly towards the same task. In doing so, the perspective instantiated by our approach provides novel coverage to the partitioning of task-relevant and -irrelevant variability implemented by the motor system along with an improved specificity regarding the functional roles of muscle couplings [21–23]. Our framework emphasises not only the role of functionally redundant (i.e. similarly tuned) muscle couplings but also of complementary, synergistic dependencies that have been shown to be important for communication and integration across specialised neural circuitry [52,53]. Thus, the present study aligns the muscle synergy concept with the current mechanistic understanding of the nervous system whilst offering an analytical approach amenable to the continued advances in large-scale data capture [16,54].

Among current approaches to muscle synergy analysis, the established synchronous, temporal and time-varying muscle synergy models are understood to each characterise unique motor features [55]. More specifically, the synchronous model captures agonist-antagonist muscle pairings, the temporal model decomposes EMG signals into functionally distinct temporal phases and the task-specific modulation of spatiotemporal invariants are quantified in the time-varying model. In a unifying framework, here we quantify space-time muscle networks that concurrently capture many of these salient motor features in a holistic and principled way whilst mapping their functional consequences to motor behaviour. These salient features included, among others, agonist-antagonist pairings and functionally meaningful inter-limb couplings that consistently appeared across task-redundant, -synergistic and -irrelevant spaces. Thus, in dissecting the muscle interactions governing coordinated movement, our framework revealed their parallel and synchronous processing of functionally similar, complementary, and task-irrelevant information. This insight aligns with several recent works also demonstrating this distributed neural architecture of parallel information processing units [1,52,53].

Our framework also revealed novel characteristics of the motor system. For instance, the task-redundant and -synergistic networks we extracted appeared to be centralised around the couplings of individual muscles, supporting recent work showing the nonhomogeneous sharing of neural drives within modules [56]. These novel spatial characteristics were driven by parallel temporal patterns representing endpoint trajectory and co-contraction related mechanisms, an insight supportive of recent work showing their parallel innervation [17,43,57]. Together, these representations encapsulated the functional interplay between task end-goal requirements and biomechanical affordances, a dynamic frequently highlighted in object manipulation experiments [58]. In other words, the task-relevant networks reflected how muscles ‘work together’ both redundantly and synergistically towards a desired end-goal state whilst continually controlling the present trajectory of the system. Meanwhile, the task-irrelevant networks demonstrated that these muscle couplings also work concomitantly towards lower-level objectives assigned mostly during the transition to this desired end state. These objectives may include gross motor functions [59], the maintenance of internal joint mechanics and reciprocal inhibition of contralateral limbs [20,44], reflecting both biomechanical- and task-level constraints that provide a structural foundation for task-specific interactions. The separate quantification of these distinct task spaces to coordinated movement opens up novel opportunities in the practical application of muscle synergy analysis [5,12,15]. For instance, these distinct representations may encapsulate different neural substrates that can be specifically assessed at the muscle-level for the purpose of bodily restoration and augmentation [60]. Uncovering their neural underpinnings is an interesting topic for future research.

In future work, we aim to complement this study’s combinatorial perspective to the muscle synergy by dissecting the unique contribution of individual muscles to motor behaviour and how they may work independently towards task performance (see magenta and cyan intersections in fig.1.3(B)). More broadly, our work here parallels related information-theoretic approaches to decomposing task-relevant brain activity [46,61], whilst addressing a current research gap across the neurosciences in effective methods to mapping large-scale, spatiotemporal neural activity to behaviour [16,54]. Future applications of this framework should include large-scale, multi-modal data captured from participants performing a wide range of natural behaviours.

In sum, this study introduced a novel perspective to the muscle synergy concept and a computational framework to extract muscle couplings that map their pairwise contributions to motor behaviour. We suggest that this approach offers novel research opportunities for investigating the underlying neural constraints on motor behaviour and the fundamental structure-function relationships generated by agent – environment interactions [14,62].

Materials and Methods

Quantifying muscle couplings in the task space

To quantify muscle couplings we used MI, a non-linear measure of dependence that captures any type of relationship between random variables. Here to estimate MI, we used a Gaussian copula-based approximation [41]. This semi-parametric estimator exploits the equivalence between MI and the negative entropy of the empirical copula (c), a function that maps a multivariate set (e.g. [mx, my] representing activities of muscles X and Y) to their joint distribution (Equation 1.1) [41]. Thus, to determine task-irrelevant muscle couplings (I(mx;my|τ)), we conditioned the negative entropy of the empirical copula for [mx, my] with respect to a task variable τ (Equation 1.2). As mentioned, [mx, my] are continuous vectors composed of individual muscle amplitudes at specific timepoints across trials while τ is a corresponding discrete (e.g. movement direction for datasets 1 and 2) or continuous (e.g. movement kinematics in dataset 3) task parameter. For discrete task variables, τ takes one value for each trial and the MI is calculated across trials using a Gaussian mixture model [41]. In the case of continuous task variables, τ varies in time within a specific trial. Thus, we compute MI at each timepoint t using the muscle activity mx(t) and the task variable value τ(t) at this time point using a closed-form parametric estimator [41]. To evaluate the task-relevance of the identified muscle couplings, we used a higher-order information theoretic measure known as co-information (co-I) (Equation.1.3) (fig.11), which quantifies the relationship between three random variables, here [mx, my], and τ. Co-I implements the inclusion-exclusion principle of combinatorics [45], whereby the sum of MIs between individual m vectors and τ (I(mx;τ)+ I(my;τ)) is compared against their composite MI (I(mx, my;τ)) as follows:

Three spatial (S1-S3) and two temporal task-synergistic muscle networks (T1-T2) were empirically identified and extracted across participants and task parameters from dataset 3 using the NIF pipeline (Panel A-B) [11,28]. Activation coefficients are presented in supplementary materials (fig.6), indicating their task parameter-specific scaling averaged across participants in the dynamic, IMU and kinematic spaces. Human body models accompanying each spatial network illustrate their respective submodular structure with node colour and size and edge width indicating community affiliation [33], network centrality and connection strength respectively [39,40].

Co-I determines the difference between the sum total information shared with τ in mx and my when observed separately and the information shared with τ when they are observed together. The adjacency matrices show how this calculation is carried out for all unique [mx, my] combinations. Redundant and synergistic muscle couplings are then separated into two equivalently sized networks. The accompanying colour bars indicate the values present in the adjacency matrix.

Negative I(mx;my;τ) corresponds to a net redundant coupling between [mx, my] about τ while positive I(mx;my;τ) indicates a net synergy. To analyse these distinct couplings separately, we parsed redundant and synergistic I(mx;my;τ) into two equivalently sized matrices and rectified the redundant couplings to make them suitable for non-negative dimensionality reduction.

Then, to produce a multiplexed view of the muscular interactions across trials, we iterated these MI computations over all unique combinations of [mx, my] and τ. The resulting MI estimates collectively form A, a symmetric adjacency matrix (i.e. ATA = I) that represents the functional connectivities between all muscle activations (fig.10). When repeated across all available task variables τ and participants, A is of dimension [No• of muscle pairs x No• of timepoint pairsx [No• of τ x No• of participants]]. Thus, by applying network-theoretic statistical tools to A, we can identify functional modules carrying the same type of (redundant/synergistic) task information (fig.2(B)).

Estimating statistical significance of muscle couplings

To isolate statistically significant dependencies, we applied a modified percolation analysis to each A [29]. This method sparsifies functional connectivities in A with respect to its percolation threshold (Pc). Pc is a critical value that specifies the probability of a nodes’ occupation in A with respect to the networks size. In random networks, a ‘giant component’ comprised of long-range connections exists above Pc but disappears as Pc tends to zero [30], while it is thought that living systems optimise adaptability by fluctuating around Pc in a state of self-organised criticality [31]. Pc was iteratively specified for each layer of A relative to equivalently sized random networks and utilised to remove insignificant network edges up to a stopping-point where this giant component begins to become affected (fig.2(C)). This procedure was carried out for each layer of A separately configured as muscle-wise couplings across temporal scales (i.e. a 3-D tensor of dimension [No• of muscles] x [No• of muscles] x [No• of timepoint pairings x No• of τ x No• of participants]) and vice versa as timepoint-wise couplings across spatial scales (fig.2(D)). The separate sparsification of each individual network layer in both alternative network configurations produced discrepancies in the output, as some connections were found to be significant in only one domain. To ameliorate this discrepancy, we employed a conservative heuristic where dependencies must be significant in both space and time to be included in the final input matrix for dimensionality reduction. Thus, the sparsified input matrices to dimensionality reduction were comprised of significant spatiotemporal task-redundant, -synergistic or - irrelevant muscle couplings.

Model-rank specification

To determine the optimal number of modules to extract, we implemented two alternative community detection algorithms generalised to multiplex networks [32,33,35]. Both forms seek to optimise a modularity criterion known as the Q-statistic that quantifies the proportion of within-community network edges compared to what would be expected from a network consisting of random connections [37]. More specifically, for a particular division of a single layer network (Equation 2.1), let δ(gi, gj)=1 if node I and j belong to the same group (g) and 0 otherwise and Aij be the number of edges between node i and j. The equivalent of Aij from a randomised network (Pij) is expected to be xs (Newman-Girvan null model) [37], where ki and kj are the node degrees and . The typical output of the Q-statistic is found within the range [0,1] with 1 indicating maximum modularity [37]. In its generalised multilayer form, the Q-statistic is given an additional term to consider couplings between layers l and r with intra- and inter-layer resolution parameters γ and ω (Equation 2.2). Here, μ is the total edge weight across the network and γ and ω were set to 1 in the current study for classical modularity [32]. We chose to implement two complementary model-rank specification approaches to address limitations related to stochasticity and scalability present in the multilayer formulation and the inability to consider inter-layer dependencies present in the mono-layer formulation [19,38]. To apply these algorithms to our data, we grouped the set of A into multiplex networks configured with respect to spatial or temporal scales (fig.2(D)). We then applied these algorithms to both space-time network configurations for individual participant/tasks. This procedure generated a binary adjacency matrix from the resulting community partition vector in each case where 1 indicated the nodes belonged to the same community and 0 otherwise [35]. Following a consensus-based approach [36], we then grouped these binary adjacency matrices into a new multiplex network and re-applied the two alternative community detection algorithms to find an optimal spatial and temporal model-rank [36].

Extraction of low-dimensional representations

Following the specification of an optimal model-rank in the spatial and temporal domains, we used these values as input parameters into dimensionality reduction (fig.2(E)). To extract a low-dimensional representation of motor behaviour across muscle couplings, we applied a sample-based non-negative matrix tri-factorisation method (sNM3F) with additional orthogonality constraints to the matrices consisting of vectorised and concatenated A [26]. More specifically, we decomposed the input three-mode tensor A of dimension [K = No. of unique muscle pairs (m) x (L = No. of time-sample pairs (t) x No. of participants + tasks)] into a set of spatial (V) and temporal (W) factors and the participant- and task-specific weighting coefficients (S) reflecting the amount of information carried by each combination of spatial-temporal factors for each participant and task parameter (Equation 3.1). In equation 3.2, this factorisation is also illustrated in vector sum notation for a single participant and task variable:

Examining the generalisability of extracted representations

To determine the generalisability of the extracted space-time muscle networks, we implemented a representational similarity analysis where we compared representations extracted from the full dataset 1-3 to equivalent networks extracted from a subset of the respective datasets. We computed the similarity between pairs of representations using Pearson’s correlation.

For dataset 1 and 2 (see below), we removed from the input data an individual participant or task at a time and compared the similarity of the decomposition outputs with those obtained from the full dataset. We repeated this for all participants and task variables and reported the average similarity as a measure of robustness of the decomposition.

For dataset 3 (see below), due to the greater number of participants and task parameters, we implemented a more stringent examination. More specifically, we firstly extracted representations from each task space individually and, using these representations as a reference, compared them against functionally similar outputs after removing randomly-sized portions of randomly selected vectors in the input data (up to the no. of column vectors - 1). We repeated this procedure for 50 iterations, and computed summary statistics by converting the coefficients to Fisher’s Z values, computing the average and standard deviation, and then reverting these values back to correlation coefficients.

Subnetwork analysis

To illustrate the relative importance of individual muscles within each network, we determined the total communicability (C(i)) of individual nodes (i) in each network (A). C(i) is defined as the row-wise sum of all matrix exponentials (e) in the adjacency matrix (A) that consider the number of walks between each pair of nodes i and j (Equation 4.1) [39,63]: To emphasise salient functional connectivities in the spatial networks, we sparsified all dependencies with a below average network communicability and illustrated the output on the accompanying human body models [40,63]. To uncover salient subnetwork structures consisting of more closely functionally related muscle activations, we applied the monolayer community detection algorithm in equation 2.1 to the extracted spatial networks [33,34].

Data acquisition and processing

To illustrate our framework, we applied it to three datasets of EMG signals recorded during different motor tasks. In dataset 1 (fig.3(A)), 7 adult participants (age: 27 ± 2 years, height: 1.77 ± 0.03 m) performed table-top point-to-point reaching movements in both forward and backwards directions and at fast and slow speeds while the activity of nine muscles on the preferred right, reaching arm (finger extensors (FE), brachioradialis (BR), biceps brachii (BI), medial-triceps (TM), lateral-triceps (TL), anterior deltoid (AD), posterior deltoid (PD), pectoralis major (PE), latissimus dorsi (LD)) were captured for a total of 640 trials per participant [26]. To enable the quantification of shared information across muscles with respect to specific task attributes, we formulated four discrete task parameters of length equal to the number of trials executed.

These discrete variables represented the trial-to-trial variation in reaching direction (fwd vs bwd), speed (high vs low), and reaching target [P1-P8] (fig.3(A)).

In dataset 2 (fig.3(B)), 3 participants performed whole-body point-to-point reaching movements in various directions and to varying heights while EMG from 30 muscles (tibialis anterior, soleus, peroneus, gastrocnemius, vastus lateralis, rectus femoris, biceps femoris, gluteus maximus, erector spinae, pectoralis major, trapezius, anterior deltoid, posterior deltoid, biceps and triceps brachii) across both hemi-bodies were captured [27]. Like dataset 1, we formulated task parameters each representing a specific task attribute across trials (∼2160 trials per participant). In this case, we formed eight discrete task parameters representing start- and end-point target, -bar, and -height and both up-down (vertical) and left-right (horizontal) reaching directions.

Dataset 3 consisted of multiple trials from 17 participants performing level-ground walking, stair- and ramp-ascents/descents with various sub-conditions (walking speed, clockwise/counter-clockwise direction, different stair/ramp inclines etc.) (fig.3(C)) [28]. These locomotion modes were performed while 11 EMG signals were captured from the right lower-limb (Gluteus medius (GlutM), right external oblique (Obl), semitendinosus (ST), gracilis (GR), biceps femoris (BF), rectus femoris (RF), vastus lateralis (VL), vastus medialis (VM), soleus (SO), tibialis anterior (TA), gastrocnemius medialis (GM)), XYZ coordinates were captured bilaterally from 32 kinematic markers and 4 IMUs and a force-plate captured accelerations and dynamic features among the lower-limbs also. More detailed breakdowns of the experimental design for each dataset can be found at their parent publications [26–28].

For all datasets, we processed the EMG signals offline using a standardised approach [10]: the EMGs for each sample were digitally full-wave rectified, low-pass filtered (Butterworth filter; 20 Hz cut-off; zero-phase distortion), normalised to 1000 time-samples and then the signals were integrated over 20 time-step intervals yielding a waveform of ∼50 time-steps. To match the time-series lengths, we resampled the kinematic, dynamic and IMU recordings of dataset 3 using cubic-spline interpolation to match the EMG signals.

eLife Assessment

The study by O'Reilly and Delis provides a valuable data-driven framework for extracting task-related muscle synergies in a step towards the understanding and practical use of synergies in real scenarios (e.g., evaluation of patients in a clinical environment). The approach is incomplete since the authors did not compare their method with classical physiologically grounded approaches for assessing muscle synergies. In this sense, the comparisons with classical approaches would clarify if physiological assemblies were preserved and were not altered to incorporate task space variables. Despite limitations, the proposed framework would interest motor control and neural engineering researchers.

Reviewer #1 (Public Review):

The proposed study provides an innovative framework for the identification of muscle synergies taking into account their task relevance. State-of-the-art techniques for extracting muscle interactions use unsupervised machine-learning algorithms applied to the envelopes of the electromyographic signals without taking into account the information related to the task being performed. In this work, the authors suggest including the task parameters in extracting muscle synergies using a network information framework previously proposed. This allows the identification of muscle interactions that are relevant, irrelevant, or redundant to the parameters of the task executed.

The proposed framework is a powerful tool to understand and identify muscle interactions for specific task parameters and it may be used to improve man-machine interfaces for the control of prostheses and robotic exoskeletons.

With respect to the network information framework recently published, this work added an important part to estimate the relevance of specific muscle interactions to the parameters of the task executed. However, the authors should better explain what is the added value of this contribution with respect to the previous one, also in terms of computational methods.

In general, the method proposed relies on several hyperparameters and cost functions that have been optimized for the specific datasets. A sensitivity analysis should be performed, varying these parameters and reporting the performance of the framework.

It is not clear how the well-known phenomenon of cross-talk during the recording of electromyographic muscle activity may affect the performance of the proposed technique and how it may bias the overall outcomes of the framework.

Reviewer #2 (Public Review):

This paper is an attempt to extend or augment muscle synergy and motor primitive ideas with task measures. The authors idea is to use information metrics (mutual information, co-information) in 'synergy' creation including task information directly. My reading of the paper is that the framework proposed radically moves from attempts to be analytic in terms of physiology and compositionality with physiological bases, instead into more descriptive ML frameworks that may not support physiological work easily.

This approach is very different from the notions of physiological compositional elements as muscle synergies and motor primitives, and to me seems to really be striving to identify task relevant coordinative couplings. This is a meta problem for more classical analyses. Classical analyses seek compositional elements stable across tasks. These elements may then be explored in causal experiments and generative simulations of coupling and control strategies. The present work does not convince me that the joint 'meta' analysis proposed with task information added is not unmoored from physiology and causal modeling in some important ways. It also neglects publications and methods that might be inconvenient to the new framework.

Information based separation has been used in muscle synergy analyses using infomax ICA, which is information not variance based at core. Though linear mixing of sources is assumed, minimized mutual information is the basis.

Physiological causal testing of synergy ideas is neglected in the literature reviews in the paper. Although these are in animal work, the clear connection of muscle synergy choices and analyses to physiology is important, and needs to be managed in the new methods proposed. Is any correspondence assumed? Possible?

Questions and concerns with the framework as an overall tool:

First, muscle based motor information sources have influences on different time scales in the task mechanics. Analyses of synergies in the methods proposed will be very much dependent on the number and quality of task variables included and how these are managed. Standardizing and comparing among labs, tasks sets and instrumentation differences is not well enough considered as a problem in this new proposed method toolset, at least in my reading. Will replication, and testing across groups ever be truly feasible in this framework? Muscle based motor information sources have influences on different time scales in the task mechanics. Kinematic analyses, dynamic analyses and force plate analyses of the same task may provide task variables that alter the results in the proposed framework it seems.

Second, there is a sampling problem in all synergy analyses. We cannot record all muscles or all task parameters. Examining synergies across multiple tasks seeks 'stationary' compositionality. Including task specific elements may or may not reinforce or give increased coordinative precision to the stationary compositionality.

To me the new methods proposed seem partly orthogonal to the ideas of stable compositionality. The 'synergies' obtained will likely differ, and are more likely to be coordinative control groupings of recurrent task and muscle motifs (based on instrumentation) which may or may not relate to core compositionality in physiology. Is there any expectation that the framework should relate to core compositionality and physiology. This is not clear in the paper as written.

It would be useful to explore the approach with a range of neuromechanical models and controllers and simulated data to explore the issues I am raising and convince readers that this analysis framework adds clarity rather than dissolving the generalizability and interpretability of analyses in terms of underlying causal mechanisms.

The authors need to better frame their work in relation to causal analyses if they are claiming links to muscle synergies analyses and claim extension/refinement. Alternatively, these may not be linked, and instead parallel approaches exploring different hypotheses and goals using different organizational data descriptors.

To me this appears a data science tool that may not help any reductionist efforts and leads into less interpretable descriptions of motor control. Not invalid, but sufficiently different that common term use muddies the water.

Reviewer #3 (Public Review):

In this study, the authors developed and tested a novel framework for extracting muscle synergies. The approach aims at removing some limitations and constrains typical of previous approaches used in the field. In particular, the authors propose a mathematical formulation that removes constrains of linearity and couple the synergies to their motor outcome, supporting the concept of functional synergies and distinguishing the task-related performance related to each synergy. While some concepts behind this work were already introduced in recent work in the field, the methodology provided here encapsulates all these features in an original formulation providing a step forward with respect to the currently available algorithms. The authors also successfully demonstrated the applicability of their method to previously available datasets of multi-joint movements.

Preliminary results positively support the scientific soundness of the presented approach and its potential. The added values of the method should be documented more in future work to understand how the presented formulation relates to previous approaches and what novel insights can be achieved in practical scenarios and confirm/exploit the potential of the theoretical findings.

Strengths:

This work proposes a novel framework that addresses physiologically non-verified hypothesis of standard muscle synergy methods: it removes restrictive model assumptions (e.g. linearity, same mixing coefficients) and the reliance on variance-accounted-for (VAF) metrics.

The method is solid and achieves the prescribed objectives at a computational level and in preliminary laboratory data.

A toolbox is available for testing the methods on a larger scale.

The paper is well written and shows a high level of innovation, original content and analysis

Weaknesses:

Task performance variables could be specified in more quantitative definition in future work (e.g.: articular angles rather than a generic starting point- end point).

The paper does not show a comparison with previous approaches (e.g.: NMF) or recently developed approaches (such as MMF).

A discussion of the likely impact of the work on the field, and the utility of the methods and data to the community.

In this work, the effort of the authors aimed at developing the field is clear. It is fundamental to develop novel frameworks for synergy extraction and use them to make them more interpretable and applicable to real scenarios, as well as more adherent to recent findings achieved in motor control and neuroscience that are not reflected in the standard models. At the same time, muscle synergies are being used more and more in research but their impact in practical scenarios is still limited, probably because synergies have rarely been analyzed in a functional context. This paper shows a very in-depth analysis and a novel framework to interpret data that links to the task space from a functional perspective. I also found that the results on the datasets are very well commented but could expand more to show why using this framework is advantageous.

There are some key points for discussion that follow from this paper which can be described more, maybe in future work, and that might contribute to major developments in the field, including:

The understanding of how the separation between relevant (redundant and synergistic) and irrelevant synergies impact on synergy analysis in practical works;

Interpreting how different synergistic organizations described in this work allows to better describe data from real scenarios (e.g.: motor recovery of patients after neurological diseases);

Discussing in detail how the presented findings compare with standard algorithms such as NMF to determine the added value provided with this approach;

Describe how redundant synergies reflect real neural organization and - if their "existence" is confirmed - how they contribute to redesign the concept of muscle synergies and of modular/synergistic control in general.

Author Response

The following is the authors’ response to the original reviews.

eLife assessment

The study by O'Reilly and Delis provides a valuable data-driven framework for extracting task-related muscle synergies in a step towards the understanding and practical use of synergies in real scenarios (e.g., evaluation of patients in a clinical environment). The approach is incomplete since the authors did not compare their method with classical physiologically grounded approaches for assessing muscle synergies. In this sense, the comparisons with classical approaches would clarify if physiological assemblies were preserved and were not altered to incorporate task space variables. Despite limitations, the proposed framework would interest motor control and neural engineering researchers.

We thank the editors for the positive assessment of our work and appreciate their constructive feedback. In our revised manuscript, we believe we have sufficiently addressed the identified limitations by a) comparing our approach to existing physiologically-based methods, providing thorough comparisons of their respective outputs, b) applying it to a dataset of post-stroke participants to demonstrate that it can identify physiologically-interpretable markers of motor recovery and c) providing examples to demonstrate how readers can interpret the novel perspective introduced.

Reviewer #1 (Public Review):

The proposed study provides an innovative framework for the identification of muscle synergies taking into account their task relevance. State-of-the-art techniques for extracting muscle interactions use unsupervised machine-learning algorithms applied to the envelopes of the electromyographic signals without taking into account the information related to the task being performed. In this work, the authors suggest including the task parameters in extracting muscle synergies using a network information framework previously proposed. This allows the identification of muscle interactions that are relevant, irrelevant, or redundant to the parameters of the task executed.

The proposed framework is a powerful tool to understand and identify muscle interactions for specific task parameters and it may be used to improve man-machine interfaces for the control of prostheses and robotic exoskeletons.

With respect to the network information framework recently published, this work added an important part to estimate the relevance of specific muscle interactions to the parameters of the task executed. However, the authors should better explain what is the added value of this contribution with respect to the previous one, also in terms of computational methods.

We thank the reviewer for their constructive comments. We have adjusted the introduction section of the manuscript to better explain the added value of this framework over previous work. Specifically, we draw the reviewer’s attention to the following updated section of the introduction:

“In [11], we considered, key limitations among current approaches to muscle synergy analysis in extracting functionally relevant and interpretable patterns of muscle activity [12]. We proposed a combinatorial approach based on information- and network-theory and dimensionality reduction (the network-information framework (NIF)) that significantly improved the generalisability of the extraction process by, among others, removing restrictive model assumptions (e.g. linearity, same mixing coefficients) and the reliance on variance-accounted-for (VAF) metrics [12]. By determining the pairwise mutual information between muscles, this innovation paved the way for the appropriate mapping of muscular interactions to the task space. To elaborate on the significance of this development, the extraction of motor patterns in isolation of the task space comes at the expense of both functional and physiological relevance [12,13]. Furthermore, effective methods for mapping large-scale physiological dynamics to behaviour is a current gap across the neurosciences [14]. Thus, here we build on this work by, for the first time, directly including task space parameters during muscle synergy extraction. In doing so, we address these current research gaps, progressing muscle synergy research and successful engineering applications in a fruitful direction [12,15,16]. This enables us, in a novel way, to dissect the concept of the muscle synergy and therefore quantify interactions between muscle activations with shared or complementary functional roles. “

In general, the method proposed relies on several hyperparameters and cost functions that have been optimized for the specific datasets. A sensitivity analysis should be performed, varying these parameters and reporting the performance of the framework.

We thank the reviewer for this comment which enabled us to clarify a potential misunderstanding. Our proposed framework does not require setting or varying hyperparameters to optimise cost functions.

For model-rank specification, a modularity maximising cost-function is used which determines what partitioning of the networks results in maximal modularity. We have offered two alternative approaches using this cost-function which consistently converge on the same solution. To further ensure the representativeness of this solution, we also offer a consensus-based approach where we apply these alternative approaches to individual participant or task data, then group the collective partitions together and re-apply the approaches. One of these approaches (Equation 2.2) requires two hyperparameters, γ and ω, which adjust the intra- and inter- network layer resolutions. As stated in the manuscript, we set both of these parameters to 1, thus nullifying their presence in the cost-function and aligning our work with the classical notion of modularity. Across the two alternative approaches to model-rank specification, the solution is unique and data-driven and has a demonstratable generalisability across datasets.

The only other cost-function present in the framework is during dimensionality reduction, which is a standard loss function used across the muscle synergy analysis literature. Thus, the approach is essentially parameter-free and we now have mentioned this more explicitly in the manuscript:

“To empirically determine the number of components to extract in a parameter-free way, we then concatenated these adjacency matrices into a multiplex network and employed network community-detection protocols to identify modules across spatial and temporal scales (fig.3(D)) [29–32,44].”

“In its generalised multilayer form, the Q-statistic is given an additional term to consider couplings between layers l and r with intra- and inter-layer resolution parameters γ and ω (Equation 2.2). Here, μ is the total edge weight across the network and γ and ω were set to 1 in the current study for classical modularity [30], thus removing the need for any hyperparameter tuning.”

It is not clear how the well-known phenomenon of cross-talk during the recording of electromyographic muscle activity may affect the performance of the proposed technique and how it may bias the overall outcomes of the framework.

Indeed artifacts such as crosstalk are a standard issue across the EMG literature and may impact the performance of subsequent analyses where prevalent in the dataset. Crosstalk is expected to be present irrespective of the task and so should not affect redundant and synergistic muscle representations, however it could be present in the task-irrelevant muscle interactions extracted. Due to the prominence of long-range functional connections with the task-irrelevant representations extracted, we suggest that such artifacts are unlikely to have played a prominent role in the extracted patterns. Nonetheless, we have recognised this possibility with the following updated sentence in the Discussion section:

“Although distinguishing task-irrelevant muscle couplings may capture artifacts such as EMG crosstalk, our results convey several physiological objectives of muscles including gross motor functions [65], the maintenance of internal joint mechanics and reciprocal inhibition of contralateral limbs [20,50].”

Reviewer #2 (Public Review):

This paper is an attempt to extend or augment muscle synergy and motor primitive ideas with task measures. The authors idea is to use information metrics (mutual information, co-information) in 'synergy' creation including task information directly. My reading of the paper is that the framework proposed radically moves from attempts to be analytic in terms of physiology and compositionality with physiological bases, instead into more descriptive ML frameworks that may not support physiological work easily.

We thank the reviewer for taking the time to provide a thorough commentary on this manuscript. An overall aim in developing this framework is to build on other recent developments in providing a more fine-grained functional architecture underlying movement control [1,2]. It is a requirement for the successful communication and introduction of this toolbox to the field to provide readers with an understanding of how to use the framework and an intuition on how to interpret the results. Thus, we agree with the reviewer that functional interpretations are of crucial use.

We also agree with the reviewer that maintaining a physiological underpinning is a desirable direction for the field and should not be made secondary to functional descriptions. In our updated version of this manuscript, we have therefore included direct comparisons with the gold-standard in the field for muscle synergy extraction, namely non-negative matrix factorisation based muscle synergy extraction (see ‘Building on current approaches to muscle synergy analysis’ and fig.5-6 of revised manuscript) [3,4]. In these comparison, we show how our framework goes beyond this current approach in terms of functional insight while still maintaining physiological relevance. Indeed, in the revised manuscript we also include a fourth dataset comprising post-stroke participants and healthy controls (Fig.6). We demonstrate, through a simple example application to this dataset, how our proposed framework can produce more predictive representations of motor impairment than the gold-standard approach. The representations we identified were discriminative of motor impairment measured via the Fugl-Meyer assessment using just one trial per participant. This improves considerably upon the sensitivity of the current approach to altered motor patterns which have predominantly required many trials and participants to gain significance [5,6]. Thus, the patterns we extract are a more comprehensive representation of the actual underlying physiological state of the participants.

This approach is very different from the notions of physiological compositional elements as muscle synergies and motor primitives, and to me seems to really be striving to identify task relevant coordinative couplings. This is a meta problem for more classical analyses. Classical analyses seek compositional elements stable across tasks. These elements may then be explored in causal experiments and generative simulations of coupling and control strategies. The present work does not convince me that the joint 'meta' analysis proposed with task information added is not unmoored from physiology and causal modeling in some important ways. It also neglects publications and methods that might be inconvenient to the new framework.

We would be very interested in receiving the reviewer’s suggestions of existing approaches that we have not incorporated here and would be happy to discuss these in the revised manuscript.

Information based separation has been used in muscle synergy analyses using infomax ICA, which is information not variance based at core. Though linear mixing of sources is assumed, minimized mutual information is the basis.

We agree with the reviewer that ICA relies on information measures, however it does not incorporate task-space information. The novelty of our approach lies in the characterisation of muscle interactions with respect to the task at hand. If the reviewer could provide references to this statement, we would be able to consider this further.

Physiological causal testing of synergy ideas is neglected in the literature reviews in the paper. Although these are in animal work, the clear connection of muscle synergy choices and analyses to physiology is important and needs to be managed in the new methods proposed. Is any correspondence assumed? Possible?

We agree with τhe reviewer that this a crucial element of muscle synergy research and will aim to address it in our future work. However, we would like to point out that the current manuscript is a “tools and resources” article aiming to introduce a new framework. In our revised manuscript, we have incorporated an application of the framework to a dataset from post-stroke patients to demonstrate the use of the framework in clinical settings to identify biomarkers and use them to make predictions of motor recovery (see Fig.6 of updated manuscript).

Questions and concerns with the framework as an overall tool:

First, muscle based motor information sources have influences on different time scales in the task mechanics. Analyses of synergies in the methods proposed will be very much dependent on the number and quality of task variables included and how these are managed. Standardizing and comparing among labs, tasks sets and instrumentation differences is not well enough considered as a problem in this new proposed method toolset, at least in my reading. Will replication, and testing across groups ever be truly feasible in this framework?

We agree with the reviewer that this important point can be a limitation of the applicability of the framework. For this reason, we chose a “holistic” approach, applying the framework to several datasets collected in different settings, and selecting different kinds of task variables to extract muscle networks from. Crucially, we used a leave-one-task-out and leave-one-participant-out cross validation procedure to specifically address this point. Our results showed that the extracted couplings are robust irrespective of the task variable and/or participant excluded and this lends credit to the generalisability of the framework.

Muscle based motor information sources have influences on different time scales in the task mechanics. Kinematic analyses, dynamic analyses and force plate analyses of the same task may provide task variables that alter the results in the proposed framework it seems.

As we have mentioned above, here we used all the above types of task variables together to illustrate the range of measures that can be included in the proposed framework and showed that the outputs are robust to the exclusion of any task/participant. This point is especially evident for dataset 3 results, where high levels of generalisability were found despite the inclusion of kinematic, dynamic and IMU data (see Table 1. of original submission and updated manuscript). We believe that this is an advantage of the approach as it allows researchers to apply the method to different kinds of measurements they may have collected and gain insights into the relationships of muscle couplings with kinematic/dynamic/force parameters. This will also enable scientists to attribute different functional roles to the identified couplings and it is something we plan to do in future applications of the framework.

Second, there is a sampling problem in all synergy analyses. We cannot record all muscles or all task parameters. Examining synergies across multiple tasks seeks 'stationary' compositionality. Including task specific elements may or may not reinforce or give increased coordinative precision to the stationary compositionality.

We fully agree that this is a limitation of all synergy analyses and aimed to consider this study a step in the direction of addressing this limitation by providing the research community with a toolbox that can be used to quantify muscle couplings that can have different levels of task specificity.

To me the new methods proposed seem partly orthogonal to the ideas of stable compositionality. The 'synergies' obtained will likely differ, and are more likely to be coordinative control groupings of recurrent task and muscle motifs (based on instrumentation) which may or may not relate to core compositionality in physiology. Is there any expectation that the framework should relate to core compositionality and physiology. This is not clear in the paper as written.

In our new analysis, we have compared the proposed approach to existing physiologically-based methodologies and showed that the new framework can capture several salient physiological features of movement that the current NMF-based approach cannot. For example, as we have moved away from optimising variance accounted for metrics, our framework can identify subtle muscle couplings that have important functional roles. These subtle couplings are often not captured in current muscle synergy analysis as, against physiological relevance, higher amplitude muscles often take prominence. Further, by directly including task parameters during extraction, we can determine the muscles that have a functional role concerning the included task parameter rather than inferring this relationship indirectly using knowledge about the task executed. In our updated manuscript, by applying the framework to post-stroke participants (see Fig.6), we were also able to demonstrate that the extracted couplings are associated with functional parameters of motor recovery and have a clear link with the physiological state of individual participants.

It would be useful to explore the approach with a range of neuromechanical models and controllers and simulated data to explore the issues I am raising and convince readers that this analysis framework adds clarity rather than dissolving the generalizability and interpretability of analyses in terms of underlying causal mechanisms.

The authors need to better frame their work in relation to causal analyses if they are claiming links to muscle synergies analyses and claim extension/refinement. Alternatively, these may not be linked, and instead parallel approaches exploring different hypotheses and goals using different organizational data descriptors.

To address the reviewers concerns here, we have included in the updated manuscript a toy example simulating situations in which pairs of muscles would have a redundant or synergistic functional relationship (see Fig.2). This simulation gives clear intuition on situations where two muscles (e.g. an antagonist-agonist pair) may share functionally similar or complementary information about task direction (left vs right). In particular, within the main text describing this figure, we state how current NMF based approaches consider muscles functionally equivalent when they share similar magnitude activations, whereas our framework captures muscles with identical task information. Thus, our work is an extension of current approaches towards understanding causal mechanisms. The suggestion to use neuromechanical models is valuable, however we consider it beyond the scope of this work. This “Tools and Resources” paper is aimed at introducing the computational framework for the analysis of large-scale muscle couplings in task space. Our future work will use this framework to address unanswered questions in the field and we hope that it will be helpful for other scientists in testing their hypotheses.

To me this appears a data science tool that may not help any reductionist efforts and leads into less interpretable descriptions of motor control. Not invalid, but sufficiently different that common term use muddies the water.

We believe that the novel evidence we provided both on simulated and real data have contributed to a better interpretability of the approach outcomes. Specifically, we have introduced examples showing the functional roles of the different types of interactions as well as the predictive power of the outputs. Concerning the use of the term synergy, we have provided a clear description throughout the manuscript regarding the interpretation of synergy vs redundancy in the novel perspective we propose. For example in the discussion section:

“ We thus sought to provide greater nuance to the notion of ‘working together’ by defining motor redundancy and synergy in information-theoretic terms [6,56]. In our framework, redundancy and synergy are terms describing functionally similar and complementary motor signals respectively, introducing a new perspective that is conceptually distinct from the traditional view of muscle synergies as a solution to the motor redundancy problem [3,6,7]. In this new definition of muscle interactions in the task space, a group of muscles can ‘work together’ either synergistically or redundantly towards the same task. In doing so, the perspective instantiated by our approach provides novel coverage to the partitioning of task-relevant and -irrelevant variability implemented by the motor system along with an improved specificity regarding the functional roles of muscle couplings [20–22]. Our framework emphasises not only the role of functionally redundant muscle couplings that result from the underlying degeneracy of the motor system, but also of complementary, synergistic dependencies that are important for communication and integration across specialised neural circuitry [57,58]. Thus, the present study aligns the muscle synergy concept with the current mechanistic understanding of the nervous system whilst offering an analytical approach amenable to the continued advances in large-scale data capture [14,59].”

Reviewer #3 (Public Review):

In this study, the authors developed and tested a novel framework for extracting muscle synergies. The approach aims at removing some limitations and constrains typical of previous approaches used in the field. In particular, the authors propose a mathematical formulation that removes constrains of linearity and couple the synergies to their motor outcome, supporting the concept of functional synergies and distinguishing the task-related performance related to each synergy. While some concepts behind this work were already introduced in recent work in the field, the methodology provided here encapsulates all these features in an original formulation providing a step forward with respect to the currently available algorithms. The authors also successfully demonstrated the applicability of their method to previously available datasets of multi-joint movements.

Preliminary results positively support the scientific soundness of the presented approach and its potential. The added values of the method should be documented more in future work to understand how the presented formulation relates to previous approaches and what novel insights can be achieved in practical scenarios and confirm/exploit the potential of the theoretical findings.

Strengths:

This work proposes a novel framework that addresses physiologically non-verified hypothesis of standard muscle synergy methods: it removes restrictive model assumptions (e.g. linearity, same mixing coefficients) and the reliance on variance-accounted-for (VAF) metrics.

The method is solid and achieves the prescribed objectives at a computational level and in preliminary laboratory data.

A toolbox is available for testing the methods on a larger scale.

The paper is well written and shows a high level of innovation, original content and analysis

Weaknesses:

Task performance variables could be specified in more quantitative definition in future work (e.g.: articular angles rather than a generic starting point- end point).

We agree with this point and will incorporate it in future work. Our aim here was to show that the framework would work with any task variable and that scientists can use it to identify the relevance of muscle interactions to different types of task parameters.

The paper does not show a comparison with previous approaches (e.g.: NMF) or recently developed approaches (such as MMF).

We have now illustrated such a comparison on two datasets and explained more how the new framework can dissect the different types of muscle groupings (see ‘Building on current approaches to muscle synergy analysis’ section and Fig.5-6 of revised manuscript).

A discussion of the likely impact of the work on the field, and the utility of the methods and data to the community.

In our revised manuscript, we have introduced 2 new applications of the framework to real data to exemplify its use for a) functional interpretability and b) identification of biomarkers (see ‘Building on current approaches to muscle synergy analysis’ section and Fig.5-6 of revised manuscript). We also point towards its use in movement restoration and augmentation devices and in the clinical setting in the discussion section:

“The separate quantification of these muscle interaction types opens up novel opportunities in the practical application of muscle synergy analysis, as demonstrated in the current study through the identification of a significant predictor of motor impairment post-stroke from single-trials [5,12,65]. For instance, these distinct representations may encapsulate different neural substrates that can be specifically assessed at the muscle-level for the purpose of bodily restoration and augmentation [66]. Uncovering their neural underpinnings is an interesting topic for future research.”

In this work, the effort of the authors aimed at developing the field is clear. It is fundamental to develop novel frameworks for synergy extraction and use them to make them more interpretable and applicable to real scenarios, as well as more adherent to recent findings achieved in motor control and neuroscience that are not reflected in the standard models. At the same time, muscle synergies are being used more and more in research but their impact in practical scenarios is still limited, probably because synergies have rarely been analyzed in a functional context. This paper shows a very in-depth analysis and a novel framework to interpret data that links to the task space from a functional perspective. I also found that the results on the datasets are very well commented but could expand more to show why using this framework is advantageous.

There are some key points for discussion that follow from this paper which can be described more, maybe in future work, and that might contribute to major developments in the field, including:

The understanding of how the separation between relevant (redundant and synergistic) and irrelevant synergies impact on synergy analysis in practical works;

We have now introduced new figures (Fig. 5 and 6) to the revised manuscript, demonstrating simple applications of the framework and providing intuition regarding the outputs. We have also added points to the Discussion commenting on the differences between types of couplings and how they can be interpreted in future works:

“Our framework emphasises not only the role of functionally redundant muscle couplings that result from the underlying degeneracy of the motor system, but also of complementary, synergistic dependencies that are important for communication and integration across specialised neural circuitry [57,58]. Thus, the present study aligns the muscle synergy concept with the current mechanistic understanding of the nervous system whilst offering an analytical approach amenable to the continued advances in large-scale data capture [14,59].”

“Although distinguishing task-irrelevant muscle couplings may capture artifacts such as EMG crosstalk, our results convey several physiological objectives of muscles including gross motor functions [64], the maintenance of internal joint mechanics and reciprocal inhibition of contralateral limbs [19,49]. Thus, task-irrelevant muscle interactions reflect both biomechanical- and task-level constraints that provide a structural foundation for task-specific couplings. The separate quantification of these muscle interaction types opens up novel opportunities in the practical application of muscle synergy analysis, as demonstrated in the current study through the identification of a significant predictor of motor impairment post-stroke from single-trials [5,12,65]. For instance, these distinct representations may encapsulate different neural substrates that can be specifically assessed at the muscle-level for the purpose of bodily restoration and augmentation [66]. Uncovering their neural underpinnings is an interesting topic for future research.”

Interpreting how different synergistic organizations described in this work allows to better describe data from real scenarios (e.g.: motor recovery of patients after neurological diseases);

We have now added an example application of the framework to a dataset of stroke patients (Fig.6) and identified a redundant muscle patterns that are predictive of functional measures.

Discussing in detail how the presented findings compare with standard algorithms such as NMF to determine the added value provided with this approach;

As indicated above, we have now shown such a comparison on two new datasets (see Fig.5-6 of revised manuscript).

Describe how redundant synergies reflect real neural organization and - if their "existence" is confirmed - how they contribute to redesign the concept of muscle synergies and of modular/synergistic control in general.

This is an important point that we have now addressed more in our Discussion by relating redundant muscle couplings to degeneracy in the motor system and synergistic couplings to integrative dynamics by higher-level processes. We have also added a simple simulation illustrating how synergistic and redundant interactions co-exist and represent different contributions to task performance (see Fig.2 of revised manuscript).

Version 2

Abstract

Abstract

The muscle synergy is a guiding concept in motor control research that relies on the general notion of muscles ‘working together’ towards task performance. However, although the synergy concept has provided valuable insights into motor coordination, muscle interactions have not been fully characterised with respect to task performance. Here, we address this research gap by proposing a novel perspective to the muscle synergy that assigns specific functional roles to muscle couplings by characterising their task-relevance. Our novel perspective provides nuance to the muscle synergy concept, demonstrating how muscular interactions can ‘work together’ in different ways: a) irrespective of the task at hand but also b) redundantly or c) complementarily towards common task-goals. To establish this perspective, we leverage information- and network-theory and dimensionality reduction methods to include discrete and continuous task parameters directly during muscle synergy extraction. Specifically, we introduce co-information as a measure of the task relevance of muscle interactions and use it to categorise such interactions as task-irrelevant (present across tasks), redundant (shared task information) or synergistic (different task information). To demonstrate these types of interactions in real data, we firstly apply the framework in a simple way, revealing its added functional and physiological relevance with respect to current approaches. We then apply the framework to large-scale datasets and extract generalizable and scale-invariant representations consisting of subnetworks of synchronised muscle couplings and distinct temporal patterns. The representations effectively capture the functional interplay between task end-goals and biomechanical affordances and the concurrent processing of functionally similar and complementary task information. The proposed framework unifies the capabilities of current approaches in capturing distinct motor features while providing novel insights and research opportunities through a nuanced perspective to the muscle synergy.

Introduction

Human movement is a highly complex behaviour, with a broad spectrum of multiplexed spatiotemporal dynamics typically exhibited for basic activities-of-daily-living [1,2]. How the central nervous system controls movement in the face of this inherent complexity to ensure efficient and reliable navigation of the environment and task performance is a nontrivial question currently under investigation in the motor control field [3,4]. The muscle synergy hypothesis is a long-withstanding proposition on the underlying neural constraints producing coordinated movement, stating that this complexity is offset by the allocation of computational resources to the spinal-level in the form of motor primitives [3–6]. These motor primitives modularly activate functional groups of muscles that are flexibly combined for the efficient construction of a movement. The conceptual underpinning of the ‘muscle synergy’ entails the idea of combinations of muscles ‘working together’ for the purpose of effective goal-orientated behaviour [7]. This emergent cohesion involves the following qualifying attributes: a repeatable muscle activation pattern common across trials and participants, a reciprocal relationship among functional muscle groups such that changes occur in one group to compensate for changes in another, and task dependence (i.e. the pattern of interdependencies among muscles must map onto task performance) [7,8]. A common approach to analyse the neural constraints underlying these motor patterns is to apply unsupervised machine-learning algorithms to electromyographic (EMG) data [9,10], with the aim of extracting a latent, low-dimensional representation.

In [11], we considered, key limitations among current approaches to muscle synergy analysis in extracting functionally relevant and interpretable patterns of muscle activity [12]. We proposed a combinatorial approach based on information- and network-theory and dimensionality reduction (the network-information framework (NIF)) that significantly improved the generalisability of the extraction process by, among others, removing restrictive model assumptions (e.g. linearity, same mixing coefficients) and the reliance on variance-accounted-for (VAF) metrics [12]. By determining the pairwise mutual information between muscles, this innovation paved the way for the appropriate mapping of muscular interactions to the task space. To elaborate on the significance of this development, the extraction of motor patterns in isolation of the task space often comes at the expense of functional and physiological relevance [12,13]. Furthermore, effective methods for mapping large-scale physiological dynamics to behaviour is a current gap across the neurosciences [14]. Thus, here we build on this work by, for the first time, directly including task space parameters during muscle synergy extraction. This enables us, in a novel way, to dissect the concept of the muscle synergy and therefore quantify interactions between muscle activations with shared or complementary functional roles.

Further to the above, in its currently defined state, the muscle synergy concept describes the role of common neural drives to functional muscle groupings working redundantly towards a common task-goal [3–5,7,15]. However, recent influential works have highlighted several other important mechanisms involved in this low-dimensional control strategy that are not well recognised by the muscle synergy concept [16–24]. Such insights include the partitioning of motor variability by the nervous system into task-relevant and -irrelevant spaces and the cooperation between functionally distinct muscle groupings in the form of cross-module functional connectivities. These observations highlight the need for a refinement of the muscle synergy concept to comprehensively describe diverse muscle interactions during movement, including their partitioning into task-relevant and –irrelevant spaces and the characterisation of their functional roles.

We thus motivate the development of a more nuanced perspective to the muscle synergy concept and the general notion of ‘working together’ that comprehensively describes the muscle interactions underlying motor behaviour. To do so, we propose an information-theoretic approach (based on the NIF pipeline) that characterises the contributions of muscle couplings to task performance. In other words, we frame the notion of ‘working together’ in more specific terms of shared information between pairs of spatiotemporal muscle activations ([mx,my] (Fig.1.3(a))) (red and green sets respectively) and a corresponding task parameter (r) (blue set) (see Venn diagrams in Fig.1.3(b) below). Among current approaches to muscle synergy analysis, the shared information between muscles (yellow and white area in Fig.1.1(a)) is quantified, using dimensionality reduction, as common patterns of variability. These common patterns are essentially task-agnostic and may contain patterns of variability a) present in specific tasks (i.e. task-relevant (white shaded area in Fig.1.1(a)) as well as b) shared across tasks (i.e. task–irrelevant (yellow shaded area in Fig.1.1(a)). Our proposed approach dissects patterns of muscle variability in space, time and across trials in terms of their task-relevance and functional similarity in a generalizable manner using mutual information (MI) (Fig.1.3(a-c)). This enables us to decompose muscle activations into muscle pair – task parameter couplings and characterise their combined functional roles. We can then extract low-dimensional representations of these muscle couplings, i.e. muscle networks with specific spatial and temporal signatures, across participants and tasks (Fig.1.3(c)).

A general outline of the proposed approach. (1.1(a-b)) We propose a novel approach to mapping muscle couplings to the task space. Among current muscle synergy analysis approaches, muscle couplings are quantified in isolation of the task solely using dimensionality reduction. Using our approach, the functional characteristics of muscle interactions can be quantified in terms of the similarity of their encoded task information. We do so by determining the coupling between [mX, mY] and a corresponding task parameter (r) using mutual information (MI). From this perspective, task-redundant muscle couplings (pink shaded area in pink-orange intersection) represent muscles cooperating towards similar task goals, while task-synergistic muscle couplings (orange shaded area in pink-orange intersection) encapsulate the task information provided by a muscle pairing acting towards complementary task goals. Muscle couplings present across tasks (i.e., task-irrelevant) are quantified by conditioning the MI between [mX, mY] pairs with respect to r (yellow intersection). (1.2) A description of redundant and synergistic interactions. (a) Net redundant interactions are defined by a greater amount of information generated by the sum of individual observation of mX and mY ([mX + mY]) than their simultaneous observation ([mX, mY]). (b) In a net synergistic interaction, [mX, mY] provides more information than [mX + mY]. (1.3(a-c)) An overview of the approach. Spatiotemporal muscle activation samples are extracted across trials from large-scale EMG datasets and concatenated into vectors, forming [mX, mY] pairs.The derived muscle couplings are then run through the NIF pipeline [11], producing low-dimensional, multiplexed space-time muscle networks.

Crucially, using this novel framework, we can separately quantify the task-irrelevant (i.e. muscle interactions present across tasks) information conveyed by a muscle coupling (yellow intersection in Fig. 1.1(b)) from the task-relevant information (pink-and-orange shaded area in Fig.1.1(b)). These task-relevant interactions can be either sub-additive/redundant (i.e. the muscle coupling provides less information about the task compared to the sum of the individual muscle patterns) or super-additive/synergistic (i.e. the muscle coupling conveys more task information than the sum of individual muscle-task encodings). Conceptually, the information a muscle interaction provides is considered redundant when all the information can essentially be found in one of the muscles (pink shaded area (Fig.1.2(a)). This redundant task information thus reveals a functional similarity between muscle activations. Alternatively, we can also identify muscles that act synergistically towards complementary task goals, meaning their variations provide different information about a motor behaviour. A key, quantifiable attribute of this complementary interaction is the emergent task information (‘synergy’) they provide when considered together (orange shaded area (fig.1.2(b)). From this novel perspective, muscle activations can ‘work together’ not just similarly towards a common task-goal but also complementarily towards different aspects of motor behaviour and concurrently towards objectives functionally irrelevant to overt task performance, thus providing a comprehensive view of the muscle interactions governing coordinated movement.

To illustrate this novel conceptual and analytical framework, we conducted two example applications to data from human participants performing naturalistic movements. These applications demonstrate the added utility of this framework to current muscle synergy analysis in terms of functional and physiological relevance and interpretability. We then applied it to three large-scale datasets, extracting generalizable and functionally interpretable space-time muscle networks with respect to both discrete and continuous task spaces. We have also made available open-source Matlab routines for readers to apply this approach to their own data (https://github.com/DelisLab/EMG2Task).

Results

Our primary aim here is to characterise muscle synergies in task space by quantifying the contributions of muscle couplings to task performance. To achieve this, we essentially reverse the analytical approach typically used in muscle synergy studies (i.e. muscle groupings are identified and inferences then made about their functional roles) [9]. More specifically, we firstly identify functional couplings between paired muscle activations by evaluating their task-relevance and then extracting representative patterns of such couplings using dimensionality reduction methods. This enables us to distinguish task-irrelevant from task-relevant muscle couplings. Of the muscle couplings that demonstrate task-relevance, we can then characterise their functional roles as either redundant or synergistic. Fig.2 illustrates a simulation to facilitate interpretations of what informational redundancy and synergy mean when applied to muscle activities in the context of task performance. Redundant task information is generated when mx and my carry identical predictive information about τ. This is distinct from current muscle synergy analysis which would consider mx and my to share information about τ if their magnitudes are equivalent. Here, τ is always L when mx is on and when my is off and R when vice versa. Thus, all the task information can be found in either mx or my alone, generating 1 bit of redundant information. Synergistic task information on the other hand is predictive task information generated only when observing both mx and my together. In the simple example shown, τ can be L when both mx and my are active or inactive. However, we can see that when both muscles are active or inactive then τ is L. Thus, no predictive task information is provided by either mx or my alone but the full 1 bit of information available is generated when observing both muscles together.

A simulation demonstrating how informational redundancy and synergy can be interpreted when applied to the muscle space. Four observations of a given muscle pair ( and ) that can fall into two equiprobable on and off activation states and a corresponding task parameter ( ) describing left (L) or right (R) movement direction.

A summary of the NIF pipeline. (A) Large-scale datasets of EMG signals are captured while participants perform various motor tasks [25–27]. (B) The MI between all unique muscle-timepoint vector (combinations with respect to a corresponding task parameter ( ) is determined [28], forming a network of functional connectivities. (C) These adjacency matrices are then analysed in terms of statistical significance and modular structure using percolation theory [11]. (D) The optimal spatial and temporal model-ranks are determined using generalised, consensus-based network community detection methods [29–32]. (E) The optimal model-ranks are used as input parameters for dimensionality reduction, where space-time muscle networks along with their underlying activation coefficients are concurrently extracted [25].

Building on current approaches to muscle synergy analysis

Current approaches to muscle synergy analysis based on non-negative matrix factorisation (NMF) have a proven use case in the extraction of functionally- and physiologically relevant motor patterns [3,12,34]. To demonstrate that the proposed framework adds to this current utility, here we firstly provide a simple example output from the proposed and current approaches (see fig.5(A-D)). This example was derived from the EMG recordings of a single-trial of a participant walking on level-ground in the counter-clockwise direction around the circuit depicted in fig.4(C) [27]. For the proposed approach, the muscle couplings were determined with respect to a single, continuous task parameter, the heel kinematic marker in the anterior-posterior direction. For the application of the current approach, we applied the spatial muscle synergy model across the same single-trial EMG recordings [35], extracting one component.

Graphical illustrations of each of the datasets analysed in the current study. (A) Dataset 1 consisted of participants executing table-top point-to-point reaching movements across four targets in forward (P1-P4) and backwards (P5-P8) directions at both fast and slow speeds [25]. The muscles recorded included the finger extensors (FE), brachioradialis (BR), biceps brachii (BI), medial-triceps (TM), lateral-triceps (TL), anterior deltoid (AD), posterior deltoid (PD), pectoralis major (PE), latissimus dorsi (LD) of the right, reaching arm. (B) For dataset 2, the activity of 30 muscles was recorded while participants performed whole-body point-to-point reaching movements across three different heights and bars and in various directions, accumulating to 72 unique reaching tasks [26]. (C) The circuit navigated by participants in dataset 3 as they executed various locomotion modes is illustrated, of which level-ground walking, stair- and ramp-ascent/descent were analysed in the current study [27]. Several sub-conditions were undertaken by participants for each locomotion mode including different walking-speeds, clockwise vs. counter-clockwise direction, different stair heights and ramp inclines etc. Participants executed these tasks while the EMG of 11 muscles on the right leg ((Gluteus medius (GlutM), right external oblique (Obl), semitendinosus (ST), gracilis (GR), biceps femoris (BF), rectus femoris (RF), vastus lateralis (VL), vastus medialis (VM), soleus (SO), tibialis anterior (TA), gastrocnemius medialis (GM)) along with kinematic, dynamic and IMU signals were captured. (D) The EMG placement for dataset 4 [Deltoideus pars clavicularis (DC), Biceps brachii (BB), Triceps brachii (TB), Flexor digitorum superficialis (FDS), Extensor digitorum (ED), Brachioradialis (BR), Flexor carpi ulnaris (FCU), Extensor carpi ulnaris (ECU), Pronator teres (PT), Flexor carpi radialis (FCR), Abductor pollicus brevis (APB), Abductor digiti minimi (ADM)] [33]. A single-trial was taken from 25 healthy and 20 post-stroke participants performing a unilateral pointing movement with the index finger and arm outstretched (task 9 of the Softpro protocol (MHH)).

A simplified example output from the proposed framework applied to a single trial of turning gait from Dataset 3. (A) Task-irrelevant, (B) Task-redundant and (C) Task-synergistic synchronous muscle couplings were quantified with respect to the heel kinematic marker (anterior-posterior direction). Human body models accompanying each spatial network illustrate their respective submodular structure with node colour and size and edge width indicating community affiliation [29], network centrality and connection strength respectively [36,37]. (D) A corresponding synergy representation from a single trial of turning gait from dataset 3 extracted using the spatial model from current approaches [35]. Each bar represents the relative weighting of each muscle in the synergy component.

The simplified representations from the proposed approach reveal the functional role of muscular interactions with respect to the heel marker and provide intuition on the types of muscle couplings that can be identified, including task-irrelevant (A) -redundant (B) and –synergistic (C) interactions. Their submodular structure are illustrated via the node colour on the accompanying human body models [36], describing muscles that have a closer functional relationship. For example, a) the hamstring muscles (ST and BF) controlling knee flexion work redundantly together with muscles involved in hip abduction (GR, GlutM and Obl) to move the heel around the circuit (Fig.4(B)), and b) calf muscles involved in ankle flexion (GM, TA) cooperatively determine heel position in synergy with the same hamstring muscles (Fig.5(C)). Moreover, these hamstring muscles together with their antagonist RF also form a task-irrelevant network, i.e. their couplings are not predictive of heel position (Fig.5(A)). Overall, task-irrelevant muscle couplings primarily capture interactions between co-agonist and agonist-antagonist muscles that are indiscriminative of heel marker position. Also, task-redundant and -synergistic couplings reveal functionally similar and dissimilar muscle combinations respectively that provide sub-additive (i.e. shared or redundant) and super-additive (i.e. complementary or synergistic) information about heel marker position.

The NMF representation (fig.5(D)) conveys important information about gait in a task-agnostic manner, thus it may contain both task-relevant and -irrelevant interactions. Intuitively, SO, GM and Obl of the outer, right-side are most prominently weighted here, perhaps representing their functional role in accelerating the body in the counterclockwise direction around the circuit whilst maintaining upright posture (fig.4(C)). This observation can only be inferred indirectly however as, amongst current approaches, no direct association is made with other muscles or with task performance. Indeed, with respect to heel position, the proposed approach reveals that SO is functionally similar to Obl and GM whilst also containing task-synergistic and -irrelevant information with GM (fig.5(A-C)). From the perspective of existing approaches, the knee flexors and extensors play a minor role during turning gait (as indicated by their relatively low weighting (fig.5(D))). However, the proposed approach conveys a central role for these muscles suggesting that the proposed framework enables targeted dissections of muscle functionality with respect to a chosen task parameter, thus revealing subtle couplings with potentially important behavioural consequences.

Next, to demonstrate the additional physiological relevance the proposed methodology brings to current muscle synergy analysis, we applied the proposed framework to single trials of pointing movements [33]). Specifically, we determined MI from a randomly selected trial of healthy and post-stroke participants with respect to the 3D position of the anterior wrist kinematic marker (WRBA) of the pointing arm. We chose WRBA as the task variable here due to its sensitivity to hand orientation. For the purpose of this simplified demonstration, we focused on a comparison between task-redundant (with respect to WRBA) and NMF-based muscle representations. We firstly generated a normative representation of this pointing motion by extracting the first component across the healthy controls using the proposed approach (Fig.6(a.1)) and NMF (Fig.6(a.2)). We then quantified the similarity of muscle representations extracted from each post-stroke participant individually to this normative reference (obtained across all healthy participants) using Pearson’s correlation and converted these values to distances (i.e. 1-r) (Fig.6(b-c)). Finally, we determined if these distances from healthy control values were predictive of the stroke survivor’s motor impairment, measured using the upper-extremity section of the Fugl-Meyer assessment (FMA-UE) (Fig.6(d)).

A simple demonstration of the physiological relevance of the proposed approach (a.1-d-1) and the traditional, NMF-based approach (a.2-d.2). From dataset 4 [33], we took the EMG signals and WRBA kinematic from 20 post-stroke and 25 healthy participants. We extracted a single normative reference of healthy controls task-redundant muscle couplings with respect to WRBA (a.1) and a corresponding normative reference using NMF only (a.2). We then extracted a single component from each post-stroke participant and compared them individually with the corresponding normative reference, computing distance values (1-r) (b-c). We finally determined the predictive relationship of these distance values with a measure of upper-extremity motor impairment derived from the Fugl-meyer assessment (FMA-UE) (d).

To briefly summarise the results, the distance of post-stroke participants from healthy controls was NMF-based approach (= 1.4±1.39, p=0.33). This finding suggests, intuitively, that the proposed approach captures redundant muscle couplings that support robust motor control and that deviations from this normative pattern of motor redundancy are linearly related to the degree of impairment. Importantly, this result was obtained using only one randomly selected trial for each participant. This simple example conveys the physiologically-relevant targeted insights that can be generated from the proposed framework. Although current approaches have demonstrated significant linear relationships with motor impairment [38–40], these assessments generally rely on large numbers of trials and participants and don’t point to specific underlying muscle interactions as provided here.

Together, through two example applications, we have demonstrated attributes of the proposed methodology that provide novel capabilities to current muscle synergy analysis. In the following, we sequentially present in more detail the three types of muscular interactions in the spatiotemporal domain across three datasets and then show the robustness of the approach and its outputs.

Representations of motor behaviour in muscle couplings

To begin, we derived pairs of muscle activation vectors x .y from benchmark datasets of human participants executing naturalistic movements, namely arm reaching (dataset 1), whole-body point-to-point reaching (dataset 2), and various locomotion modes (dataset 3) [25–27,33] (see Fig.4 for the experimental design of each dataset and ‘Materials and methods’ for an outline of the experimental setups and EMG data pre-processing) (fig.3(A)). For datasets 1 and 2, we determine the MI between x .y vectors with respect to several discrete task parameters representing specific task attributes (e.g. reaching direction, speed etc.), while for dataset 3 we determine the MI with respect to corresponding kinematic, dynamic and inertial motion unit (IMU) features sampled across trials.

Having extracted the muscle pair-task interdependencies representing a specific intersection in fig.1.1(b), we next sought to find a parsimonious representation of motor behaviour that is consistent across tasks and participants for dataset 1-3 (fig.3(E)) [11,25]. To produce this sparse, low-dimensional representation, we undertook the following intermediary steps:

We modelled the MI values as adjacency matrices in the spatial or temporal domain and identified dependencies that were statistically significant using percolation theory (fig.3(C)) [41,42]. By assuming the muscle networks operate near a state of self-organised criticality [43], we effectively isolated dependencies that were above chance-level occurrence, thus empirically sparsifying the networks.

To empirically determine the number of components to extract in a parameter-free way, we then concatenated these adjacency matrices into a multiplex network and employed network community-detection protocols to identify modules across spatial and temporal scales (fig.3(D)) [29–32,44]. Having detected the spatiotemporal modular structure, we then returned the sparsified networks to their original format and used the number of modules identified as input parameters into dimensionality reduction (fig.3(E)) [25].

By optimising a modularity-maximising cost-function [45,46], the community detection protocols we employed consistently identified three spatial (S1-S3) and two temporal (T1-T2) modules as representative of the underlying task-redundant, -synergistic and -irrelevant informational dynamics. Following their extraction, we further analysed the spatial networks from each dataset in terms of their submodular structure by applying network-theoretic tools [29,32,37]. In doing so, we identified subnetworks within each spatial network and interesting patterns of network centrality, i.e. the relative importance of a node in a network. The spatial and temporal networks of each dataset output are illustrated in panels A and B (fig.7-12) of the following sections. They are accompanied by human body models where node colour and size indicate subnetwork community affiliation and network centrality respectively [36]. The networks we extracted operate in parallel within spatial and temporal domains while having an all-to-all correspondence across domains, i.e. any spatial component can be combined with any temporal component via a task-specific coefficient (illustrated in panel C for dataset 1 and 2 outputs and in the supplementary materials (fig.4-6) for dataset 3) [25]. Unlike similar muscle synergy extraction approaches, dimensionality reduction in the NIF pipeline doesn’t seek to approximate the variance of recorded EMG data but to identify sets of muscles that share the same type of interaction. Thus, the multiplexing coefficients extracted in this framework are instead interpreted as the participant- and task-specific scaling of information overlap.

Three spatial (S1-S3) and two temporal task-irrelevant muscle networks (T1-T2) were empirically identified and extracted across participants and task parameters from dataset 2 using the NIF pipeline (Panel A-B) [11,26]. (Panel C) Activation coefficients are presented to the right of the networks, indicating their task parameter-specific scaling averaged across participants. Human body models accompanying each spatial network illustrate their respective submodular structure with node colour and size and edge width indicating community affiliation [29], network centrality and connection strength respectively [36,37].

Task-irrelevant muscle couplings

To quantify the task-irrelevant contributions of muscular interactions to motor behaviour, we conditioned the MI between [mx, my] with respect to τ (see ‘Materials and methods’ section) [28]. This conditioning effectively removes the task-relevant information, leaving information produced by pairwise muscle variations that are task-indiscriminative. Following a run through the NIF pipeline (fig.3) [11], the output from dataset 2 and 3 are presented in fig.7-8 respectively while the output from dataset 1 is presented in the supplementary materials (fig.1). The task-irrelevant space-time muscle networks we extracted from dataset 1 and 2 shared several structural features with their task-agnostic counterparts extracted in the preliminary study [11], supporting recent work showing that functional muscle network structure are heavily influenced by task-irrelevant factors such as anatomical constraints [47]. The functional connectivities identified here captured known contributions of spatiotemporal muscular interactions to aspects of motor behaviour common across tasks and participants which we outline briefly here.

The temporal networks from dataset 1 and 2 captured mostly co-activations from movement onset – mid-movement and movement cessation, indicating that some co-contraction mechanisms were consistently task-irrelevant across trials. The temporal networks for dataset 2 were more diffuse compared to dataset 1, probably reflecting the more variable role of passive forces in generating movements to different heights captured in this dataset’s experimental design [26,48]. Furthermore, this co-contraction mechanism was more parsimoniously represented as a single network in dataset 3 (T1 in fig.8B), where passive forces in contrast likely played a consistently resistive role during locomotion. Interestingly, T1 for dataset 3 corresponded equivalently high for all three task spaces when corresponding with S2 which consisted of upper-leg extensors (see Supplementary materials (fig.4)). Muscle couplings indicative of agonist-antagonist pairings were also identified as separate subnetworks in S3 of dataset 3 (fig.8A). More specifically, their functional segregation appeared to be based on their distinct functional roles in forward propulsion (red nodes) and deceleration (blue nodes) during the mid-stance phase of gait, as indicated by the prominent correspondence with T2 across task spaces (see Supplementary materials (fig.4)). This finding reflects the consistent agonistic and antagonistic contributions of muscular interactions across locomotive tasks.

Three spatial (S1-S3) and two temporal task-irrelevant muscle networks (T1-T2) were empirically identified and extracted across participants and task parameters from dataset 3 using the NIF pipeline (Panel A-B) [11,27]. Activation coefficients are presented in supplementary materials (fig.4), indicating their task parameter-specific scaling averaged across participants in the dynamic, IMU and kinematic spaces. Human body models accompanying each spatial network illustrate their respective submodular structure with node colour and size and edge width indicating community affiliation [29], network centrality and connection strength respectively [36,37].

The gross motor function of muscle couplings was another characteristic of task-irrelevant muscle couplings that pervaded across the datasets analysed here. For instance, AD had a central role in S2 of dataset 1 while also displaying a unique pattern of connectivity with tibial musculature in S3 of dataset 2. Similarly, GlutM had a central role in S1 of dataset 3 (fig.8A). We further found a common pattern of task-irrelevant connectivity in S2 across datasets, namely the musculature about a hinge joint (elbow in dataset 1 and 2, knee in dataset 3) coupled with proximal shoulder or hip musculature, indicative of their biomechanical affordances. Finally, the passive, left-arm was connected with the tibial musculature of S3 in dataset 2 (green nodes (Fig.7(A)). To probe the underlying function of this connectivity in the left-arm, we inspected the original EMG signals. We observed periodic, tonic activations across tasks, reflective of reciprocal inhibition of contralateral limb musculature that enables unilateral movement [49].

Task-redundant muscle couplings

To characterise the functional role of task-relevant muscle couplings, we employed a higher-order information-theoretic concept known as co-information (co-I) [28,50,51]. This metric quantifies the MI between three random variables and may take on positive values (net synergistic) and negative values (net redundant) (see Fig.13 of ‘Materials and methods’ section). Co-I quantifies the task-relevant information shared between [mx, my] independently of the information generated by task-irrelevant muscular interactions. In doing so, it also defines the functional relationship between [mx, my] overall as redundant or complementary. Following the quantification of co-I for all [mx, my] and corresponding τ (pink area in orange-and-pink intersection (fig.1.3(B)), we parsed the negative values indicating redundancy into a separate matrix and rectified them. In fig.9-10, we illustrate the output following the extraction of task-redundant space-time muscle networks from datasets 2-3 across tasks and participants respectively while the output for dataset 1 is presented in supplementary materials fig.2. In the co-I formulation, task-redundant muscle couplings can be interpreted as muscle couplings that overall shared a common task-relevant functional role. For example, with reference to Figure 9 here, muscles in the networks presented in S1 (fig.9A) carry redundant information about the movement endpoint (fig.9C) with the temporal profile T1 (fig.9B) whereas S3 (fig.9A) contains muscle networks carrying redundant information about the starting point (fig.9C) with the same temporal profile T1 (fig.9B).

Three spatial (S1-S3) and two temporal task-redundant muscle networks (T1-T2) were empirically identified and extracted across participants and task parameters from dataset 2 using the NIF pipeline (Panel A-B) [11,26]. (Panel C) Activation coefficients are presented to the right of the networks, indicating their task parameter-specific scaling averaged across participants. Human body models accompanying each spatial network illustrate their respective submodular structure with node colour and size and edge width indicating community affiliation [29], network centrality and connection strength respectively [36,37].

Both dataset 1 (supplementary materials fig.2) and dataset 3 (fig.10) outputs display similar patterns of muscle couplings at the same spatial scale of an individual, task-relevant limbs’ musculature, with an emphasis on the coupling of specific muscles with all other muscles. For dataset 1, FE, BI and BR displayed this integrative pattern across S1-S3 respectively while BF, TA and ST demonstrated this pattern in dataset 3 also. The muscle networks encapsulated several functionally interpretable couplings such as the agonist-antagonist pairings of the BI and TM of S2 (dataset 1 (supplementary materials fig.2(A)) and the task redundancy of ankle dorsi-flexors, and knee/hip flexors during sloped walking for example in S2 of dataset 3 (fig.10(A)) [52]. The functional interpretation of these muscle connectivity patterns was in line with the extracted task-specific activations. For instance, S2 of dataset 1 was modulated most prominently by reaching direction when corresponding with T2, commensurate with the biomechanical affordances of this upper-arm muscle network. Furthermore, S2 of dataset 3 was specifically modulated by the right-thigh kinematic marker along the y-axis (up-down direction) for both T1 and T2 (see Supplementary materials (fig.5)). The centrality of task-redundant muscle couplings in dataset 1 and 3 suggests particular muscle activations drive the task-specific variations in the reaching arm and stepping leg muscle activities towards a common behavioural goal. It is also worth noting that the magnitude of these functional connectivity patterns appeared to be proportional to anatomical distance, as evidenced by the magnitude of connection strengths, a finding supportive of previous related research [47].

Three spatial (S1-S3) and two temporal task-redundant muscle networks (T1-T2) were empirically identified and extracted across participants and task parameters from dataset 3 using the NIF pipeline (A-B) [11,27]. Activation coefficients are presented in supplementary materials document 2 (fig.2), indicating their task parameter-specific scaling averaged across participants in the dynamic, IMU and kinematic spaces. Human body models accompanying each spatial network illustrate their respective submodular structure with node colour and size and edge width indicating community affiliation [29], network centrality and connection strength respectively [36,37].

Meanwhile at the greater spatial scale of dataset 2 (fig.9(A))), task-redundant muscle couplings were anatomically compartmentalised to the upper- and lower-body. This functional segregation was emphasised at the subnetwork level also, where the upper- and lower-body musculature of S3 for instance formed distinct submodules (blue and red nodes). Amongst the task-specific activations in dataset 2, S1 carried redundant task information about end-point target, -height and up-down direction when corresponding with T1. T2 for dataset 2 on the other hand contained mostly temporally proximal dependencies along the diagonal, suggestive of co-contraction mechanisms, which became more diffuse near movement cessation. These endpoint trajectory and co-contraction related temporal patterns were qualitatively similar to T1 and T2 of both dataset 1 and 3 respectively (see fig.10 and Supplementary materials fig.2 respectively).

Task-synergistic muscle couplings

Similarly, we isolated task-synergistic muscle couplings by parsing, instead, the positive co-I values from the computations conducted across all [mx, my] and corresponding into a sparse matrix where all redundant couplings were set to zero (see Fig.13 of ‘Materials and methods’ section). Task-synergistic muscle couplings here can be interpreted as a [mx, my] pair that provide complementary (i.e. functionally dissimilar) task information, thus more information is gained by observing [mx, my] together rather than separately (orange area in orange-and-pink intersection (fig.1.3(B)). In Fig.11-12, we illustrate the task-synergistic space-time muscle networks from datasets 2-3 respectively (dataset 1 output is presented in the supplementary materials fig.3).

Three spatial (S1-S3) and two temporal task-synergistic muscle networks (T1-T2) were empirically identified and extracted across participants and task parameters from dataset 2 using the NIF pipeline (Panel A-B) [11,26]. (Panel C) Activation coefficients are presented to the right of the networks, indicating their task parameter-specific scaling averaged across participants. Human body models accompanying each spatial network illustrate their respective submodular structure with node colour and size and edge width indicating community affiliation [29], network centrality and connection strength respectively [36,37].

Across datasets, the muscle networks could be characterised by the transmission of complementary task information between functionally specialised muscle groups. The overall structure of the synergistic muscle networks was similar to their task-redundant counterparts. For instance, the emphasis of an individual muscle’s connectivity with all other muscles was evident among synergistic couplings in dataset 3 (see VM, VL and RF of S1, GR, TA and SO of S2 and GlutM of S3 in fig.12A). This correspondence with task-redundant networks demonstrates that parallel and synchronous exchanges of redundant and synergistic task information underlie task-specific variations across trials (e.g. S3 of dataset 2 in fig. 9A and fig.11A). Interestingly, despite the similarity between the redundant and synergistic muscle networks, the way they are combined to encode task information differs depending on the type of interaction (synergistic vs. redundant, e.g. panel C of fig.11 in comparison to panel C of fig.9).

Concerning the temporal activations of these networks, the task-synergistic structure of dataset 2 (fig.11B) was also relatively unchanged compared to the task-redundant structure (fig.9B). This suggests that the task end-goal and co-contraction related mechanisms provided both redundant and synergistic task information concurrently during whole-body reaching movements. In contrast, a different view of synergistic information exchange is provided in datasets 1 and 3 (supplementary fig.3 and fig.12 here respectively), where T1 and T2 consist of more idiosyncratic activations that together appear to reflect the task end-goal related patterns found elsewhere. More specifically, in both datasets we found two distinct patterns of task-end goal related activity where early and late timepoints during movement initiation operated in parallel to provide complementary task information (see fig. 12B and supp. fig. 3B).

Three spatial (S1-S3) and two temporal task-synergistic muscle networks (T1-T2) were empirically identified and extracted across participants and task parameters from dataset 3 using the NIF pipeline (Panel A-B) [11,27]. Activation coefficients are presented in supplementary materials (fig.6), indicating their task parameter-specific scaling averaged across participants in the dynamic, IMU and kinematic spaces. Human body models accompanying each spatial network illustrate their respective submodular structure with node colour and size and edge width indicating community affiliation [29], network centrality and connection strength respectively [36,37].

Co-I determines the difference between the sum total information shared with r in mx and my when observed separately and the information shared with r when they are observed together. The adjacency matrices show how this calculation is carried out for all unique [mx, my] combinations. Redundant and synergistic muscle couplings are then separated into two equivalently sized networks. The accompanying colour bars indicate the values present in the adjacency matrix.

Generalisability of the extracted space-time muscle networks

To ascertain the generalisability of the extracted representations presented here beyond any subset of the input data, we conducted a similarity analysis through a leave-n-out cross validation procedure. In more detail, we compared the space-time networks extracted from the full dataset (illustrated in fig.4-9 and supplementary materials here) to the networks extracted from a subset of the data (see ‘Materials and methods’ section). Across datasets, a high level of concordance was found on average (∼0.9 correlation, see table1). This trend was evident across all datasets and for task-redundant, -synergistic and -irrelevant spatial and temporal networks. Dataset 1 and 2 findings demonstrate that the extracted networks are generalisable beyond any individual participant or task. Dataset 3 results go further by demonstrating that the extracted patterns are generalizable beyond any randomly selected and randomly sized subset of the input data. The highest correlations on average were consistently among temporal networks, replicating previous findings [11,53,54]. Although the spatial networks demonstrated a lower average correlation, this was substantially higher compared to previous applications of datasets 1 and 2 [11,25,53], suggesting that the inclusion of task parameters here captures the inter-participant differences more effectively. When comparing representations extracted from each continuous task space in dataset 3, kinematic features consistently had the highest average correlation and lowest variability compared to dynamic and IMU feature spaces. These findings support our change in the interpretation of the extracted activation coefficients away from conventional approaches where representational biases towards particular participants and/or tasks are inferred.

To further probe how the underlying assumption of an all-to-all correspondence between spatial and temporal representations made by sNM3F influenced the generalisability of the extracted networks, we compared its performance to non-negative Canonical-Polyadic (CP) tensor decomposition, which assumes an opposing one-to-one correspondence between components. We found that although CP has demonstrated a considerable capacity to de-mix neural data into simplified and interpretable low-dimensional components [55], its application here resulted in poor generalisability of the extracted patterns (∼0.5 correlation on average). This finding suggests that the all-to-all correspondence implemented by sNM3F identifies a more generalizable representation and should be favourably considered in future applications of this framework.

Discussion

The aim of the current study was to dissect the muscle synergy concept and offer a novel, more nuanced perspective on how muscles ‘work together’ to achieve a common behavioural goal. To do so, we introduced a computational approach based on the NIF pipeline (fig.3), enabling the effective decomposition of muscle interactions and the comprehensive description of their functional roles. Through the direct inclusion of task parameters in the extraction of muscle synergies, a novel perspective was produced where muscles ‘work together’ not just towards a common task-goal, but also concomitantly towards complementary task objectives and lower-level functions irrelevant for overt task performance. The functional architectures we uncovered were comprised of distributed subnetworks of synchronous muscle couplings and driven by distinct temporal patterns operating in parallel. Example applications to simple real and simulated EMG datasets revealed the additional capabilities provided by the proposed framework to current muscle synergy analysis in terms of functional and physiological relevance and interpretability. When applied to large-scale data, the proposed methodology extracted representations scale-invariant to dataset complexity and motor behaviours whilst being generalizable beyond any subset of the data. We thus present this framework as a useful analytical approach for mechanistic investigations on large-scale neural data through this novel perspective to the muscle synergy.

The ‘muscle synergy’ is a major guiding concept in the motor control field [3–7,15], providing a conceptual framework to the problem of motor redundancy that centres around the general notion of ‘working together’. In its current conceptualisation, ‘working together’ describes how the nervous system functionally groups muscles in a task-dependent manner through common neural drives to simplify movement control. This idea has undergone continued refinement since its early conception [3,6,7], with a notable progression being the introduction of the qualifying attributes: a sharing pattern, reciprocity and task dependence [7]. Nonetheless, recent influential works revealing other important mechanisms for the simplification of motor control highlight the generality of the current perspective offered by this concept [16–24]. We thus sought to provide greater nuance to the notion of ‘working together’ by defining motor redundancy and synergy in information-theoretic terms [6,56]. In our framework, redundancy and synergy are terms describing functionally similar and complementary motor signals respectively, introducing a new perspective that is conceptually distinct from the traditional view of muscle synergies as a solution to the motor redundancy problem [3,6,7]. In this new definition of muscle interactions in the task space, a group of muscles can ‘work together’ either synergistically or redundantly towards the same task. In doing so, the perspective instantiated by our approach provides novel coverage to the partitioning of task-relevant and - irrelevant variability implemented by the motor system along with an improved specificity regarding the functional roles of muscle couplings [20–22]. Our framework emphasises not only the role of functionally redundant muscle couplings that result from the underlying degeneracy of the motor system, but also of complementary, synergistic dependencies that are important for communication and integration across specialised neural circuitry [57,58]. Thus, the present study aligns the muscle synergy concept with the current mechanistic understanding of the nervous system whilst offering an analytical approach amenable to the continued advances in large-scale data capture [14,59].

Among current approaches to muscle synergy analysis, the established synchronous, temporal and time-varying muscle synergy models are understood to each characterise unique motor features [60]. More specifically, the synchronous model captures agonist-antagonist muscle pairings, the temporal model decomposes EMG signals into functionally distinct temporal phases and the task-specific modulation of spatiotemporal invariants are quantified in the time-varying model. In a unifying framework, here we quantify space-time muscle networks that concurrently capture many of these salient motor features in a holistic and principled way whilst mapping their functional consequences to motor behaviour. These salient features included, among others, agonist-antagonist pairings and functionally meaningful inter-limb couplings that consistently appeared across task-redundant, -synergistic and -irrelevant spaces. Thus, in dissecting the muscle interactions governing coordinated movement, our framework revealed their parallel and synchronous processing of functionally similar, complementary, and task-irrelevant information. This insight aligns with several recent works also demonstrating this distributed neural architecture of parallel information processing units [1,57,58].

Our framework also revealed novel characteristics of the motor system. For instance, the task-redundant and -synergistic networks we extracted appeared to be structured around the coupling between a prime-mover muscle and several supporting muscles, supporting recent work showing the nonhomogeneous sharing of neural drives within modules [61]. These novel spatial characteristics were driven by parallel temporal patterns representing endpoint trajectory and co-contraction related mechanisms, an insight supportive of recent work showing their parallel innervation [16,48,62]. Together, these representations encapsulated the functional interplay between task end-goal requirements and biomechanical affordances, a dynamic frequently highlighted in object manipulation experiments [63]. In other words, the task-relevant networks reflected how muscles ‘work together’ both redundantly and synergistically towards a desired end-goal state whilst, in parallel, continually controlling the present trajectory of the system. Meanwhile, the task-irrelevant networks demonstrated that these muscle couplings also work concomitantly towards lower-level objectives assigned mostly during the transition to this desired end state. Although distinguishing task-irrelevant muscle couplings may capture artifacts such as EMG crosstalk, our results convey several physiological objectives of muscles including gross motor functions [64], the maintenance of internal joint mechanics and reciprocal inhibition of contralateral limbs [19,49]. Thus, task-irrelevant muscle interactions reflect both biomechanical- and task-level constraints that provide a structural foundation for task-specific couplings. The separate quantification of these muscle interaction types opens up novel opportunities in the practical application of muscle synergy analysis, as demonstrated in the current study through the identification of a significant predictor of motor impairment post-stroke from single-trials [5,12,65]. For instance, these distinct representations may encapsulate different neural substrates that can be specifically assessed at the muscle-level for the purpose of bodily restoration and augmentation [66]. Uncovering their neural underpinnings is an interesting topic for future research.

Indeed, in future work, we aim to complement this study’s combinatorial perspective to the muscle synergy by dissecting the unique contribution of individual muscles to motor behaviour and how they may work independently towards task performance (see magenta and cyan intersections in fig.1.3(B)). More broadly, our work here parallels related information-theoretic approaches to decomposing task-relevant brain activity [51,67], whilst addressing a current research gap across the neurosciences in effective methods to mapping large-scale, spatiotemporal neural activity to behaviour [14,59]. Future applications of this framework should include large-scale, multi-modal data captured from participants performing a wide range of natural behaviours.

In sum, this study introduced a novel perspective to the muscle synergy concept and a computational framework to extract muscle couplings that map their pairwise contributions to motor behaviour. We suggest that this approach offers novel research opportunities for investigating the underlying neural constraints on motor behaviour and the fundamental structure-function relationships generated by agent – environment interactions [15,68].

Materials and Methods

Quantifying muscle couplings in the task space

To quantify muscle couplings we used MI, a non-linear measure of dependence that captures any type of relationship between random variables. Here to estimate MI, we used a Gaussian copula-based approximation [28]. This semi-parametric estimator exploits the equivalence between MI and the negative entropy of the empirical copula (c), a function that maps a multivariate set (e.g. [mx, my] representing activities of muscles X and Y) to their joint distribution (Equation 1.1) [28]. Thus, to determine task-irrelevant muscle couplings (I(mx; my|τ)), we conditioned the negative entropy of the empirical copula for [mx, my] with respect to a task variable τ (Equation 1.2). As mentioned, [mx, my] are continuous vectors composed of individual muscle amplitudes at specific timepoints across trials while is a corresponding discrete (e.g. movement direction for datasets 1 and 2) or continuous (e.g. movement kinematics in dataset 3) task parameter. For discrete task variables, τ takes one value for each trial and the MI is calculated across trials using a Gaussian mixture model [28]. In the case of continuous task variables, τ varies in time within a specific trial. Thus, we compute MI at each timepoint t using the muscle activity mx(t) and the task variable value τ(t) at this time point using a closed-form parametric estimator [28]. To evaluate the task-relevance of the identified muscle couplings, we used a higher-order information theoretic measure known as co-information (co-I) (Equation.1.3) (fig.11), which quantifies the relationship between three random variables, here [mx, my], and τ. Co-I implements the inclusion-exclusion principle of combinatorics [50], whereby the sum of MIs between individual m vectors and τ(I(mx; τ) = I(my; τ)) is compared against their composite MI (mx, my; τ) as follows: Negative I(mx; my;τ) corresponds to a net redundant coupling between [mx, my] about τ while positive I(mx; my;τ) indicates a net synergy. To analyse these distinct couplings separately, we parsed redundant and synergistic I(mx; my;τ) into two equivalently sized matrices and rectified the redundant couplings to make them suitable for non-negative dimensionality reduction.

Then, to produce a multiplexed view of the muscular interactions across trials, we iterated these MI computations over all unique combinations of [mx, my] and τ. The resulting MI estimates collectively form A, a symmetric adjacency matrix (i.e. ATA = I) that represents the functional connectivities between all muscle activations (fig.10). When repeated across all available task variables τ and participants, A is of dimension [No. of muscle pairs x No. of timepoints pairs x [No. of τ x No. of participants]. Thus, by applying network-theoretic statistical tools to A, we can identify functional modules carrying the same type of (redundant/synergistic) task information (fig.2(B)).

Estimating statistical significance of muscle couplings

To isolate statistically significant dependencies, we applied a modified percolation analysis to each A[41]. This method sparsifies functional connectivities in A with respect to its percolation threshold (Pc). Pc is a critical value that specifies the probability of a nodes’ occupation in A with respect to the networks size. In random networks, a ‘giant component’ comprised of long-range connections exists above Pc but disappears as Pc tends to zero [42], while it is thought that living systems optimise adaptability by fluctuating around Pc in a state of self-organised criticality [43]. Preliminary testing of this method showed it to be at least equivalent to permutation-testing each MI value in the network and thus, much more computationally efficient. Pc was therefore iteratively specified for each layer of A relative to equivalently sized random networks and utilised to remove insignificant network edges up to a stopping-point where this giant component begins to become affected (fig.3(C)). This procedure was carried out for each layer of A separately configured as muscle-wise couplings across temporal scales (i.e. a 3-D tensor of dimension [No. of muscle] x [No. of muscle] x [No. of timepoints pairings x No. of τ x No. of participants]) and vice versa as timepoint-wise couplings across spatial scales (fig.3(D)). The separate sparsification of eachindividual network layer in both alternative network configurations produced discrepancies in the output, as some connections were found to be significant in only one domain. To ameliorate this discrepancy, we employed a conservative heuristic where dependencies must be significant in both space and time to be included in the final input matrix for dimensionality reduction. Thus, the sparsified input matrices to dimensionality reduction were comprised of significant spatiotemporal task-redundant, -synergistic or - irrelevant muscle couplings.

Model-rank specification

To determine the optimal number of modules to extract, we implemented two alternative community detection algorithms generalised to multiplex networks [29,30,44]. Both forms seek to optimise a modularity criterion known as the Q-statistic that quantifies the proportion of within-community network edges compared to what would be expected from a network consisting of random connections [45]. More specifically, for a particular division of a single layer network (Equation 2.1), let, δ(gi, gj) = 1 if node i and j belong to the same group (g) and 0 otherwise and Aij be the number of edges between node i and j. The equivalent of Aij from a randomised network (Pij) is expected to be (Newman-Girvan null model) [45], where Ki and Kj are the node degrees and . The typical output of the Q-statistic is found within the range [0,1] with 1 indicating maximum modularity [45]. In its generalised multilayer form, the Q-statistic is given an additional term to consider couplings between layers l and r with intra- and inter-layer resolution parameters γ and ω (Equation 2.2). Here, μ is the total edge weight across the network and γ and ω were set to 1 in the current study for classical modularity [30], thus removing the need for any hyperparameter tuning. We chose to implement two complementary model-rank specification approaches to address limitations related to stochasticity and scalability present in the multilayer formulation and the inability to consider inter-layer dependencies present in the mono-layer formulation [18,46]. To apply these algorithms to our data, we grouped the set of A into multiplex networks configured with respect to spatial or temporal scales (fig.3(D)). We then applied these algorithms to both space-time network configurations for individual participant/tasks. This procedure generated a binary adjacency matrix from the resulting community partition vector in each case where 1 indicated the nodes belonged to the same community and 0 otherwise [44]. Following a consensus-based approach [31], we then grouped these binary adjacency matrices into a new multiplex network and re-applied the two alternative community detection algorithms to find an optimal spatial and temporal model-rank [31].

Extraction of low-dimensional representations

Following the specification of an optimal model-rank in the spatial and temporal domains, we used these values as input parameters into dimensionality reduction (fig.3(E)). To extract a low-dimensional representation of motor behaviour across muscle couplings, we applied a sample-based non-negative matrix tri-factorisation method (sNM3F) with additional orthogonality constraints to the matrices consisting of vectorised and concatenated [25]. More specifically, we decomposed the input three-mode tensor A of dimension [K = No. of unique muscle pairs (m) x (L = No. of time-sample pairs (t) x No. of participants + tasks)] into a set of spatial (V) and temporal (W) factors and the participant- and task-specific weighting coefficients (S) reflecting the amount of information carried by each combination of spatial-temporal factors for each participant and task parameter (Equation 3.1). In equation 3.2, this factorisation is also illustrated in vector sum notation for a single participant and task variable:

Examining the generalisability of extracted representations

To determine the generalisability of the extracted space-time muscle networks, we implemented a representational similarity analysis where we compared representations extracted from the full dataset 1-3 to equivalent networks extracted from a subset of the respective datasets. We computed the similarity between pairs of representations using Pearson’s correlation.

For dataset 1 and 2 (see below), we removed from the input data an individual participant or task at a time and compared the similarity of the decomposition outputs with those obtained from the full dataset. We repeated this for all participants and task variables and reported the average similarity as a measure of robustness of the decomposition.

For dataset 3 (see below), due to the greater number of participants and task parameters, we implemented a more stringent examination. More specifically, we firstly extracted representations from each task space individually and, using these representations as a reference, compared them against functionally similar outputs after removing randomly sized portions of randomly selected vectors in the input data (up to the no. of column vectors - 1). We repeated this procedure for 50 iterations, and computed summary statistics by converting the coefficients to Fisher’s Z values, computing the average and standard deviation, and then reverting these values back to correlation coefficients.

Subnetwork analysis

To illustrate the relative importance of individual muscles within each network, we determined the total communicability (C(i)) of individual nodes (i) in each network (A). C(i) is defined as the row-wise sum of all matrix exponentials (e) in the adjacency matrix (A) that consider the number of walks between each pair of nodes i and j (Equation 4.1) [37,69]: To emphasise salient functional connectivities in the spatial networks, we sparsified all dependencies with a below average network communicability and illustrated the output on the accompanying human body models [36,69]. To uncover salient subnetwork structures consisting of more closely functionally related muscle activations, we applied the monolayer community detection algorithm in equation 2.1 to the extracted spatial networks [29,32].

Data acquisition and processing

To illustrate our framework, we applied it to three datasets of EMG signals recorded during different motor tasks. In dataset 1 (fig.4(A)), 7 adult participants (age: 27 ± 2 years, height: 1.77 ± 0.03 m) performed table-top point-to-point reaching movements in both forward and backwards directions and at fast and slow speeds while the activity of nine muscles on the preferred right, reaching arm (finger extensors (FE), brachioradialis (BR), biceps brachii (BI), medial-triceps (TM), lateral-triceps (TL), anterior deltoid (AD), posterior deltoid (PD), pectoralis major (PE), latissimus dorsi (LD)) were captured for a total of 640 trials per participant [25]. To enable the quantification of shared information across muscles with respect to specific task attributes, we formulated four discrete task parameters of length equal to the number of trials executed. These discrete variables represented the trial-to-trial variation in reaching direction (fwd vs bwd), speed (high vs low), and reaching target [P1-P8] (fig.3(A)).

In dataset 2 (fig.4(B)), 3 participants performed whole-body point-to-point reaching movements in various directions and to varying heights while EMG from 30 muscles (tibialis anterior, soleus, peroneus, gastrocnemius, vastus lateralis, rectus femoris, biceps femoris, gluteus maximus, erector spinae, pectoralis major, trapezius, anterior deltoid, posterior deltoid, biceps and triceps brachii) across both hemi-bodies were captured [26]. Like dataset 1, we formulated task parameters each representing a specific task attribute across trials (∼2160 trials per participant). In this case, we formed eight discrete task parameters representing start- and end-point target, -bar, and -height and both up-down (vertical) and left-right (horizontal) reaching directions.

Dataset 3 consisted of multiple trials from 17 participants performing level-ground walking, stair- and ramp-ascents/descents with various sub-conditions (walking speed, clockwise/counter-clockwise direction, different stair/ramp inclines etc.) (fig.4(C)) [27]. These locomotion modes were performed while 11 EMG signals were captured from the right lower-limb (Gluteus medius (GlutM), right external oblique (Obl), semitendinosus (ST), gracilis (GR), biceps femoris (BF), rectus femoris (RF), vastus lateralis (VL), vastus medialis (VM), soleus (SO), tibialis anterior (TA), gastrocnemius medialis (GM)), XYZ coordinates were captured bilaterally from 32 kinematic markers and 4 IMUs and a force-plate captured accelerations and dynamic features among the lower-limbs also. More detailed breakdowns of the experimental design for each dataset can be found at their parent publications [25–27].

For all datasets, we processed the EMG signals offline using a standardised approach [10]: the EMGs for each sample were digitally full-wave rectified, low-pass filtered (Butterworth filter; 20 Hz cut-off; zero-phase distortion), normalised to 1000 time-samples and then the signals were integrated over 20 time-step intervals yielding a waveform of ∼50 time-steps. To match the time-series lengths, we resampled the kinematic, dynamic and IMU recordings of dataset 3 using cubic-spline interpolation to match the EMG signals.

Supporting information

eLife Assessment

The work by O'Reilly and Delis is important to extend the synergy ideas using methods from signal processing and information theory to cluster muscles and task parameters, thereby advancing our understanding of the modular architecture of motor control. The method is innovative, and the findings are compelling from theoretical and practical perspectives. The work will be of broad interest to motor control and neural engineering researchers.

Reviewer #1 (Public Review):

The proposed study provides an innovative framework for the identification of muscle synergies taking into account their task relevance. State-of-the-art techniques for extracting muscle interactions use unsupervised machine-learning algorithms applied to the envelopes of the electromyographic signals without taking into account the information related to the task being performed. In this work, the authors suggest including the task parameters in extracting muscle synergies using a network information framework previously proposed. This allows the identification of muscle interactions that are relevant, irrelevant, or redundant to the parameters of the task executed.

The proposed framework is a powerful tool to understand and identify muscle interactions for specific task parameters and it may be used to improve man-machine interfaces for the control of prostheses and robotic exoskeletons.

With respect to the network information framework recently published, this work added an important part to estimate the relevance of specific muscle interactions to the parameters of the task executed. However, the authors should better explain what is the added value of this contribution with respect to the previous one, also in terms of computational methods.

It is not clear how the well-known phenomenon of cross-talk during the recording of electromyographic muscle activity may affect the performance of the proposed technique and how it may bias the overall outcomes of the framework.

Reviewer #2 (Public Review):

This paper is an attempt to extend or augment muscle synergy and motor primitive ideas with task measures. The authors idea is to use information metrics (mutual information, co-information) in 'synergy' constraint creation that includes task information directly. By using task related information and muscle information sources and then sparsification, the methods construct task relevant network communities among muscles, together with task redundant communities, and task irrelevant communities. This process of creating network communities may then constrain and help to guide subsequent synergy identification using the authors published sNM3F algorithm to detect spatial and temporal synergies.

The revised paper is much clearer and examples are helpful in various ways. However, figure 2 as presented does not convincingly show why task muscle mutual information helps in separating synergies, though it is helpful in defining the various network communities used in the toy example.

The impact of the information theoretic constraints developed as network communities on subsequent synergy separation are posited to be benign and to improve over other methods (e.g., NNMF). However, not fully addressed are the possible impacts of the methods on compositionality links with physiological bases, and the possibility remains of the methods sometimes instead leading to modules that represent more descriptive ML frameworks that may not support physiological work easily. Accordingly, there is a caveat. This is recognized and acknowledged by the authors in their rebuttal of the prior review. It will remain for other work to explore this issue, likely through testing on detailed high degree of freedom artificial neuromechanical models and tasks. This possible issue with the strategy here likely needs to be fully acknowledged in the paper.

The approach of the methods seeks to identify task relevant coordinative couplings. This is a meta problem for more classical synergy analyses. Classical analyses seek compositional elements stable across tasks. These elements may then be explored in causal experiments and generative simulations of coupling and control strategies. However, task-based understanding of synergy roles and functional uses is significant and is clearly likely to be aided by methods in this study.

Information based separation has been used in muscle synergy analyses using infomax ICA, which is information based at core. Though linear mixing of sources is assumed in ICA, minimized mutual information among source (synergy) drives is the basis of the separation and detects low variance synergy contributions (e.g., see Yang, Logan, Giszter, 2019). In the work in this paper, instead, mutual information approaches are used to cluster muscles and task features into network communities preceding the SNM3F algorithm use for separation, rather than using minimized information in separation. This contrast of an accretive or agglomerative mutual information strategy here used to cluster into networks, versus a minimizing mutual information source separation used in infomax ICA epitomizes a key difference in approach here.

Physiological causal testing of synergy ideas is neglected in the literature reviews in the paper. Although these are only in animal work (Hart and Giszter, 2010; Takei and Seki, 2017), the clear connection of muscle synergy analysis choices to physiology is important, and eventually these issues need to be better managed and understood in relation to the new methods proposed here, even if not in this paper.

Analyses of synergies using the methods the paper has proposed will likely be very much dependent on the number and quality of task variables included and how these are managed, and the impacts of these on the ensuing sparsification and network communities used prior to SNM3F. The authors acknowledge this in their response. This caveat should likely be made very explicit in the paper.

It would be useful in the future to explore the approach described with a range of simulated data to better understand the caveats, and optimizations for best practices in this approach.

Reviewer #3 (Public Review):

In this study, the authors developed and tested a novel framework for extracting muscle synergies. The approach aims at removing some limitations and constraints typical of previous approaches used in the field. In particular, the authors propose a mathematical formulation that removes constraints of linearity and couples the synergies to their motor outcome, supporting the concept of functional synergies and distinguishing the task-related performance related to each synergy. While some concepts behind this work were already introduced in recent work in the field, the methodology provided here encapsulates all these features in an original formulation providing a step forward with respect to the currently available algorithms. The authors also successfully demonstrated the applicability of their method to previously available datasets of multi-joint movements.

Preliminary results positively support the scientific soundness of the presented approach and its potential. The added values of the method should be documented more in future work to understand how the presented formulation relates to previous approaches and what novel insights can be achieved in practical scenarios and confirm/exploit the potential of the theoretical findings.

In their revision, the authors have implemented major revisions and improved their paper. The work was already of good quality and now it has improved further. The authors were able to successfully:

- improve the clarity of the writing (e.g.: better explaining the rationale and the aims of the paper);

- extend the clarification of some of the key novel concepts introduced in their work, like the redundant synergies;

- show a scenario in which their approach might be useful for increasing the understanding of motor control in patients with respect to traditional algorithms such as NMF. In particular, their example illustrates why considering the task space is a fundamental step forward when extracting muscle synergies, improving the practical and physiological interpretation of the results.

Author Response

The following is the authors’ response to the previous reviews.

Public Reviews:

Reviewer #1 (Public Review):

The proposed study provides an innovative framework for the identification of muscle synergies taking into account their task relevance. State-of-the-art techniques for extracting muscle interactions use unsupervised machine-learning algorithms applied to the envelopes of the electromyographic signals without taking into account the information related to the task being performed. In this work, the authors suggest including the task parameters in extracting muscle synergies using a network information framework previously proposed. This allows the identification of muscle interactions that are relevant, irrelevant, or redundant to the parameters of the task executed.

The proposed framework is a powerful tool to understand and identify muscle interactions for specific task parameters and it may be used to improve man-machine interfaces for the control of prostheses and robotic exoskeletons.

With respect to the network information framework recently published, this work added an important part to estimate the relevance of specific muscle interactions to the parameters of the task executed. However, the authors should better explain what is the added value of this contribution with respect to the previous one, also in terms of computational methods.

It is not clear how the well-known phenomenon of cross-talk during the recording of electromyographic muscle activity may affect the performance of the proposed technique and how it may bias the overall outcomes of the framework.

We thank reviewer 1 for their useful commentary on this manuscript.

Reviewer #2 (Public Review):

This paper is an attempt to extend or augment muscle synergy and motor primitive ideas with task measures. The authors idea is to use information metrics (mutual information, co-information) in 'synergy' constraint creation that includes task information directly. By using task related information and muscle information sources and then sparsification, the methods construct task relevant network communities among muscles, together with task redundant communities, and task irrelevant communities. This process of creating network communities may then constrain and help to guide subsequent synergy identification using the authors published sNM3F algorithm to detect spatial and temporal synergies.

The revised paper is much clearer and examples are helpful in various ways. However, figure 2 as presented does not convincingly show why task muscle mutual information helps in separating synergies, though it is helpful in defining the various network communities used in the toy example.

The impact of the information theoretic constraints developed as network communities on subsequent synergy separation are posited to be benign and to improve over other methods (e.g., NNMF). However, not fully addressed are the possible impacts of the methods on compositionality links with physiological bases, and the possibility remains of the methods sometimes instead leading to modules that represent more descriptive ML frameworks that may not support physiological work easily. Accordingly, there is a caveat. This is recognized and acknowledged by the authors in their rebuttal of the prior review. It will remain for other work to explore this issue, likely through testing on detailed high degree of freedom artificial neuromechanical models and tasks. This possible issue with the strategy here likely needs to be fully acknowledged in the paper.

The approach of the methods seeks to identify task relevant coordinative couplings. This is a meta problem for more classical synergy analyses. Classical analyses seek compositional elements stable across tasks. These elements may then be explored in causal experiments and generative simulations of coupling and control strategies. However, task-based understanding of synergy roles and functional uses is significant and is clearly likely to be aided by methods in this study.

Information based separation has been used in muscle synergy analyses using infomax ICA, which is information based at core. Though linear mixing of sources is assumed in ICA, minimized mutual information among source (synergy) drives is the basis of the separation and detects low variance synergy contributions (e.g., see Yang, Logan, Giszter, 2019). In the work in this paper, instead, mutual information approaches are used to cluster muscles and task features into network communities preceding the SNM3F algorithm use for separation, rather than using minimized information in separation. This contrast of an accretive or agglomerative mutual information strategy here used to cluster into networks, versus a minimizing mutual information source separation used in infomax ICA epitomizes a key difference in approach here.

Physiological causal testing of synergy ideas is neglected in the literature reviews in the paper. Although these are only in animal work (Hart and Giszter, 2010; Takei and Seki, 2017), the clear connection of muscle synergy analysis choices to physiology is important, and eventually these issues need to be better managed and understood in relation to the new methods proposed here, even if not in this paper.

Analyses of synergies using the methods the paper has proposed will likely be very much dependent on the number and quality of task variables included and how these are managed, and the impacts of these on the ensuing sparsification and network communities used prior to SNM3F. The authors acknowledge this in their response. This caveat should likely be made very explicit in the paper.

It would be useful in the future to explore the approach described with a range of simulated data to better understand the caveats, and optimizations for best practices in this approach.

A key component of the reviewers’ arguments here is their reductionist view of muscle synergies vs the emergentist view presented in our work here. In the reductionist lens, muscle groupings are the units (‘building blocks’) of coordinated movement and thus the space of intermuscular interactions is of particular interest for understanding movement construction. On the other hand, the emergentist view suggests that muscle groupings emerge from interactions between constituent parts (as quantified here using information theory, synergistic information is the information found when both activities are observed together). This is in line with recent work in the field showing modular control at the intramuscular level, exemplifying a scale-free phenomena. Nonetheless, we consider these approaches to muscle synergy research as complementary and beneficial for the field overall going forward.

Reviewer #3 (Public Review):

In this study, the authors developed and tested a novel framework for extracting muscle synergies. The approach aims at removing some limitations and constraints typical of previous approaches used in the field. In particular, the authors propose a mathematical formulation that removes constraints of linearity and couples the synergies to their motor outcome, supporting the concept of functional synergies and distinguishing the task-related performance related to each synergy. While some concepts behind this work were already introduced in recent work in the field, the methodology provided here encapsulates all these features in an original formulation providing a step forward with respect to the currently available algorithms. The authors also successfully demonstrated the applicability of their method to previously available datasets of multi-joint movements.

Preliminary results positively support the scientific soundness of the presented approach and its potential. The added values of the method should be documented more in future work to understand how the presented formulation relates to previous approaches and what novel insights can be achieved in practical scenarios and confirm/exploit the potential of the theoretical findings.

In their revision, the authors have implemented major revisions and improved their paper. The work was already of good quality and now it has improved further. The authors were able to successfully:

improve the clarity of the writing (e.g.: better explaining the rationale and the aims of the paper);

extend the clarification of some of the key novel concepts introduced in their work, like the redundant synergies;

show a scenario in which their approach might be useful for increasing the understanding of motor control in patients with respect to traditional algorithms such as NMF. In particular, their example illustrates why considering the task space is a fundamental step forward when extracting muscle synergies, improving the practical and physiological interpretation of the results.

We thank reviewer 3 for their constructive commentary on this manuscript.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

Figure 3 should report the distances between reaching points in panel A and the actual length distances of the walking paths in panel C.

The caption of fig.3 concerning the experimental setup of the datasets analysed has been updated with the following for dataset 1: “(A) Dataset 1 consisted of participants executing table-top point-to-point reaching movements (40cm distance from starting point P0) across four targets in forward (P1-P4) and backwards (P5-P8) directions at both fast and slow speeds (40 repetitions per task) [25]. The muscles recorded included the finger extensors (FE), brachioradialis (BR), biceps brachii (BI), medial-triceps (TM), lateral-triceps (TL), anterior deltoid (AD), posterior deltoid (PD), pectoralis major (PE), latissimus dorsi (LD) of the right, reaching arm.”. For dataset 3, to the best of the authors knowledge, this information was not given in the original paper.

Figure 4, what is the unit of the data shown?

The unit of bits is now mentioned in the toy example figure caption and in the caption of fig.5

Figure 4, the characteristics of the interactions are not fully clear, and the graphical representation should be improved.

We have made steps to improve the clarity of the figures presented.

For dataset 3, τ was the movement kinematics, but it is not specified how the task parameters were formulated. Did the authors use the data from all 32 kinematic markers, 4 IMUs, and force plates? If yes, it should be specified why all these signals were used. For sure, there will be signals included that are not relevant to the specific task. Did the authors select specific signals based on their relevance to the task (e.g., ankle kinematics)?

We have now clarified this in the text as follows: “For datasets 1 and 2, we determine the MI between vectors with respect to several discrete task parameters representing specific task attributes (e.g. reaching direction, speed etc.), while for dataset 3 we determined the task-relevant and -irrelevant muscles couplings in an unassuming way by quantifying them with respect to all available kinematic, dynamic and inertial motion unit (IMU) features.”

How did the authors endure that crosstalk did not affect their analysis, particularly between, e.g., finger extensors and brachioradialis and posterior deltoid and anterior deltoid (dataset 1)?

We have addressed this point in the previous round of reviews and made an explicit statement regarding cross-talk in the discussion section: “Although distinguishing task-irrelevant muscle couplings may capture artifacts such as EMG crosstalk, our results convey several physiological objectives of muscles including gross motor functions [66], the maintenance of internal joint mechanics and reciprocal inhibition of contralateral limbs [19,51].”

It would be informative to add some examples of not trivial/obvious task-related synergistic muscle combinations that have been extracted in the three datasets. Most of the examples reported in the manuscript are well-known biomechanically and quite intuitive, so they do not improve our understanding of synergistic muscle control in humans.

Our framework improves our understanding of synergistic motor control by enabling the formal quantification of synergistic muscle interactions, a capability not present among current approaches. Regarding the implications of this advance in terms of concrete examples, we have further clarified our examples presented in the results section, for example:

“Across datasets, many the muscle networks could be characterised by the transmission of complementary task information between functionally specialised muscle groups, many of which identified among the task-redundant representations (Fig.9-10 and Supp. Fig.2). The most obvious example of this is the S3 synergist muscle network of dataset 2 (Fig.11), which captures the complementary interaction between task-redundant submodules identified previously (S3 (Fig.9)).”

The description shows how our framework can extract the cross-module interactions that align with the higher-level objectives of the system, here the synergistic connectivity between the upper and lower body modules. Current approaches can only capture redundant and task-irrelevant interactions. Thus our framework provides additional insight into movement control.

The number of participations in dataset 2 is very limited and should be increased. We appreciate the reviewer's comment and would like to point out that for dataset 2 our aim was to increase the number of muscles (30), tasks (72) and trials for each task (30) which produced a very large dataset for each participant. This came at the expense of low number of participants, however all our statistical analyses here can be performed at the single-participant level. Furthermore, dataset 3 includes 25 participants and it enables us to demonstrate the reliability of the findings across participants.

Reviewer #2 (Recommendations For The Authors):

I believe it is important in the future to explore the approach proposed with a range of simulation data and neuromechanical models, to explore the issues I have raised and that you have acknowledged, though I agree it is likely out of scope for the paper here.

We agree with the reviewer that this would be valuable future work and indeed plan to do this in our future research.

The Github code for this paper should likely include the various data sets used in the paper and figures, appropriately anonymized, in order to allow the data to be explored and analyses replicated and package demonstrated to be exercised fully by a new user.

We thank the reviewer for this suggestion. Dataset3 is already available online at https://doi.org/10.1016/j.jbiomech.2021.110320. We will also make the other 2 datasets publicly available on our lab website very soon. Until then, as stated in the manuscript, we will make them available to anyone upon reasonable request.

Reviewer #3 (Recommendations For The Authors):

I have the following open points to suggest to the authors:

First, I recommend improving the quality of the figures: in the pdf version I downloaded, some writings are impossible to read.

We fully agree with the reviewer and note that in the pdf version of the paper, the figures are a lot worse than in the submitted word document submitted. Nevertheless, we will make further improvements on the figures as requested.

Even though the manuscript has improved, I still feel that some points were not addressed or were only partially addressed. In particular:

The proposed comparison with NMF helps understanding why incorporating the task space is useful (and I fully agree with the authors about this point as the main reason to propose their contribution). However, the comparison does not help the reader to understand whether the synergies incorporating the task space are biased by the introduction of the task variables.

This question can be also reformulated as: are muscle synergies modified when task space variables are incorporated? Is the "weight" on task coefficients affecting the composition of muscle synergies? If so, the added interpretational power is achieved at the cost of losing the information regarding the neural substrate of synergies? I understand this point is not immediate to show, but it would increase the quality of the work.

Reference to previous approaches that aimed at including task variables into synergy extraction are still missing in the paper. Even though it is not required to provide quantitative comparisons with other available approaches, there are at most 2-3 available algorithms in the literature (kinematics-EMG; force-EMG), that should not be neglected in this work. What did previous approaches achieve? What was improved with this approach? What was not improved?

Previous attempts of extracting synergies with non-linear approaches could also be described more.

In the latest version of the manuscript, we have referenced both the mixed NMF and autoencoders based algorithms. In both the introduction and discussion section of the manuscript, we also specify that our framework quantifies and decomposes muscle interactions in a novel way that cannot be done by other current approaches. In the results section we use examples from 3 different datasets to make this point clear, providing intuition on the use cases of our framework.

Author Response

The following is the authors’ response to the previous reviews.

Public Reviews:

Reviewer #1 (Public Review):

The proposed study provides an innovative framework for the identification of muscle synergies taking into account their task relevance. State-of-the-art techniques for extracting muscle interactions use unsupervised machine-learning algorithms applied to the envelopes of the electromyographic signals without taking into account the information related to the task being performed. In this work, the authors suggest including the task parameters in extracting muscle synergies using a network information framework previously proposed. This allows the identification of muscle interactions that are relevant, irrelevant, or redundant to the parameters of the task executed.

The proposed framework is a powerful tool to understand and identify muscle interactions for specific task parameters and it may be used to improve man-machine interfaces for the control of prostheses and robotic exoskeletons.

With respect to the network information framework recently published, this work added an important part to estimate the relevance of specific muscle interactions to the parameters of the task executed. However, the authors should better explain what is the added value of this contribution with respect to the previous one, also in terms of computational methods.

It is not clear how the well-known phenomenon of cross-talk during the recording of electromyographic muscle activity may affect the performance of the proposed technique and how it may bias the overall outcomes of the framework.

We thank reviewer 1 for their useful commentary on this manuscript.

Reviewer #2 (Public Review):

This paper is an attempt to extend or augment muscle synergy and motor primitive ideas with task measures. The authors idea is to use information metrics (mutual information, co-information) in 'synergy' constraint creation that includes task information directly. By using task related information and muscle information sources and then sparsification, the methods construct task relevant network communities among muscles, together with task redundant communities, and task irrelevant communities. This process of creating network communities may then constrain and help to guide subsequent synergy identification using the authors published sNM3F algorithm to detect spatial and temporal synergies.

The revised paper is much clearer and examples are helpful in various ways. However, figure 2 as presented does not convincingly show why task muscle mutual information helps in separating synergies, though it is helpful in defining the various network communities used in the toy example.

The impact of the information theoretic constraints developed as network communities on subsequent synergy separation are posited to be benign and to improve over other methods (e.g., NNMF). However, not fully addressed are the possible impacts of the methods on compositionality links with physiological bases, and the possibility remains of the methods sometimes instead leading to modules that represent more descriptive ML frameworks that may not support physiological work easily. Accordingly, there is a caveat. This is recognized and acknowledged by the authors in their rebuttal of the prior review. It will remain for other work to explore this issue, likely through testing on detailed high degree of freedom artificial neuromechanical models and tasks. This possible issue with the strategy here likely needs to be fully acknowledged in the paper.

The approach of the methods seeks to identify task relevant coordinative couplings. This is a meta problem for more classical synergy analyses. Classical analyses seek compositional elements stable across tasks. These elements may then be explored in causal experiments and generative simulations of coupling and control strategies. However, task-based understanding of synergy roles and functional uses is significant and is clearly likely to be aided by methods in this study.

Information based separation has been used in muscle synergy analyses using infomax ICA, which is information based at core. Though linear mixing of sources is assumed in ICA, minimized mutual information among source (synergy) drives is the basis of the separation and detects low variance synergy contributions (e.g., see Yang, Logan, Giszter, 2019). In the work in this paper, instead, mutual information approaches are used to cluster muscles and task features into network communities preceding the SNM3F algorithm use for separation, rather than using minimized information in separation. This contrast of an accretive or agglomerative mutual information strategy here used to cluster into networks, versus a minimizing mutual information source separation used in infomax ICA epitomizes a key difference in approach here.

Physiological causal testing of synergy ideas is neglected in the literature reviews in the paper. Although these are only in animal work (Hart and Giszter, 2010; Takei and Seki, 2017), the clear connection of muscle synergy analysis choices to physiology is important, and eventually these issues need to be better managed and understood in relation to the new methods proposed here, even if not in this paper.

Analyses of synergies using the methods the paper has proposed will likely be very much dependent on the number and quality of task variables included and how these are managed, and the impacts of these on the ensuing sparsification and network communities used prior to SNM3F. The authors acknowledge this in their response. This caveat should likely be made very explicit in the paper.

It would be useful in the future to explore the approach described with a range of simulated data to better understand the caveats, and optimizations for best practices in this approach.

A key component of the reviewers’ arguments here is their reductionist view of muscle synergies vs the emergentist view presented in our work here. In the reductionist lens, muscle groupings are the units (‘building blocks’) of coordinated movement and thus the space of intermuscular interactions is of particular interest for understanding movement construction. On the other hand, the emergentist view suggests that muscle groupings emerge from interactions between constituent parts (as quantified here using information theory, synergistic information is the information found when both activities are observed together). This is in line with recent work in the field showing modular control at the intramuscular level, exemplifying a scale-free phenomena. Nonetheless, we consider these approaches to muscle synergy research as complementary and beneficial for the field overall going forward.

Reviewer #3 (Public Review):

In this study, the authors developed and tested a novel framework for extracting muscle synergies. The approach aims at removing some limitations and constraints typical of previous approaches used in the field. In particular, the authors propose a mathematical formulation that removes constraints of linearity and couples the synergies to their motor outcome, supporting the concept of functional synergies and distinguishing the task-related performance related to each synergy. While some concepts behind this work were already introduced in recent work in the field, the methodology provided here encapsulates all these features in an original formulation providing a step forward with respect to the currently available algorithms. The authors also successfully demonstrated the applicability of their method to previously available datasets of multi-joint movements.

Preliminary results positively support the scientific soundness of the presented approach and its potential. The added values of the method should be documented more in future work to understand how the presented formulation relates to previous approaches and what novel insights can be achieved in practical scenarios and confirm/exploit the potential of the theoretical findings.

In their revision, the authors have implemented major revisions and improved their paper. The work was already of good quality and now it has improved further. The authors were able to successfully:

improve the clarity of the writing (e.g.: better explaining the rationale and the aims of the paper);

extend the clarification of some of the key novel concepts introduced in their work, like the redundant synergies;

show a scenario in which their approach might be useful for increasing the understanding of motor control in patients with respect to traditional algorithms such as NMF. In particular, their example illustrates why considering the task space is a fundamental step forward when extracting muscle synergies, improving the practical and physiological interpretation of the results.

We thank reviewer 3 for their constructive commentary on this manuscript.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

Figure 3 should report the distances between reaching points in panel A and the actual length distances of the walking paths in panel C.

The caption of fig.3 concerning the experimental setup of the datasets analysed has been updated with the following for dataset 1: “(A) Dataset 1 consisted of participants executing table-top point-to-point reaching movements (40cm distance from starting point P0) across four targets in forward (P1-P4) and backwards (P5-P8) directions at both fast and slow speeds (40 repetitions per task) [25]. The muscles recorded included the finger extensors (FE), brachioradialis (BR), biceps brachii (BI), medial-triceps (TM), lateral-triceps (TL), anterior deltoid (AD), posterior deltoid (PD), pectoralis major (PE), latissimus dorsi (LD) of the right, reaching arm.”. For dataset 3, to the best of the authors knowledge, this information was not given in the original paper.

Figure 4, what is the unit of the data shown?

The unit of bits is now mentioned in the toy example figure caption and in the caption of fig.5

Figure 4, the characteristics of the interactions are not fully clear, and the graphical representation should be improved.

We have made steps to improve the clarity of the figures presented.

For dataset 3, τ was the movement kinematics, but it is not specified how the task parameters were formulated. Did the authors use the data from all 32 kinematic markers, 4 IMUs, and force plates? If yes, it should be specified why all these signals were used. For sure, there will be signals included that are not relevant to the specific task. Did the authors select specific signals based on their relevance to the task (e.g., ankle kinematics)?

We have now clarified this in the text as follows: “For datasets 1 and 2, we determine the MI between vectors with respect to several discrete task parameters representing specific task attributes (e.g. reaching direction, speed etc.), while for dataset 3 we determined the task-relevant and -irrelevant muscles couplings in an unassuming way by quantifying them with respect to all available kinematic, dynamic and inertial motion unit (IMU) features.”

How did the authors endure that crosstalk did not affect their analysis, particularly between, e.g., finger extensors and brachioradialis and posterior deltoid and anterior deltoid (dataset 1)?

We have addressed this point in the previous round of reviews and made an explicit statement regarding cross-talk in the discussion section: “Although distinguishing task-irrelevant muscle couplings may capture artifacts such as EMG crosstalk, our results convey several physiological objectives of muscles including gross motor functions [66], the maintenance of internal joint mechanics and reciprocal inhibition of contralateral limbs [19,51].”

It would be informative to add some examples of not trivial/obvious task-related synergistic muscle combinations that have been extracted in the three datasets. Most of the examples reported in the manuscript are well-known biomechanically and quite intuitive, so they do not improve our understanding of synergistic muscle control in humans.

Our framework improves our understanding of synergistic motor control by enabling the formal quantification of synergistic muscle interactions, a capability not present among current approaches. Regarding the implications of this advance in terms of concrete examples, we have further clarified our examples presented in the results section, for example:

“Across datasets, many the muscle networks could be characterised by the transmission of complementary task information between functionally specialised muscle groups, many of which identified among the task-redundant representations (Fig.9-10 and Supp. Fig.2). The most obvious example of this is the S3 synergist muscle network of dataset 2 (Fig.11), which captures the complementary interaction between task-redundant submodules identified previously (S3 (Fig.9)).”

The description shows how our framework can extract the cross-module interactions that align with the higher-level objectives of the system, here the synergistic connectivity between the upper and lower body modules. Current approaches can only capture redundant and task-irrelevant interactions. Thus our framework provides additional insight into movement control.

The number of participations in dataset 2 is very limited and should be increased. We appreciate the reviewer's comment and would like to point out that for dataset 2 our aim was to increase the number of muscles (30), tasks (72) and trials for each task (30) which produced a very large dataset for each participant. This came at the expense of low number of participants, however all our statistical analyses here can be performed at the single-participant level. Furthermore, dataset 3 includes 25 participants and it enables us to demonstrate the reliability of the findings across participants.

Reviewer #2 (Recommendations For The Authors):

I believe it is important in the future to explore the approach proposed with a range of simulation data and neuromechanical models, to explore the issues I have raised and that you have acknowledged, though I agree it is likely out of scope for the paper here.

We agree with the reviewer that this would be valuable future work and indeed plan to do this in our future research.

The Github code for this paper should likely include the various data sets used in the paper and figures, appropriately anonymized, in order to allow the data to be explored and analyses replicated and package demonstrated to be exercised fully by a new user.

We thank the reviewer for this suggestion. Dataset3 is already available online at https://doi.org/10.1016/j.jbiomech.2021.110320. We will also make the other 2 datasets publicly available on our lab website very soon. Until then, as stated in the manuscript, we will make them available to anyone upon reasonable request.

Reviewer #3 (Recommendations For The Authors):

I have the following open points to suggest to the authors:

First, I recommend improving the quality of the figures: in the pdf version I downloaded, some writings are impossible to read.

We fully agree with the reviewer and note that in the pdf version of the paper, the figures are a lot worse than in the submitted word document submitted. Nevertheless, we will make further improvements on the figures as requested.

Even though the manuscript has improved, I still feel that some points were not addressed or were only partially addressed. In particular:

The proposed comparison with NMF helps understanding why incorporating the task space is useful (and I fully agree with the authors about this point as the main reason to propose their contribution). However, the comparison does not help the reader to understand whether the synergies incorporating the task space are biased by the introduction of the task variables.

This question can be also reformulated as: are muscle synergies modified when task space variables are incorporated? Is the "weight" on task coefficients affecting the composition of muscle synergies? If so, the added interpretational power is achieved at the cost of losing the information regarding the neural substrate of synergies? I understand this point is not immediate to show, but it would increase the quality of the work.

Reference to previous approaches that aimed at including task variables into synergy extraction are still missing in the paper. Even though it is not required to provide quantitative comparisons with other available approaches, there are at most 2-3 available algorithms in the literature (kinematics-EMG; force-EMG), that should not be neglected in this work. What did previous approaches achieve? What was improved with this approach? What was not improved?

Previous attempts of extracting synergies with non-linear approaches could also be described more.

In the latest version of the manuscript, we have referenced both the mixed NMF and autoencoders based algorithms. In both the introduction and discussion section of the manuscript, we also specify that our framework quantifies and decomposes muscle interactions in a novel way that cannot be done by other current approaches. In the results section we use examples from 3 different datasets to make this point clear, providing intuition on the use cases of our framework.

Version 3

Abstract

Abstract

The muscle synergy is a guiding concept in motor control research that relies on the general notion of muscles ‘working together’ towards task performance. However, although the synergy concept has provided valuable insights into motor coordination, muscle interactions have not been fully characterised with respect to task performance. Here, we address this research gap by proposing a novel perspective to the muscle synergy that assigns specific functional roles to muscle couplings by characterising their task-relevance. Our novel perspective provides nuance to the muscle synergy concept, demonstrating how muscular interactions can ‘work together’ in different ways: a) irrespective of the task at hand but also b) redundantly or c) complementarily towards common task-goals. To establish this perspective, we leverage information- and network-theory and dimensionality reduction methods to include discrete and continuous task parameters directly during muscle synergy extraction. Specifically, we introduce co-information as a measure of the task relevance of muscle interactions and use it to categorise such interactions as task-irrelevant (present across tasks), redundant (shared task information) or synergistic (different task information). To demonstrate these types of interactions in real data, we firstly apply the framework in a simple way, revealing its added functional and physiological relevance with respect to current approaches. We then apply the framework to large-scale datasets and extract generalizable and scale-invariant representations consisting of subnetworks of synchronised muscle couplings and distinct temporal patterns. The representations effectively capture the functional interplay between task end-goals and biomechanical affordances and the concurrent processing of functionally similar and complementary task information. The proposed framework unifies the capabilities of current approaches in capturing distinct motor features while providing novel insights and research opportunities through a nuanced perspective to the muscle synergy.

Introduction

Human movement is a highly complex behaviour, with a broad spectrum of multiplexed spatiotemporal dynamics typically exhibited for basic activities-of-daily-living [1,2]. How the central nervous system controls movement in the face of this inherent complexity to ensure efficient and reliable navigation of the environment and task performance is a nontrivial question currently under investigation in the motor control field [3,4]. The muscle synergy hypothesis is a long-withstanding proposition on the underlying neural constraints producing coordinated movement, stating that this complexity is offset by the allocation of computational resources to the spinal-level in the form of motor primitives [3–6]. These motor primitives modularly activate functional groups of muscles that are flexibly combined for the efficient construction of a movement. The conceptual underpinning of the ‘muscle synergy’ entails the idea of combinations of muscles ‘working together’ for the purpose of effective goal-orientated behaviour [7]. This emergent cohesion involves the following qualifying attributes: a repeatable muscle activation pattern common across trials and participants, a reciprocal relationship among functional muscle groups such that changes occur in one group to compensate for changes in another, and task dependence (i.e. the pattern of interdependencies among muscles must map onto task performance) [7,8]. A common approach to analyse the neural constraints underlying these motor patterns is to apply unsupervised machine-learning algorithms to electromyographic (EMG) data [9,10], with the aim of extracting a latent, low-dimensional representation.

In [11], we considered, key limitations among current approaches to muscle synergy analysis in extracting functionally relevant and interpretable patterns of muscle activity [12]. We proposed a combinatorial approach based on information- and network-theory and dimensionality reduction (the network-information framework (NIF)) that significantly improved the generalisability of the extraction process by, among others, removing restrictive model assumptions (e.g. linearity, same mixing coefficients) and the reliance on variance-accounted-for (VAF) metrics [12]. By determining the pairwise mutual information between muscles, this innovation paved the way for the appropriate mapping of muscular interactions to the task space. To elaborate on the significance of this development, the extraction of motor patterns in isolation of the task space often comes at the expense of functional and physiological relevance [12,13]. Furthermore, effective methods for mapping large-scale physiological dynamics to behaviour is a current gap across the neurosciences [14]. Thus, here we build on this work by, for the first time, directly including task space parameters during muscle synergy extraction. This enables us, in a novel way, to dissect the concept of the muscle synergy and therefore quantify interactions between muscle activations with shared or complementary functional roles.

Further to the above, in its currently defined state, the muscle synergy concept describes the role of common neural drives to functional muscle groupings working redundantly towards a common task-goal [3–5,7,15]. However, recent influential works have highlighted several other important mechanisms involved in this low-dimensional control strategy that are not well recognised by the muscle synergy concept [16–24]. Such insights include the partitioning of motor variability by the nervous system into task-relevant and -irrelevant spaces and the cooperation between functionally distinct muscle groupings in the form of cross-module functional connectivities. These observations highlight the need for a refinement of the muscle synergy concept to comprehensively describe diverse muscle interactions during movement, including their partitioning into task-relevant and –irrelevant spaces and the characterisation of their functional roles.

We thus motivate the development of a more nuanced perspective to the muscle synergy concept and the general notion of ‘working together’ that comprehensively describes the muscle interactions underlying motor behaviour. To do so, we propose an information-theoretic approach (based on the NIF pipeline) that characterises the contributions of muscle couplings to task performance. In other words, we frame the notion of ‘working together’ in more specific terms of shared information between pairs of spatiotemporal muscle activations ([mx, my] (Fig.1.3(a))) (red and green sets respectively) and a corresponding task parameter (τ) (blue set) (see Venn diagrams in Fig.1.3(b) below). Among current approaches to muscle synergy analysis, the shared information between muscles (yellow and white area in Fig.1.1(a)) is quantified, using dimensionality reduction, as common patterns of variability. These common patterns are essentially task-agnostic and may contain patterns of variability a) present in specific tasks (i.e. task-relevant (white shaded area in Fig.1.1(a)) as well as b) shared across tasks (i.e. task–irrelevant (yellow shaded area in Fig.1.1(a)). Our proposed approach dissects patterns of muscle variability in space, time and across trials in terms of their task-relevance and functional similarity in a generalizable manner using mutual information (MI) (Fig.1.3(a-c)). This enables us to decompose muscle activations into muscle pair – task parameter couplings and characterise their combined functional roles. We can then extract low-dimensional representations of these muscle couplings, i.e. muscle networks with specific spatial and temporal signatures, across participants and tasks (Fig.1.3(c)).

A general outline of the proposed approach. (1.1(a-b)) We propose a novel approach to mapping muscle couplings to the task space. Among current muscle synergy analysis approaches, muscle couplings are quantified in isolation of the task solely using dimensionality reduction. Using our approach, the functional characteristics of muscle interactions can be quantified in terms of the similarity of their encoded task information. We do so by determining the coupling between [mx, my] and a corresponding task parameter (τ) using mutual information (MI). From this perspective, task-redundant muscle couplings (pink shaded area in pink-orange intersection) represent muscles cooperating towards similar task goals, while task-synergistic muscle couplings (orange shaded area in pink-orange intersection) encapsulate the task information provided by a muscle pairing acting towards complementary task goals. Muscle couplings present across tasks (i.e., task-irrelevant) are quantified by conditioning the MI between [mx, my] pairs with respect to τ (yellow intersection). (1.2) A description of redundant and synergistic interactions. (a) Net redundant interactions are defined by a greater amount of information generated by the sum of individual observation of mx and my ([mx + my]) than their simultaneous observation ([mx, my]). (b) In a net synergistic interaction, [mx, my] provides more information than [mx + my]. (1.3(a-c)) An overview of the approach. Spatiotemporal muscle activation samples are extracted across trials from large-scale EMG datasets and concatenated into vectors, forming [mx, my] pairs.The derived muscle couplings are then run through the NIF pipeline [11], producing low-dimensional, multiplexed space-time muscle networks.

Crucially, using this novel framework, we can separately quantify the task-irrelevant (i.e. muscle interactions present across tasks) information conveyed by a muscle coupling (yellow intersection in Fig. 1.1(b)) from the task-relevant information (pink-and-orange shaded area in Fig.1.1(b)). These task-relevant interactions can be either sub-additive/redundant (i.e. the muscle coupling provides less information about the task compared to the sum of the individual muscle patterns) or super-additive/synergistic (i.e. the muscle coupling conveys more task information than the sum of individual muscle-task encodings). Conceptually, the information a muscle interaction provides is considered redundant when all the information can essentially be found in one of the muscles (pink shaded area (Fig.1.2(a)). This redundant task information thus reveals a functional similarity between muscle activations. Alternatively, we can also identify muscles that act synergistically towards complementary task goals, meaning their variations provide different information about a motor behaviour. A key, quantifiable attribute of this complementary interaction is the emergent task information (‘synergy’) they provide when considered together (orange shaded area (fig.1.2(b)). From this novel perspective, muscle activations can ‘work together’ not just similarly towards a common task-goal but also complementarily towards different aspects of motor behaviour and concurrently towards objectives functionally irrelevant to overt task performance, thus providing a comprehensive view of the muscle interactions governing coordinated movement.

To illustrate this novel conceptual and analytical framework, we conducted two example applications to data from human participants performing naturalistic movements. These applications demonstrate the added utility of this framework to current muscle synergy analysis in terms of functional and physiological relevance and interpretability. We then applied it to three large-scale datasets, extracting generalizable and functionally interpretable space-time muscle networks with respect to both discrete and continuous task spaces. We have also made available open-source Matlab routines for readers to apply this approach to their own data (https://github.com/DelisLab/EMG2Task).

Results

Our primary aim here is to characterise muscle synergies in task space by quantifying the contributions of muscle couplings to task performance. To achieve this, we essentially reverse the analytical approach typically used in muscle synergy studies (i.e. muscle groupings are identified and inferences then made about their functional roles) [9]. More specifically, we firstly identify functional couplings between paired muscle activations by evaluating their task-relevance and then extracting representative patterns of such couplings using dimensionality reduction methods. This enables us to distinguish task-irrelevant from task-relevant muscle couplings. Of the muscle couplings that demonstrate task-relevance, we can then characterise their functional roles as either redundant or synergistic. Fig.2 illustrates a simulation to facilitate interpretations of what informational redundancy and synergy mean when applied to muscle activities in the context of task performance. Redundant task information is generated when mx and my carry identical predictive information about τ. This is distinct from current muscle synergy analysis which would consider mx and my to share information about τ if their magnitudes are equivalent. Here, τ is always L when mx is on and when my is off and R when vice versa. Thus, all the task information can be found in either mx or my alone, generating 1 bit of redundant information. Synergistic task information on the other hand is predictive task information generated only when observing both mx and my together. In the simple example shown, τ can be L when both mx and my are active or inactive. However, we can see that when both muscles are active or inactive then τ is L. Thus, no predictive task information is provided by either mx or my alone but the full 1 bit of information available is generated when observing both muscles together.

A simulation demonstrating how informational redundancy and synergy can be interpreted when applied to the muscle space. Four observations of a given muscle pair (mx and my) that can fall into two equiprobable on and off activation states and a corresponding task parameter (τ) describing left (L) or right (R) movement direction. Observing either mx or my in the redundancy example gives 1 bit of information while observing both mx and my together in the synergy example gives 1 bit of information.

Building on current approaches to muscle synergy analysis

Current approaches to muscle synergy analysis based on non-negative matrix factorisation (NMF) have a proven use case in the extraction of functionally- and physiologically relevant motor patterns [3,12,34–36]. To demonstrate that the proposed framework adds to this current utility, here we firstly provide a simple example output from the proposed and current approaches (see fig.5(A-D)). This example was derived from the EMG recordings of a single-trial of a participant walking on level-ground in the counter-clockwise direction around the circuit depicted in fig.4(C) [27]. For the proposed approach, the muscle couplings were determined with respect to a single, continuous task parameter, the heel kinematic marker in the anterior-posterior direction. For the application of the current approach, we applied the spatial muscle synergy model across the same single-trial EMG recordings [37], extracting one component.

The simplified representations from the proposed approach reveal the functional role of muscular interactions with respect to the heel marker and provide intuition on the types of muscle couplings that can be identified, including task-irrelevant (A) -redundant (B) and –synergistic (C) interactions. Their submodular structure are illustrated via the node colour on the accompanying human body models [38], describing muscles that have a closer functional relationship. For example, a) the hamstring muscles (ST and BF) controlling knee flexion work redundantly together with muscles involved in hip abduction (GR, GlutM and Obl) to move the heel around the circuit (Fig.4(B)), and b) calf muscles involved in ankle flexion (GM, TA) cooperatively determine heel position in synergy with the same hamstring muscles (Fig.5(C)). Moreover, these hamstring muscles together with their antagonist RF also form a task-irrelevant network, i.e. their couplings are not predictive of heel position (Fig.5(A)). Overall, task-irrelevant muscle couplings primarily capture interactions between co-agonist and agonist-antagonist muscles that are indiscriminative of heel marker position. Also, task-redundant and -synergistic couplings reveal functionally similar and dissimilar muscle combinations respectively that provide sub-additive (i.e. shared or redundant) and super-additive (i.e. complementary or synergistic) information about heel marker position.

The NMF representation (fig.5(D)) conveys important information about gait in a task-agnostic manner, thus it may contain both task-relevant and -irrelevant interactions. Intuitively, SO, GM and Obl of the outer, right-side are most prominently weighted here, perhaps representing their functional role in accelerating the body in the counterclockwise direction around the circuit whilst maintaining upright posture (fig.4(C)). This observation can only be inferred indirectly however as, amongst current approaches, no direct association is made with other muscles or with task performance. Indeed, with respect to heel position, the proposed approach reveals that SO is functionally similar to Obl and GM whilst also containing task-synergistic and -irrelevant information with GM (fig.5(A-C)). From the perspective of existing approaches, the knee flexors and extensors play a minor role during turning gait (as indicated by their relatively low weighting (fig.5(D))). However, the proposed approach conveys a central role for these muscles suggesting that the proposed framework enables targeted dissections of muscle functionality with respect to a chosen task parameter, thus revealing subtle couplings with potentially important behavioural consequences.

Next, to demonstrate the additional physiological relevance the proposed methodology brings to current muscle synergy analysis, we applied the proposed framework to single trials of pointing movements performed by 20 participants with stroke and 25 healthy controls from dataset 4 (fig.4(D)) (‘MMH’ task 9 [33]). Specifically, we determined MI from a randomly selected trial of healthy and post-stroke participants with respect to the 3D position of the anterior wrist kinematic marker (WRBA) of the pointing arm. We chose WRBA as the task variable here due to its sensitivity to hand orientation. For the purpose of this simplified demonstration, we focused on a comparison between task-redundant (with respect to WRBA) and NMF-based muscle representations. We firstly generated a normative representation of this pointing motion by extracting the first component across the healthy controls using the proposed approach (Fig.6(a.1)) and NMF (Fig.6(a.2)). We then quantified the similarity of muscle representations extracted from each post-stroke participant individually to this normative reference (obtained across all healthy participants) using Pearson’s correlation and converted these values to distances (i.e. 1-r) (Fig.6(b-c)). Finally, we determined if these distances from healthy control values were predictive of the stroke survivor’s motor impairment, measured using the upper-extremity section of the Fugl-Meyer assessment (FMA-UE) (Fig.6(d)).

To briefly summarise the results, the distance of post-stroke participants from healthy controls was found to be predictive of motor impairment for the proposed approach (β= -8.52±2.2, p=0.0012) but not the NMF-based approach (β= 1.4±1.39, p=0.33). This finding suggests, intuitively, that the proposed approach captures redundant muscle couplings that support robust motor control and that deviations from this normative pattern of motor redundancy are linearly related to the degree of impairment. Importantly, this result was obtained using only one randomly selected trial for each participant. This simple example conveys the physiologically-relevant targeted insights that can be generated from the proposed framework. Although current approaches have demonstrated significant linear relationships with motor impairment [40–42], these assessments generally rely on large numbers of trials and participants and don’t point to specific underlying muscle interactions as provided here.

Together, through two example applications, we have demonstrated attributes of the proposed methodology that provide novel capabilities to current muscle synergy analysis. In the following, we sequentially present in more detail the three types of muscular interactions in the spatiotemporal domain across three datasets and then show the robustness of the approach and its outputs.

Representations of motor behaviour in muscle couplings

To begin, we derived pairs of muscle activation vectors [mx, my] from benchmark datasets of human participants executing naturalistic movements, namely arm reaching (dataset 1), whole-body point- to-point reaching (dataset 2), and various locomotion modes (dataset 3) [25–27,33] (see Fig.4 for the experimental design of each dataset and ‘Materials and methods’ for an outline of the experimental setups and EMG data pre-processing) (fig.3(A)). For datasets 1 and 2, we determine the MI between [mx, my] vectors with respect to several discrete task parameters representing specific task attributes (e.g. reaching direction, speed etc.), while for dataset 3 we determined the task-relevant and -irrelevant muscles couplings in an assumption-free way by quantifying them with respect to all available kinematic, dynamic and inertial motion unit (IMU) features.

A summary of the NIF pipeline. (A) Large-scale datasets of EMG signals are captured while participants perform various motor tasks [25–27]. (B) The MI between all unique muscle-timepoint vector ([mx, my]) combinations with respect to a corresponding task parameter (τ) is determined [28], forming a network of functional connectivities. (C) These adjacency matrices are then analysed in terms of statistical significance and modular structure using percolation theory [11]. (D) The optimal spatial and temporal model-ranks are determined using generalised, consensus-based network community detection methods [29–32]. (E) The optimal model-ranks are used as input parameters for dimensionality reduction, where space-time muscle networks along with their underlying activation coefficients are concurrently extracted [25].

Having extracted the muscle pair-task interdependencies representing a specific intersection in fig.1.1(b), we next sought to find a parsimonious representation of motor behaviour that is consistent across tasks and participants for dataset 1-3 (fig.3(E)) [11,25]. To produce this sparse, low-dimensional representation, we undertook the following intermediary steps:

We modelled the MI values as adjacency matrices in the spatial or temporal domain and identified dependencies that were statistically significant using percolation theory (fig.3(C)) [43,44]. By assuming the muscle networks operate near a state of self-organised criticality [45], we effectively isolated dependencies that were above chance-level occurrence, thus empirically sparsifying the networks.

To empirically determine the number of components to extract in a parameter-free way, we then concatenated these adjacency matrices into a multiplex network and employed network community-detection protocols to identify modules across spatial and temporal scales (fig.3(D)) [29–32,46]. Having detected the spatiotemporal modular structure, we then returned the sparsified networks to their original format and used the number of modules identified as input parameters into dimensionality reduction (fig.3(E)) [25].

By optimising a modularity-maximising cost-function [47,48], the community detection protocols we employed consistently identified three spatial (S1-S3) and two temporal (T1-T2) modules as representative of the underlying task-redundant, -synergistic and -irrelevant informational dynamics. Following their extraction, we further analysed the spatial networks from each dataset in terms of their submodular structure by applying network-theoretic tools [29,32,39]. In doing so, we identified subnetworks within each spatial network and interesting patterns of network centrality, i.e. the relative importance of a node in a network. The spatial and temporal networks of each dataset output are illustrated in panels A and B (fig.7-12) of the following sections. They are accompanied by human body models where node colour and size indicate subnetwork community affiliation and network centrality respectively [38]. The networks we extracted operate in parallel within spatial and temporal domains while having an all-to-all correspondence across domains, i.e. any spatial component can be combined with any temporal component via a task-specific coefficient (illustrated in panel C for dataset 1 and 2 outputs and in the supplementary materials (fig.4-6) for dataset 3) [25]. Unlike similar muscle synergy extraction approaches, dimensionality reduction in the NIF pipeline doesn’t seek to approximate the variance of recorded EMG data but to identify sets of muscles that share the same type of interaction. Thus, the multiplexing coefficients extracted in this framework are instead interpreted as the participant- and task-specific scaling of information overlap.

Graphical illustrations of each of the datasets analysed in the current study. (A) Dataset 1 consisted of participants executing table-top point-to-point reaching movements (40cm distance from starting point P0) across four targets in forward (P1-P4) and backwards (P5-P8) directions at both fast and slow speeds (40 repetitions per task) [25]. The muscles recorded included the finger extensors (FE), brachioradialis (BR), biceps brachii (BI), medial-triceps (TM), lateral-triceps (TL), anterior deltoid (AD), posterior deltoid (PD), pectoralis major (PE), latissimus dorsi (LD) of the right, reaching arm. (B) For dataset 2, the activity of 30 muscles was recorded while participants performed whole-body point-to-point reaching movements across three different heights and bars and in various directions, accumulating to 72 unique reaching tasks [26]. (C) The circuit navigated by participants in dataset 3 as they executed various locomotion modes is illustrated, of which level-ground walking, stair- and ramp-ascent/descent were analysed in the current study [27]. Several sub-conditions were undertaken by participants for each locomotion mode including different walking-speeds, clockwise vs. counter-clockwise direction, different stair heights and ramp inclines etc. Participants executed these tasks while the EMG of 11 muscles on the right leg ((Gluteus medius (GlutM), right external oblique (Obl), semitendinosus (ST), gracilis (GR), biceps femoris (BF), rectus femoris (RF), vastus lateralis (VL), vastus medialis (VM), soleus (SO), tibialis anterior (TA), gastrocnemius medialis (GM)) along with kinematic, dynamic and IMU signals were captured. (D) The EMG placement for dataset 4 [Deltoideus pars clavicularis (DC), Biceps brachii (BB), Triceps brachii (TB), Flexor digitorum superficialis (FDS), Extensor digitorum (ED), Brachioradialis (BR), Flexor carpi ulnaris (FCU), Extensor carpi ulnaris (ECU), Pronator teres (PT), Flexor carpi radialis (FCR), Abductor pollicus brevis (APB), Abductor digiti minimi (ADM)] [33]. A single-trial was taken from 25 healthy and 20 post-stroke participants performing a unilateral pointing movement with the index finger and arm outstretched (task 9 of the Softpro protocol (MHH)).

A simplified example output from the proposed framework applied to a single trial of turning gait from Dataset 3. (A) Task-irrelevant, (B) Task-redundant and (C) Task-synergistic synchronous muscle couplings were quantified (the unit of shared information is 1 bit) with respect to the heel kinematic marker (anterior-posterior direction). Human body models accompanying each spatial network illustrate their respective submodular structure with node colour and size and edge width indicating community affiliation [29], network centrality and connection strength respectively [38,39]. (D) A corresponding synergy representation from a single trial of turning gait from dataset 3 extracted using the spatial model from current approaches [37]. Each bar represents the relative weighting of each muscle in the synergy component.

A simple demonstration of the physiological relevance of the proposed approach (a.1-d-1) and the traditional, NMF-based approach (a.2-d.2). From dataset 4 [33], we took the EMG signals and WRBA kinematic from 20 post-stroke and 25 healthy participants. We extracted a single normative reference of healthy controls task-redundant muscle couplings with respect to WRBA (a.1) and a corresponding normative reference using NMF only (a.2). We then extracted a single component from each post-stroke participant and compared them individually with the corresponding normative reference, computing distance values (1-r) (b-c). We finally determined the predictive relationship of these distance values with a measure of upper-extremity motor impairment derived from the Fugl-meyer assessment (FMA-UE) (d).

Task-irrelevant muscle couplings

To quantify the task-irrelevant contributions of muscular interactions to motor behaviour, we conditioned the MI between [mx, my] with respect to τ (see ‘Materials and methods’ section) [28]. This conditioning effectively removes the task-relevant information, leaving information produced by pairwise muscle variations that are task-indiscriminative. Following a run through the NIF pipeline (fig.3) [11], the output from dataset 2 and 3 are presented in fig.7-8 respectively while the output from dataset 1 is presented in the supplementary materials (fig.1). The task-irrelevant space-time muscle networks we extracted from dataset 1 and 2 shared several structural features with their task-agnostic counterparts extracted in the preliminary study [11], supporting recent work showing that functional muscle network structure are heavily influenced by task-irrelevant factors such as anatomical constraints [49]. The functional connectivities identified here captured known contributions of spatiotemporal muscular interactions to aspects of motor behaviour common across tasks and participants which we outline briefly here.

Three spatial (S1-S3) and two temporal task-irrelevant muscle networks (T1-T2) were empirically identified and extracted across participants and task parameters from dataset 2 using the NIF pipeline (Panel A-B) [11,26]. (Panel C) Activation coefficients are presented to the right of the networks, indicating their task parameter-specific scaling averaged across participants. Human body models accompanying each spatial network illustrate their respective submodular structure with node colour and size and edge width indicating community affiliation [29], network centrality and connection strength respectively [38,39].

Three spatial (S1-S3) and two temporal task-irrelevant muscle networks (T1-T2) were empirically identified and extracted across participants and task parameters from dataset 3 using the NIF pipeline (Panel A-B) [11,27]. Activation coefficients are presented in supplementary materials (fig.4), indicating their task parameter-specific scaling averaged across participants in the dynamic, IMU and kinematic spaces. Human body models accompanying each spatial network illustrate their respective submodular structure with node colour and size and edge width indicating community affiliation [29], network centrality and connection strength respectively [38,39].

The temporal networks from dataset 1 and 2 captured mostly co-activations from movement onset – mid-movement and movement cessation, indicating that some co-contraction mechanisms were consistently task-irrelevant across trials. The temporal networks for dataset 2 were more diffuse compared to dataset 1, probably reflecting the more variable role of passive forces in generating movements to different heights captured in this dataset’s experimental design [26,50]. Furthermore, this co-contraction mechanism was more parsimoniously represented as a single network in dataset 3 (T1 in fig.8B), where passive forces in contrast likely played a consistently resistive role during locomotion. Interestingly, T1 for dataset 3 corresponded equivalently high for all three task spaces when corresponding with S2 which consisted of upper-leg extensors (see Supplementary materials (fig.4)). Muscle couplings indicative of agonist-antagonist pairings were also identified as separate subnetworks in S3 of dataset 3 (fig.8A). More specifically, their functional segregation appeared to be based on their distinct functional roles in forward propulsion (red nodes) and deceleration (blue nodes) during the mid-stance phase of gait, as indicated by the prominent correspondence with T2 across task spaces (see Supplementary materials (fig.4)). This finding reflects the consistent agonistic and antagonistic contributions of muscular interactions across locomotive tasks.

The gross motor function of muscle couplings was another characteristic of task-irrelevant muscle couplings that pervaded across the datasets analysed here. For instance, AD had a central role in S2 of dataset 1 while also displaying a unique pattern of connectivity with tibial musculature in S3 of dataset 2. Similarly, GlutM had a central role in S1 of dataset 3 (fig.8A). We further found a common pattern of task-irrelevant connectivity in S2 across datasets, namely the musculature about a hinge joint (elbow in dataset 1 and 2, knee in dataset 3) coupled with proximal shoulder or hip musculature, indicative of their biomechanical affordances. Finally, the passive, left-arm was connected with the tibial musculature of S3 in dataset 2 (green nodes (Fig.7(A)). To probe the underlying function of this connectivity in the left-arm, we inspected the original EMG signals. We observed periodic, tonic activations across tasks, reflective of reciprocal inhibition of contralateral limb musculature that enables unilateral movement [51].

Task-redundant muscle couplings

To characterise the functional role of task-relevant muscle couplings, we employed a higher-order information-theoretic concept known as co-information (co-I) [28,52,53]. This metric quantifies the MI between three random variables and may take on positive values (net synergistic) and negative values (net redundant) (see Fig.13 of ‘Materials and methods’ section). Co-I quantifies the task-relevant information shared between [mx, my] independently of the information generated by task-irrelevant muscular interactions. In doing so, it also defines the functional relationship between [mx, my] overall as redundant or complementary. Following the quantification of co-I for all [mx, my] and corresponding τ (pink area in orange-and-pink intersection (fig.1.3(B)), we parsed the negative values indicating redundancy into a separate matrix and rectified them. In fig.9-10, we illustrate the output following the extraction of task-redundant space-time muscle networks from datasets 2-3 across tasks and participants respectively while the output for dataset 1 is presented in supplementary materials fig.2. In the co-I formulation, task-redundant muscle couplings can be interpreted as muscle couplings that overall shared a common task-relevant functional role. For example, with reference to Figure 9 here, muscles in the networks presented in S1 (fig.9A) carry redundant information about the movement endpoint (fig.9C) with the temporal profile T1 (fig.9B) whereas S3 (fig.9A) contains muscle networks carrying redundant information about the starting point (fig.9C) with the same temporal profile T1 (fig.9B).

Three spatial (S1-S3) and two temporal task-redundant muscle networks (T1-T2) were empirically identified and extracted across participants and task parameters from dataset 2 using the NIF pipeline (Panel A-B) [11,26]. (Panel C) Activation coefficients are presented to the right of the networks, indicating their task parameter-specific scaling averaged across participants. Human body models accompanying each spatial network illustrate their respective submodular structure with node colour and size and edge width indicating community affiliation [29], network centrality and connection strength respectively [38,39].

Both dataset 1 (supplementary materials fig.2) and dataset 3 (fig.10) outputs display similar patterns of muscle couplings at the same spatial scale of an individual, task-relevant limbs’ musculature, with an emphasis on the coupling of specific muscles with all other muscles. For dataset 1, FE, BI and BR displayed this integrative pattern across S1-S3 respectively while BF, TA and ST demonstrated this pattern in dataset 3 also. The muscle networks encapsulated several functionally interpretable couplings such as the agonist-antagonist pairings of the BI and TM of S2 (dataset 1 (supplementary materials fig.2(A)) and the task redundancy of ankle dorsi-flexors, and knee/hip flexors during sloped walking for example in S2 of dataset 3 (fig.10(A)) [54]. The functional interpretation of these muscle connectivity patterns was in line with the extracted task-specific activations. For instance, S2 of dataset 1 was modulated most prominently by reaching direction when corresponding with T2, commensurate with the biomechanical affordances of this upper-arm muscle network. Furthermore, S2 of dataset 3 was specifically modulated by the right-thigh kinematic marker along the y-axis (up-down direction) for both T1 and T2 (see Supplementary materials (fig.5)). The centrality of task-redundant muscle couplings in dataset 1 and 3 suggests particular muscle activations drive the task-specific variations in the reaching arm and stepping leg muscle activities towards a common behavioural goal. It is also worth noting that the magnitude of these functional connectivity patterns appeared to be proportional to anatomical distance, as evidenced by the magnitude of connection strengths, a finding supportive of previous related research [49].

Three spatial (S1-S3) and two temporal task-redundant muscle networks (T1-T2) were empirically identified and extracted across participants and task parameters from dataset 3 using the NIF pipeline (A-B) [11,27]. Activation coefficients are presented in supplementary materials document 2 (fig.2), indicating their task parameter-specific scaling averaged across participants in the dynamic, IMU and kinematic spaces. Human body models accompanying each spatial network illustrate their respective submodular structure with node colour and size and edge width indicating community affiliation [29], network centrality and connection strength respectively [38,39].

Meanwhile at the greater spatial scale of dataset 2 (fig.9(A))), task-redundant muscle couplings were anatomically compartmentalised to the upper- and lower-body. This functional segregation was emphasised at the subnetwork level also, where the upper- and lower-body musculature of S3 for instance formed distinct submodules (blue and red nodes). Amongst the task-specific activations in dataset 2, S1 carried redundant task information about end-point target, -height and up-down direction when corresponding with T1. T2 for dataset 2 on the other hand contained mostly temporally proximal dependencies along the diagonal, suggestive of co-contraction mechanisms, which became more diffuse near movement cessation. These endpoint trajectory and co-contraction related temporal patterns were qualitatively similar to T1 and T2 of both dataset 1 and 3 respectively (see fig.10 and Supplementary materials fig.2 respectively).

Task-synergistic muscle couplings

Similarly, we isolated task-synergistic muscle couplings by parsing, instead, the positive co-I values from the computations conducted across all [mx, my] and corresponding τ into a sparse matrix where all redundant couplings were set to zero (see Fig.13 of ‘Materials and methods’ section). Task-synergistic muscle couplings here can be interpreted as a [mx, my] pair that provide complementary (i.e. functionally dissimilar) task information, thus more information is gained by observing [mx, my] together rather than separately (orange area in orange-and-pink intersection (fig.1.3(B)). In Fig.11-12, we illustrate the task-synergistic space-time muscle networks from datasets 2-3 respectively (dataset 1 output is presented in the supplementary materials fig.3).

Three spatial (S1-S3) and two temporal task-synergistic muscle networks (T1-T2) were empirically identified and extracted across participants and task parameters from dataset 2 using the NIF pipeline (Panel A-B) [11,26]. (Panel C) Activation coefficients are presented to the right of the networks, indicating their task parameter-specific scaling averaged across participants. Human body models accompanying each spatial network illustrate their respective submodular structure with node colour and size and edge width indicating community affiliation [29], network centrality and connection strength respectively [38,39].

Three spatial (S1-S3) and two temporal task-synergistic muscle networks (T1-T2) were empirically identified and extracted across participants and task parameters from dataset 3 using the NIF pipeline (Panel A-B) [11,27]. Activation coefficients are presented in supplementary materials (fig.6), indicating their task parameter-specific scaling averaged across participants in the dynamic, IMU and kinematic spaces. Human body models accompanying each spatial network illustrate their respective submodular structure with node colour and size and edge width indicating community affiliation [29], network centrality and connection strength respectively [38,39].

Co-I determines the difference between the sum total information shared with τ in mx and my when observed separately and the information shared with τ when they are observed together. The adjacency matrices show how this calculation is carried out for all unique [mx, my] combinations. Redundant and synergistic muscle couplings are then separated into two equivalently sized networks. The accompanying colour bars indicate the values present in the adjacency matrix.

Across datasets, muscle networks could be characterised by the transmission of complementary task information between functionally specialised muscle groups, many of which were identified among the task-redundant representations (Fig.9-10 and Supp. Fig.2). The most obvious example of this is the S3 synergist muscle network of dataset 2 (Fig.11), which captures the complementary interaction between task-redundant submodules identified previously (S3 (Fig.9)). A particularly consistent structural feature was the emphasis of an individual muscles’ connectivity with all other muscles which was evident among synergistic couplings in dataset 3 (see VM, VL and RF of S1, GR, TA and SO of S2 and GlutM of S3 in fig.12A). This structural similarity demonstrates that parallel and synchronous exchanges of redundant and synergistic task information underlie task-specific variations across trials (e.g. S3 of dataset 2 in fig. 9A and fig.11A). Interestingly, despite the similarity between the redundant and synergistic muscle networks, the way they are combined to encode task information differs depending on the type of interaction (synergistic vs. redundant, e.g. panel C of fig.11 in comparison to panel C of fig.9).

Concerning the temporal activations of these networks, the task-synergistic structure of dataset 2 (fig.11B) was also relatively unchanged compared to the task-redundant structure (fig.9B). This suggests that the task end-goal and co-contraction related mechanisms provided both redundant and synergistic task information concurrently during whole-body reaching movements. In contrast, a different view of synergistic information exchange is provided in datasets 1 and 3 (supplementary fig.3 and fig.12 here respectively), where T1 and T2 consist of more idiosyncratic activations that together appear to reflect the task end-goal related patterns found elsewhere. More specifically, in both datasets we found two distinct patterns of task- end goal related activity where early and late timepoints during movement initiation operated in parallel to provide complementary task information (see fig. 12B and supp. fig. 3B).

Generalisability of the extracted space-time muscle networks

To ascertain the generalisability of the extracted representations presented here beyond any subset of the input data, we conducted a similarity analysis through a leave-n-out cross validation procedure. In more detail, we compared the space-time networks extracted from the full dataset (illustrated in fig.4-9 and supplementary materials here) to the networks extracted from a subset of the data (see ‘Materials and methods’ section). Across datasets, a high level of concordance was found on average (∼0.9 correlation, see table1). This trend was evident across all datasets and for task-redundant, -synergistic and -irrelevant spatial and temporal networks. Dataset 1 and 2 findings demonstrate that the extracted networks are generalisable beyond any individual participant or task. Dataset 3 results go further by demonstrating that the extracted patterns are generalizable beyond any randomly selected and randomly sized subset of the input data. The highest correlations on average were consistently among temporal networks, replicating previous findings [11,55,56]. Although the spatial networks demonstrated a lower average correlation, this was substantially higher compared to previous applications of datasets 1 and 2 [11,25,55], suggesting that the inclusion of task parameters here captures the inter-participant differences more effectively. When comparing representations extracted from each continuous task space in dataset 3, kinematic features consistently had the highest average correlation and lowest variability compared to dynamic and IMU feature spaces. These findings support our change in the interpretation of the extracted activation coefficients away from conventional approaches where representational biases towards particular participants and/or tasks are inferred.

To further probe how the underlying assumption of an all-to-all correspondence between spatial and temporal representations made by sNM3F influenced the generalisability of the extracted networks, we compared its performance to non-negative Canonical-Polyadic (CP) tensor decomposition, which assumes an opposing one-to-one correspondence between components. We found that although CP has demonstrated a considerable capacity to de-mix neural data into simplified and interpretable low-dimensional components [57], its application here resulted in poor generalisability of the extracted patterns (∼0.5 correlation on average). This finding suggests that the all-to-all correspondence implemented by sNM3F identifies a more generalizable representation and should be favourably considered in future applications of this framework.

Discussion

The aim of the current study was to dissect the muscle synergy concept and offer a novel, more nuanced perspective on how muscles ‘work together’ to achieve a common behavioural goal. To do so, we introduced a computational approach based on the NIF pipeline (fig.3), enabling the effective decomposition of muscle interactions and the comprehensive description of their functional roles. Through the direct inclusion of task parameters in the extraction of muscle synergies, a novel perspective was produced where muscles ‘work together’ not just towards a common task-goal, but also concomitantly towards complementary task objectives and lower-level functions irrelevant for overt task performance. The functional architectures we uncovered were comprised of distributed subnetworks of synchronous muscle couplings and driven by distinct temporal patterns operating in parallel. Example applications to simple real and simulated EMG datasets revealed the additional capabilities provided by the proposed framework to current muscle synergy analysis in terms of functional and physiological relevance and interpretability. When applied to large-scale data, the proposed methodology extracted representations scale-invariant to dataset complexity and motor behaviours whilst being generalizable beyond any subset of the data. We thus present this framework as a useful analytical approach for mechanistic investigations on large-scale neural data through this novel perspective to the muscle synergy.

The ‘muscle synergy’ is a major guiding concept in the motor control field [3–7,15], providing a conceptual framework to the problem of motor redundancy that centres around the general notion of ‘working together’. In its current conceptualisation, ‘working together’ describes how the nervous system functionally groups muscles in a task-dependent manner through common neural drives to simplify movement control. This idea has undergone continued refinement since its early conception [3,6,7], with a notable progression being the introduction of the qualifying attributes: a sharing pattern, reciprocity and task dependence [7]. Nonetheless, recent influential works revealing other important mechanisms for the simplification of motor control highlight the generality of the current perspective offered by this concept [16–24]. We thus sought to provide greater nuance to the notion of ‘working together’ by defining motor redundancy and synergy in information-theoretic terms [6,58]. In our framework, redundancy and synergy are terms describing functionally similar and complementary motor signals respectively, introducing a new perspective that is conceptually distinct from the traditional view of muscle synergies as a solution to the motor redundancy problem [3,6,7]. In this new definition of muscle interactions in the task space, a group of muscles can ‘work together’ either synergistically or redundantly towards the same task. In doing so, the perspective instantiated by our approach provides novel coverage to the partitioning of task-relevant and - irrelevant variability implemented by the motor system along with an improved specificity regarding the functional roles of muscle couplings [20–22]. Our framework emphasises not only the role of functionally redundant muscle couplings that result from the underlying degeneracy of the motor system, but also of complementary, synergistic dependencies that are important for communication and integration across specialised neural circuitry [59,60]. Thus, the present study aligns the muscle synergy concept with the current mechanistic understanding of the nervous system whilst offering an analytical approach amenable to the continued advances in large-scale data capture [14,61].

Among current approaches to muscle synergy analysis, the established synchronous, temporal and time-varying muscle synergy models are understood to each characterise unique motor features [62]. More specifically, the synchronous model captures agonist-antagonist muscle pairings, the temporal model decomposes EMG signals into functionally distinct temporal phases and the task-specific modulation of spatiotemporal invariants are quantified in the time-varying model. In a unifying framework, here we quantify space-time muscle networks that concurrently capture many of these salient motor features in a holistic and principled way whilst mapping their functional consequences to motor behaviour. These salient features included, among others, agonist-antagonist pairings and functionally meaningful inter-limb couplings that consistently appeared across task-redundant, -synergistic and -irrelevant spaces. Thus, in dissecting the muscle interactions governing coordinated movement, our framework revealed their parallel and synchronous processing of functionally similar, complementary, and task-irrelevant information. This insight aligns with several recent works also demonstrating this distributed neural architecture of parallel information processing units [1,59,60].

Our framework also revealed novel characteristics of the motor system. For instance, the task-redundant and -synergistic networks we extracted appeared to be structured around the coupling between a prime-mover muscle and several supporting muscles, supporting recent work showing the nonhomogeneous sharing of neural drives within modules [63]. These novel spatial characteristics were driven by parallel temporal patterns representing endpoint trajectory and co-contraction related mechanisms, an insight supportive of recent work showing their parallel innervation [16,50,64]. Together, these representations encapsulated the functional interplay between task end-goal requirements and biomechanical affordances, a dynamic frequently highlighted in object manipulation experiments [65]. In other words, the task-relevant networks reflected how muscles ‘work together’ both redundantly and synergistically towards a desired end-goal state whilst, in parallel, continually controlling the present trajectory of the system. Meanwhile, the task-irrelevant networks demonstrated that these muscle couplings also work concomitantly towards lower-level objectives assigned mostly during the transition to this desired end state. Although distinguishing task-irrelevant muscle couplings may capture artifacts such as EMG crosstalk, our results convey several physiological objectives of muscles including gross motor functions [66], the maintenance of internal joint mechanics and reciprocal inhibition of contralateral limbs [19,51]. Thus, task-irrelevant muscle interactions reflect both biomechanical- and task-level constraints that provide a structural foundation for task-specific couplings. The separate quantification of these muscle interaction types opens up novel opportunities in the practical application of muscle synergy analysis, as demonstrated in the current study through the identification of a significant predictor of motor impairment post-stroke from single-trials [5,12,67]. For instance, these distinct representations may encapsulate different neural substrates that can be specifically assessed at the muscle-level for the purpose of bodily restoration and augmentation [68]. Uncovering their neural underpinnings is an interesting topic for future research.

Indeed, in future work, we aim to complement this study’s combinatorial perspective to the muscle synergy by dissecting the unique contribution of individual muscles to motor behaviour and how they may work independently towards task performance (see magenta and cyan intersections in fig.1.3(B)). More broadly, our work here parallels related information-theoretic approaches to decomposing task-relevant brain activity [53,69], whilst addressing a current research gap across the neurosciences in effective methods to mapping large-scale, spatiotemporal neural activity to behaviour [14,61]. Future applications of this framework should include large-scale, multi-modal data captured from participants performing a wide range of natural behaviours.

In sum, this study introduced a novel perspective to the muscle synergy concept and a computational framework to extract muscle couplings that map their pairwise contributions to motor behaviour. We suggest that this approach offers novel research opportunities for investigating the underlying neural constraints on motor behaviour and the fundamental structure-function relationships generated by agent – environment interactions [15,70].

Materials and Methods

Quantifying muscle couplings in the task space

To quantify muscle couplings we used MI, a non-linear measure of dependence that captures any type of relationship between random variables. Here to estimate MI, we used a Gaussian copula-based approximation [28]. This semi-parametric estimator exploits the equivalence between MI and the negative entropy of the empirical copula (c), a function that maps a multivariate set (e.g. [mx, my] representing activities of muscles X and Y) to their joint distribution (Equation 1.1) [28]. Thus, to determine task-irrelevant muscle couplings (I(mx; my|τ)), we conditioned the negative entropy of the empirical copula for [mx, my] with respect to a task variable τ (Equation 1.2). As mentioned, [mx, my] are continuous vectors composed of individual muscle amplitudes at specific timepoints across trials while τ is a corresponding discrete (e.g. movement direction for datasets 1 and 2) or continuous (e.g. movement kinematics in dataset 3) task parameter. For discrete task variables, τ takes one value for each trial and the MI is calculated across trials using a Gaussian mixture model [28]. In the case of continuous task variables, τ varies in time within a specific trial. Thus, we compute MI at each timepoint t using the muscle activity mx(t) and the task variable value τ(t) at this time point using a closed-form parametric estimator [28]. To evaluate the task-relevance of the identified muscle couplings, we used a higher-order information theoretic measure known as co-information (co-I) (Equation.1.3) (fig.11), which quantifies the relationship between three random variables, here [mx, my], and τ. Co-I implements the inclusion-exclusion principle of combinatorics [52], whereby the sum of MIs between individual m vectors and τ (I(mx; τ) + I(my; τ)) is compared against their composite MI (I(mx, my; τ)) as follows: Negative I(mx; my; τ) corresponds to a net redundant coupling between [mx, my] about τ while positive I(mx; my; τ) indicates a net synergy. To analyse these distinct couplings separately, we parsed redundant and synergistic I(mx; my; τ) into two equivalently sized matrices and rectified the redundant couplings to make them suitable for non-negative dimensionality reduction.

Then, to produce a multiplexed view of the muscular interactions across trials, we iterated these MI computations over all unique combinations of [mx, my] and τ. The resulting MI estimates collectively form A, a symmetric adjacency matrix (i.e. ATA = I) that represents the functional connectivities between all muscle activations (fig.10). When repeated across all available task variables τ and participants, A is of dimension [No. of muscle pairs x No. of timepoint pairs x [No. of τ x No. of participants]]. Thus, by applying network-theoretic statistical tools to A, we can identify functional modules carrying the same type of (redundant/synergistic) task information (fig.2(B)).

Estimating statistical significance of muscle couplings

To isolate statistically significant dependencies, we applied a modified percolation analysis to each A [43]. This method sparsifies functional connectivities in A with respect to its percolation threshold (Pc). Pc is a critical value that specifies the probability of a nodes’ occupation in A with respect to the networks size. In random networks, a ‘giant component’ comprised of long-range connections exists above Pc but disappears as Pc tends to zero [44], while it is thought that living systems optimise adaptability by fluctuating around Pc in a state of self-organised criticality [45]. Preliminary testing of this method showed it to be at least equivalent to permutation-testing each MI value in the network and thus, much more computationally efficient. Pc was therefore iteratively specified for each layer of A relative to equivalently sized random networks and utilised to remove insignificant network edges up to a stopping-point where this giant component begins to become affected (fig.3(C)). This procedure was carried out for each layer of A separately configured as muscle-wise couplings across temporal scales (i.e. a 3-D tensor of dimension [No. of muscles] x [No. of muscles] x [No. of timepoint pairings x No. of τ x No. of participants]) and vice versa as timepoint-wise couplings across spatial scales (fig.3(D)). The separate sparsification of each individual network layer in both alternative network configurations produced discrepancies in the output, as some connections were found to be significant in only one domain. To ameliorate this discrepancy, we employed a conservative heuristic where dependencies must be significant in both space and time to be included in the final input matrix for dimensionality reduction. Thus, the sparsified input matrices to dimensionality reduction were comprised of significant spatiotemporal task-redundant, -synergistic or - irrelevant muscle couplings.

Model-rank specification

To determine the optimal number of modules to extract, we implemented two alternative community detection algorithms generalised to multiplex networks [29,30,46]. Both forms seek to optimise a modularity criterion known as the Q-statistic that quantifies the proportion of within-community network edges compared to what would be expected from a network consisting of random connections [47]. More specifically, for a particular division of a single layer network (Equation 2.1), let δ(gi, gj)=1 if node i and j belong to the same group (g) and 0 otherwise and Aij be the number of edges between node i and j. The equivalent of Aij from a randomised network (Pij) is expected to be (Newman-Girvan null model) [47], where ki and kj are the node degrees and . The typical output of the Q-statistic is found within the range [0,1] with 1 indicating maximum modularity [47]. In its generalised multilayer form, the Q-statistic is given an additional term to consider couplings between layers l and r with intra- and inter-layer resolution parameters γ and ω (Equation 2.2). Here, μ is the total edge weight across the network and γ and ω were set to 1 in the current study for classical modularity [30], thus removing the need for any hyperparameter tuning. We chose to implement two complementary model-rank specification approaches to address limitations related to stochasticity and scalability present in the multilayer formulation and the inability to consider inter-layer dependencies present in the mono-layer formulation [18,48]. To apply these algorithms to our data, we grouped the set of A into multiplex networks configured with respect to spatial or temporal scales (fig.3(D)). We then applied these algorithms to both space-time network configurations for individual participant/tasks. This procedure generated a binary adjacency matrix from the resulting community partition vector in each case where 1 indicated the nodes belonged to the same community and 0 otherwise [46]. Following a consensus-based approach [31], we then grouped these binary adjacency matrices into a new multiplex network and re-applied the two alternative community detection algorithms to find an optimal spatial and temporal model-rank [31].

Extraction of low-dimensional representations

Following the specification of an optimal model-rank in the spatial and temporal domains, we used these values as input parameters into dimensionality reduction (fig.3(E)). To extract a low-dimensional representation of motor behaviour across muscle couplings, we applied a sample-based non-negative matrix tri-factorisation method (sNM3F) with additional orthogonality constraints to the matrices consisting of vectorised and concatenated A [25]. More specifically, we decomposed the input three-mode tensor A of dimension [K = No. of unique muscle pairs (m) x (L = No. of time-sample pairs (t) x No. of participants + tasks)] into a set of spatial (V) and temporal (W) factors and the participant- and task-specific weighting coefficients (S) reflecting the amount of information carried by each combination of spatial-temporal factors for each participant and task parameter (Equation 3.1). In equation 3.2, this factorisation is also illustrated in vector sum notation for a single participant and task variable:

Examining the generalisability of extracted representations

To determine the generalisability of the extracted space-time muscle networks, we implemented a representational similarity analysis where we compared representations extracted from the full dataset 1-3 to equivalent networks extracted from a subset of the respective datasets. We computed the similarity between pairs of representations using Pearson’s correlation.

For dataset 1 and 2 (see below), we removed from the input data an individual participant or task at a time and compared the similarity of the decomposition outputs with those obtained from the full dataset. We repeated this for all participants and task variables and reported the average similarity as a measure of robustness of the decomposition.

For dataset 3 (see below), due to the greater number of participants and task parameters, we implemented a more stringent examination. More specifically, we firstly extracted representations from each task space individually and, using these representations as a reference, compared them against functionally similar outputs after removing randomly sized portions of randomly selected vectors in the input data (up to the no. of column vectors - 1). We repeated this procedure for 50 iterations, and computed summary statistics by converting the coefficients to Fisher’s Z values, computing the average and standard deviation, and then reverting these values back to correlation coefficients.

Subnetwork analysis

To illustrate the relative importance of individual muscles within each network, we determined the total communicability (C(i)) of individual nodes (i) in each network (A). C(i) is defined as the row-wise sum of all matrix exponentials (e) in the adjacency matrix (A) that consider the number of walks between each pair of nodes i and j (Equation 4.1) [39,71]: To emphasise salient functional connectivities in the spatial networks, we sparsified all dependencies with a below average network communicability and illustrated the output on the accompanying human body models [38,71]. To uncover salient subnetwork structures consisting of more closely functionally related muscle activations, we applied the monolayer community detection algorithm in equation 2.1 to the extracted spatial networks [29,32].

Data acquisition and processing

To illustrate our framework, we applied it to three datasets of EMG signals recorded during different motor tasks. In dataset 1 (fig.4(A)), 7 adult participants (age: 27 ± 2 years, height: 1.77 ± 0.03 m) performed table-top point-to-point reaching movements in both forward and backwards directions and at fast and slow speeds while the activity of nine muscles on the preferred right, reaching arm (finger extensors (FE), brachioradialis (BR), biceps brachii (BI), medial-triceps (TM), lateral-triceps (TL), anterior deltoid (AD), posterior deltoid (PD), pectoralis major (PE), latissimus dorsi (LD)) were captured for a total of 640 trials per participant [25]. To enable the quantification of shared information across muscles with respect to specific task attributes, we formulated four discrete task parameters of length equal to the number of trials executed. These discrete variables represented the trial-to-trial variation in reaching direction (fwd vs bwd), speed (high vs low), and reaching target [P1-P8] (fig.3(A)).

In dataset 2 (fig.4(B)), 3 participants performed whole-body point-to-point reaching movements in various directions and to varying heights while EMG from 30 muscles (tibialis anterior, soleus, peroneus, gastrocnemius, vastus lateralis, rectus femoris, biceps femoris, gluteus maximus, erector spinae, pectoralis major, trapezius, anterior deltoid, posterior deltoid, biceps and triceps brachii) across both hemi-bodies were captured [26]. Like dataset 1, we formulated task parameters each representing a specific task attribute across trials (∼2160 trials per participant). In this case, we formed eight discrete task parameters representing start- and end-point target, -bar, and -height and both up-down (vertical) and left-right (horizontal) reaching directions.

Dataset 3 consisted of multiple trials from 17 participants performing level-ground walking, stair- and ramp-ascents/descents with various sub-conditions (walking speed, clockwise/counter-clockwise direction, different stair/ramp inclines etc.) (fig.4(C)) [27]. These locomotion modes were performed while 11 EMG signals were captured from the right lower-limb (Gluteus medius (GlutM), right external oblique (Obl), semitendinosus (ST), gracilis (GR), biceps femoris (BF), rectus femoris (RF), vastus lateralis (VL), vastus medialis (VM), soleus (SO), tibialis anterior (TA), gastrocnemius medialis (GM)), XYZ coordinates were captured bilaterally from 32 kinematic markers and 4 IMUs and a force-plate captured accelerations and dynamic features among the lower-limbs also. More detailed breakdowns of the experimental design for each dataset can be found at their parent publications [25–27].

For all datasets, we processed the EMG signals offline using a standardised approach [10]: the EMGs for each sample were digitally full-wave rectified, low-pass filtered (Butterworth filter; 20 Hz cut-off; zero-phase distortion), normalised to 1000 time-samples and then the signals were integrated over 20 time- step intervals yielding a waveform of ∼50 time-steps. To match the time-series lengths, we resampled the kinematic, dynamic and IMU recordings of dataset 3 using cubic-spline interpolation to match the EMG signals.

Supporting information